### 2017-12-01 manuscript/main.tex
latest draft

%% Submissions for peer-review must enable line-numbering 
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need 
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
% \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions

\title{Testing the usability of large scale citizen science data for phenology modeling}

\author[1]{Shawn D. Taylor}
\author[2]{Ethan P. White}
\affil[1]{UFL}
\affil[2]{UFL}
\corrauthor[1]{Shawn D. Taylor}{shawntaylor@weecology.org}

% \keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section*{Introduction}

Citizen science data is becoming increasingly important for ecological research, with the potential for significant impact in many fields \citep{dickinson2010, tulloch2013, kelling2009}. Citizen science efforts have helped document range shifts \citep{hitch2007}, study the effect of landscape fragmentation on amphibians \citep{cosentino2014}, and informed ecological theory of community composition \citep{locey2013}. A primary benefit of citizen science data is observations made at numerous sites across regional to continental scales. This helps researchers explain large scale patterns across natural gradients, but can also introduce new sources error because of observer skill, variation in sampling effort, and spatial bias \citep{dickinson2010}. These issues can be accounted for in models but first must be identified and quantified. As new citizen science datasets are made available, they must be evaluated for their own biases and limitations.

The National Phenology Network (NPN) is an organization which accepts observations of phenology from volunteers throughout North America and makes the data freely available \citep{schwartz2012a}. Plant phenology is important for studies of ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}. Robust phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant communities. At the largest scale variation in the timing of spring leaf out and fall senescence account for a significant amount of error in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long term forecasts \citep{richardson2012}. At local scales, because of species specific responses to temperature and precipitation, changes in phenology can alter the flower community available to pollinators and seed predators throughout the growing season \citep{diez2012, caradonna2014, ogilvie2017, theobald2017}.

The NPN dataset has already being used to study, among many things, variation in oak phenology \citep{gerst2017}, large scale community phenology models \citep{melaas2016}, and forecasting long term phenology trends \citep{jeong2013}. While there has been several studies looking at best practices when using this dataset \citep{crimmins2017, gerst2016} no study to date has looked a the underlying validity of the data. With hundreds of observers across North America, the observer and sampling bias for species identification and the true date of phenological events is potentially high. Thought volunteers can be very accurate at distinguishing different phenophases \citep{fuccillo2015} observations can still be made sporadically throughout the growing season, some years skipped, or visits to a particular site discontinued entirely. These biases could adversely affect process based phenology models, which are species specific and depend on high precision in the data \citep{chuine2000}.

The NPN dataset is complicated further by the spatial variation of the temperature response within species. Phenology models currently have trouble capturing the spatial variation in warming requirements \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}. Most phenology models are process based and designed to reflect physological processes driving phenological events. Yet models parameterized using observations from one site rarely  transfer to other sites, and models using pooled data from many sites do not transfer well to individual sites \citep{garcia-mozo2008, basler2016}. This variation within species can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995,diez2012}. While this complicates further any testing of dataset validity, it's worth it because there is a large opportunity with this dataset for further study of phenology across spatial gradients.

Many phenological studies in the past have relied upon datasets with observations collected at a single location over a long timeframe by a single group \citep{cook2012, roberts2015, clark2014}. These datasets are assumed to not suffer from the potential biases mentioned above, and many also have high temporal coverage of several decades. They are disadvantageous as the inference from the models can only be applied to species located at the site, and only to local climate variability. These datasets have been invaluable for ecology for several decades, but datasets with a larger spatial extent and more complete species pool will be needed to model the complex phenology dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}. The NPN datasets has the potential to meet these needs, and to date no study has assessed how well NPN data compares to long term study phenology datasets.

Here we will compare process based phenology models built using only the NPN data with models built with data from Long Term Study (LTS) datasets. We will also attempt to explain the differences in esimates of phenological events. The specific research questions are:
\begin{enumerate}
\item Do parameter estimates of species level models differ when they are built with the NPN dataset versus an LTS dataset?
\item Do estimates of phenological events differ between the NPN and LTS derived models?
\item How well do models built with the NPN dataset transfer to LTS observations, and vice verse?
\end{enumerate}

\section*{Methods}

\subsection*{Datasets}

The National Phenology Network collections began in 2009 and continue to the present. Observations are based on status-based monitoring, where observers answer “yes”, “no”, or “unsure” when asked if a plant has a specific phenological phase present \citep{denny2014}. We downloaded all NPN observations from 2009-2016 for the phenophases Breaking Leaf Buds, Breaking Needle Buds, Emerging Needles, and Open Flowers. These phenophases apply to deciduous broadleafs, evergreen conifers, pines, and all angiosperms, respectively. Hereafter we will refer to these as either Flowers for the Open Flower phenophase, or Leaves for all other phenophases. We used moderately strict subsetting of NPN phenology observations as outlined in \cite{crimmins2017}. First, "yes" observations for individual plants were only kept if they were preceded by a "no" observation within 30 days. Observations for Leaves past day of year (DOY) 172 and for Flowers past DOY 213 were dropped. Finally only species that had greater than 30 total observations were kept. \cite{crimmins2017} only kept observations that were preceded by a "no" within 15 days, and also grouped multiple observations at single sites in a year to a single observations. We used 30 days to allow for a greater number of species to be compared. We did not group together multiple individuals at a single site to better incorporate intra-site variability. In addition to these steps we also inferred the date of each phenophase as the midpoint between each "yes" observation and the preceding "no" observation.

Notes on LTS data preproccesing

\renewcommand{\arraystretch}{2}\tabcolsep=5pt
\begin{center}
    \begin{tabular}{ | l | l | l | l |}
    \hline
    Dataset Name & Habitat &  Phenological Event\newline (Num. Species) & Reference \\ \hline
    Harvard & N.E. Deciduous\newline Forest & Budburst (16)\newline Flowering (15) & ... \\
    Jornada & Chihuahuan Desert & Flowering (2) & ... \\
    HJ Andrews \newline Experimental Forest & N.W. Wet Coniferous \newline Forest & Budburst (6)\newline Flowering (6) & ... \\
    Hubbard Brook & N.E. Deciduous \newline Forest & Budburst (3) & ... \\
    \hline
    \end{tabular}
\end{center}

To compare against the NPN dataset we used four LTS datasets from North America (Table 1). Three major ecosystem types and 24 species are represented in these datasets. As all four have been collected by single groups we make the assumption that they are free of any bias in species identification or sampling effort. Observation metrics varied among these four datasets due to different protocols, so were first converted to binary "yes" and "no" observations for each phenophase (see supplement for details). As with the NPN dataset we infer the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation.

\subsection*{Models}

For each species/phenophase in each of the five datasets we fit a range of phenology models from a naive mean day of year estimate to a more complex two phase forcing model (Table 2). With the exception of two models (naive and linear) the general model form predicts that a phenological event will happen when sufficient forcing units, \(F^{*}\), accumulate from a particular start day of the year. The day forcing accumulation begins, $t_{1}$, can either be estimated or fixed. For the growing degree day model (GDD) forcing units are the degrees in Celsius above an estimated threshold. The Fixed GDD model uses the same form but has fixed values for start day (Jan 1) and temperature threshold (0C). The Alternating model has a variable number of required forcing units defined as function of the total number of days below 0C (NCD). The Uniforc model is similar to the GDD model but has forcing units derived a sigmoid function.

The naive model uses only the mean DOY from all prior observations as the estimated DOY. The linear model uses a regression with the mean spring (Jan. - March) temperature as the only independent variable and DOY as the response variable.

\begin{center}
{\def\arraystretch{2}\tabcolsep=5pt
    \begin{tabular}{ | l | c | c | p{1.3cm} | l |}
    \hline
    Name & DOY Estimator & Forcing Equations & Total\newline Parameters & Reference \\ \hline
    Naive & \( \overline{DOY} \) & - & 1 & - \\
    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(x_{t})\geq F* $  & $R_{f}(x_{t}) = min(T, 0)$ & 1 & - \\
    Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(x_{t})\geq F* $ & $ R_{f}(x_{t}) = max(T - T^{*}, 0) $  & 3 & - \\
    Alternating & $\sum_{t=0}^{DOY}R_{f}(x_{t})\geq a + be^{cNCD(t)} $ & $R_{f}(x_{t}) = min(T, 0) $ & 3 & \citep{cannell1983} \\
    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(x_{t})\geq F* $ & $ R_{f}(x_{t}) = \frac{1}{1 + e^{b(x-c)}} $ & 4 & \citep{chuine2000} \\

    \hline
    \end{tabular}
    }
\end{center}

Parameterization was done using differential evolution to minimize the root mean square error (RMSE) of the estimated day of year of the phenological event. Differential evolution is a global optimization algorithm similar to simulated annealing, which has traditionally been used to fit complex phenology models \citep{storn1997, chuine2000}. Initial testing showed that differential evolution produced more consistent results than simulated annealing and other similar algorithms. Confidence intervals for parameters were obtained by bootstrapping, where individual models were re-fit 250 times using a random sample, with replacement, of the data. A random 20\% of all data was held out from model fitting for use in later evaluation.

\subsection*{Analysis}

As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set of parameterized using only LTS data. The distribution for each parameter in each model could then be compared between the two sets using a Mann-Whitney test. This was chosen due to the majority of parameters having non-normal distributions derived from the bootstrapped estimates. Statistically signficiant differences in these parameters, which are designed to represent biological mechanisms, would imply that the NPN dataset does not match

To compare estimates of LTS and NPN derived models we calculated the root mean squre difference (RMSD) between the two estimates as:

$$ RMSD = \sqrt[]{ \frac{\sum_{i=1}^{n}(\widehat{DOY_{NPN,i}} - \widehat{DOY_{LTS,i}})^{2}}{n}} $$

where $n$ is the number of estimates produced for a single species and phenophase, and $LTS_{i}$ and $NPN_{i}$ are DOY estimates produced from the respective models. A RMSD value close to 0 implies that an NPN model produced very similar estimates to the corresponding LTS model. For each species and phenophase 12 RMSD values were calculated from the 6 phenology models as well as 2 spatial extents. The first spatial extent used only estimates from the all the species/phenophase observations from the NPN datasets, thus representing a large spatial extent. The second was from all observations from the LTS datasets, representing a single site. The RMSD does not consider actual observations as it is meant to infer the correspondence of the NPN and LTS datasets beyond parameter comparison.

For all models the actual error was calculated using the 20\% of held out data.

$$ RMSE = \sqrt[]{ \frac{\sum_{i=1}^{n}(\widehat{DOY_{i}} - DOY_{i})^{2}}{n}} $$

Pairwise model error comparison

$$ RMSE_{difference} = \widehat{RMSE_{NPN,s}} - RMSE_{LTS,s} $$

for all $$S = s_{1},...,s_{39}$$ species

\begin{figure}[]
	\centering
		\includegraphics[width=1\textwidth]{fig_1_param_comparison.png}
	\caption{Comparisons of parameters between NPN and LTS derived models. Each point represents a parameter value for a specific species/phenophase, and is the mean value from 250 bootstrapped  models. The black line is a 1:1 line.}
\end{figure}

\section*{Results}

All parameters had statistically different distributions between LTS and NPN derived estimates when tested with a Mann-Whitney U test. When comparing the mean values of parameters among all species and phenophases there was correspondence among the simple models with 1-2 parameters (Fig. 1). As model complexity increased this correspondence between the two datasets decreased. The Naive model showed a distinct late bias in the mean DOY for phenologcial events, likely resulting from the LTS datasets being mostly in the northern United States. The Fixed GDD model had the best correspondence between the two datasets. The outlier for this model, \textit{Larrea tridentata}, has phenology largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{fig_2_estimate_compare.png}
	\caption{Difference between NPN and LTS derived estimates. Each point is, for a single species/phenophase, the root mean square difference between LTS and NPN derived estimates for a specified model. A shows the RMSD over the extent of the NPN dataset, B shows the RMSD at the single sites represented by the LTS datasets.}
\end{figure}

When comparing estimates of phenological events between the two sets of models, many NPN models produced similar estimates to LTS models (Fig. 2). When estimating at local LTS sites, NPN models produced estimates within 10 days, on average, of LTS models except when using the Naive model (Fig. 2b). When estimating across the spatial extent of NPN data, except when using the Linear and Naive model, most NPN models produced estimates within 20 days of LTS models (Fig. 2a).

Figure 3 shows the errors for all different model types when tested against the held out data. Models predicting NPN observations (Fig. 3a) has higher error rates than predictions for LTS observations (Fig. 3b). For the former models derived from NPN data made the best predictions, while LTS derived models performed best for the latter. Figure 4 shows the pairwise comparison of LTS and NPN errors. Similar to Figure 3, the NPN modeld performed best on NPN observations and LTS models performed best on LTS observations. 
\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{fig_3_error_compare.png}
	\caption{Root mean square error of model predictions compared to observations. The top plot is for all held out observations in the NPN dataset. The bottom plot is for all held out observations in the LTS datasets.}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{fig_4_pairwise_error.png}
	\caption{Density plots for pairwise RMSE comparison. The x axis shows the distribution for $RMSE_{NPN} - RMSE_{LTS}$  for the  39 species/phenophase combinations.}
\end{figure}

-Even with different parameter values, some species/models had extremely similar
estimates, suggesting these models had

\section*{Discussion}

Despite having statistically different parameters to LTS derived models, many NPN derived models were able to produce similar estimates for phenological events. This is (likely) due to models having a non-linear parameter space and being non-identifiable. For example two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 can produce nearly identical results in many, but not all, scenarios. As model complexity icnreases the ability to produce similar estimates increases. Indeed, the Uniforic models with NPN derived parameters still had estimates within 10 days of LTS derived parameters for most species (Fig. 2) despite having very different parameters (Fig. 1).

This makes the interpretation of these parameters questionable, which is troublesome as they are designed to be biologically relevant. \cite{chuine2016} found similar issues, where models estimating endodormancy break using only observed budbreak did not match direct measurements of endodormancy break, suggesting that phenology models which are fit using only the observed phenophase cannot accurately infer the internal mechanisms described in the models. \cite{basler2016} suggests that the thermal forcing in models is responsible for most of the accuracy, while any additional parameters are only fit to residuals and sensitive to over fitting. Here our simplest model, the Fixed GDD model which uses only a warming component, had reasonable (thought statistically different) correspondence between LTS and NPN datasets. This correspondence decreases as more parameters are added to models, suggesting only simple models should be used when using the NPN dataset to make inferences about plant physiology. Parameter estimates of more complex models do not agree with models from the same species in LTS datasets, and it is possible even the LTS datasets are over fitting these models.

Even with similar parameters, the Fixed GDD model was not the best performing model when tested on out of sample data. While the values of additional parameters may be incorrect in the above sense, they nonetheless provide more predictive power. Considering only NPN derived models predicting out of sample NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1). The GDD and Uniforc models have three and four parameters, respectively. Thus the simpler GDD model shows promise for building models with the NPN dataset and making forecasts across large scales.

Other models show promise for forecasting under different scenarios. For example if using data from a large spatial extent to make predictions at single site our results show the Fixed GDD model is the best option, as performance under this scenario is very similar to performance of Fixed GDD models built from LTS data (Fig. 4). In the opposite scenario of making large scale predictions using only data from a single site the Alternating is likely the best option, as most LTS derived models have a RMSE only 5 days greater on average than the corresponding NPN model when tested on out of sample NPN data (Fig. 4). The Alternating model is also one of the best performing model overall for large scale NPN data (Fig. 3a).

Conclusion
Despite being collected across a large spatial extend and by hundreds of observers the NPN dataset can be used to build phenology models which make comparable predictions to LTS datasets. Indeed, if making predictions across a large spatial extent the NPN derived models are preferable to LTS derived models. If using the NPN dataset to make inferences about plant physiology researchers should take not to use overly complicated models, as parameter estimates from complex NPN derived models do not agree with LTS derived models. Whether this disagreement stems from the spatial variation in plant phenology requirements, or in the irregular sampling of the NPN dataset is an avenue for future study.

\section*{Acknowledgments}

\bibliography{refs}

\end{document}

### 2017-12-22 manuscript/main.tex
manuscript progress

@@
 
-\title{Testing the usability of large scale citizen science data for phenology modeling}
-
+%% \title{Testing the usability of large scale citizen science data for phenology modeling}
+\title{Comparison of large scale citizen science data and long term study data for phenology modeling}
 \author[1]{Shawn D. Taylor}
@@
 Plant phenology is important for studies of ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}.
-Robust phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant communities.
+Robust phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant and animal communities.
 At the largest scale variation in the timing of spring leaf out and fall senescence account for a significant amount of error in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long term forecasts \citep{richardson2012}.
@@
 With hundreds of observers across North America, the observer and sampling bias for species identification and the true date of phenological events is potentially high.
-Thought volunteers can be very accurate at distinguishing different phenophases \citep{fuccillo2015} observations can still be made sporadically throughout the growing season, some years skipped, or visits to a particular site discontinued entirely.
+Though volunteers can be very accurate at distinguishing different phenophases \citep{fuccillo2015} observations can still be made sporadically throughout the growing season, some years skipped, or visits to a particular site discontinued entirely.
 These biases could adversely affect process based phenology models, which are species specific and depend on high precision in the data \citep{chuine2000}.
@@
 The NPN datasets has the potential to meet these needs, and to date no study has assessed how well NPN data compares to long term study phenology datasets.
-
 Here we will compare process based phenology models built using only the NPN data with models built with data from Long Term Study (LTS) datasets.
-We will also attempt to explain the differences in esimates of phenological events.
 The specific research questions are:
@@
 
-Notes on LTS data preproccesing
+To compare against the NPN dataset we used four LTS datasets from North America (Table 1).
+Three major ecosystem types and 24 species are represented in these datasets.
+As all four have been collected by single groups we make the assumption that they are free of any bias in species identification or sampling effort.
+Observation metrics varied among these four datasets due to different protocols, so were first converted to binary "yes" and "no" observations for each phenophase (see supplement for details).
+As with the NPN dataset we infer the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation.
 
@@
     Dataset Name & Habitat &  Phenological Event\newline (Num. Species) & Reference \\ \hline
-    Harvard & N.E. Deciduous\newline Forest & Budburst (16)\newline Flowering (15) & ... \\
+    Harvard & N.E. Deciduous\newline Forest & Budburst (16)\newline Flowering (15) & \citep{okeefe2000} \\
     Jornada & Chihuahuan Desert & Flowering (2) & ... \\
@@
 \end{center}
-
-To compare against the NPN dataset we used four LTS datasets from North America (Table 1).
-Three major ecosystem types and 24 species are represented in these datasets.
-As all four have been collected by single groups we make the assumption that they are free of any bias in species identification or sampling effort.
-Observation metrics varied among these four datasets due to different protocols, so were first converted to binary "yes" and "no" observations for each phenophase (see supplement for details).
-As with the NPN dataset we infer the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation.
 
@@
 
-As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set of parameterized using only LTS data.
-The distribution for each parameter in each model could then be compared between the two sets using a Mann-Whitney test.
+As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set parameterized using only LTS data.
+The distribution for each parameter in each model was then compared between the two sets using a Mann-Whitney test.
 This was chosen due to the majority of parameters having non-normal distributions derived from the bootstrapped estimates.
-Statistically signficiant differences in these parameters, which are designed to represent biological mechanisms, would imply that the NPN dataset does not match
+Statistically significant differences in these parameters, which are designed to represent biological mechanisms, would imply that the LTS and NPN datasets do not agree on
 
-To compare estimates of LTS and NPN derived models we calculated the root mean squre difference (RMSD) between the two estimates as:
+Even models with different structures can produce similar estimates of phenological events \citep{basler2016}.
+Thus we also compared estimates between LTS and NPN derived models using the root mean square difference (RMSD) between the two estimates:
 
@@
 where $n$ is the number of estimates produced for a single species and phenophase, and $LTS_{i}$ and $NPN_{i}$ are DOY estimates produced from the respective models.
-A RMSD value close to 0 implies that an NPN model produced very similar estimates to the corresponding LTS model.
+A RMSD value close to 0 implies that the LTS and NP derived models produced very similar estimates.
 For each species and phenophase 12 RMSD values were calculated from the 6 phenology models as well as 2 spatial extents.
-The first spatial extent used only estimates from the all the species/phenophase observations from the NPN datasets, thus representing a large spatial extent.
-The second was from all observations from the LTS datasets, representing a single site.
+The first spatial extent used estimates from a the localities of the respective species in the NPN dataset, thus representing a large spatial extent.
+The second used estimates from the localities of the LTS datasets, thus representing a single site.
 The RMSD does not consider actual observations as it is meant to infer the correspondence of the NPN and LTS datasets beyond parameter comparison.
 
-For all models the actual error was calculated using the 20\% of held out data.
+For all models the actual error was calculated using the using the RMSE:
 
 $$ RMSE = \sqrt[]{ \frac{\sum_{i=1}^{n}(\widehat{DOY_{i}} - DOY_{i})^{2}}{n}} $$
+
+For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, held out NPN and LTS observations, and the six phenology models.
 
@@
 All parameters had statistically different distributions between LTS and NPN derived estimates when tested with a Mann-Whitney U test.
-When comparing the mean values of parameters among all species and phenophases there was correspondence among the simple models with 1-2 parameters (Fig.
+Visually comparing the mean values of parameters among all species and phenophases showed a correspondence among the simple models with 1-2 parameters (Fig.
 1).
@@
 The Naive model showed a distinct late bias in the mean DOY for phenologcial events, likely resulting from the LTS datasets being mostly in the northern United States.
-The Fixed GDD model had the best correspondence between the two datasets.
+Though it still has a small bias the Fixed GDD model had the best correspondence between the two datasets (see suppliment for zoomed in figure).
 The outlier for this model, \textit{Larrea tridentata}, has phenology largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
@@
 2b).
-When estimating across the spatial extent of NPN data, except when using the Linear and Naive model, most NPN models produced estimates within 20 days of LTS models (Fig.
+When estimating across the spatial extent of NPN data, except when using the Linear and Naive model, most NPN and LTS models produced estimates within 20 days of each other (Fig.
 2a).
@@
 Models predicting NPN observations (Fig.
-3a) has higher error rates than predictions for LTS observations (Fig.
+3a) have higher error rates than predictions for LTS observations (Fig.
 3b).
@@
 Figure 4 shows the pairwise comparison of LTS and NPN errors.
-Similar to Figure 3, the NPN modeld performed best on NPN observations and LTS models performed best on LTS observations.
+Similar to Figure 3, the NPN models performed best on NPN observations and LTS models performed best on LTS observations.
+
 \begin{figure}
@@
 For example two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 can produce nearly identical results in many, but not all, scenarios.
-As model complexity icnreases the ability to produce similar estimates increases.
+As model complexity increases the ability to produce similar estimates increases.
 Indeed, the Uniforic models with NPN derived parameters still had estimates within 10 days of LTS derived parameters for most species (Fig.
@@
 Here our simplest model, the Fixed GDD model which uses only a warming component, had reasonable (thought statistically different) correspondence between LTS and NPN datasets.
-This correspondence decreases as more parameters are added to models, suggesting only simple models should be used when using the NPN dataset to make inferences about plant physiology.
+This correspondence decreases as more parameters are added to models, suggesting that if the NPN dataset is being used to make inferences about plant physology simple models could give a more consistent results.
 Parameter estimates of more complex models do not agree with models from the same species in LTS datasets, and it is possible even the LTS datasets are over fitting these models.
@@
 While the values of additional parameters may be incorrect in the above sense, they nonetheless provide more predictive power.
+Across the two spatial scales the GDD, Uniforc, and Alternating model were generally the best performing (Fig.
+3).
 Considering only NPN derived models predicting out of sample NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1).
+One goal of ecological forecasting is to inform decision makers about the future state of systems (dietz?).
+Thus the best performing model should be used in these cases to give the best possible forecast.
 The GDD and Uniforc models have three and four parameters, respectively.
@@
 Other models show promise for forecasting under different scenarios.
-For example if using data from a large spatial extent to make predictions at single site our results show the Fixed GDD model is the best option, as performance under this scenario is very similar to performance of Fixed GDD models built from LTS data (Fig.
+For example if using data from a large spatial extent to make predictions at single sites our results show the Fixed GDD model is the best option, as performance under this scenario is very similar to performance of Fixed GDD models built from LTS data (Fig.
 4).
-In the opposite scenario of making large scale predictions using only data from a single site the Alternating is likely the best option, as most LTS derived models have a RMSE only 5 days greater on average than the corresponding NPN model when tested on out of sample NPN data (Fig.
+In the opposite scenario of making large scale predictions using only data from a single site the Alternating model is likely the best option, as most LTS derived models have a RMSE only 5 days greater on average than the corresponding NPN model when tested on out of sample NPN data (Fig.
 4).
@@
 
-Conclusion
+The spatially corrected NPN models were essentially identical to the non-corrected ones in terms of parameters, estimates of phenological events, and out of sample error rates.
+Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, yet models which attempt to model this variation did not significatnly improve estimates
+
+\section*{Conclusion}
 Despite being collected across a large spatial extend and by hundreds of observers the NPN dataset can be used to build phenology models which make comparable predictions to LTS datasets.
 Indeed, if making predictions across a large spatial extent the NPN derived models are preferable to LTS derived models.
-If using the NPN dataset to make inferences about plant physiology researchers should take not to use overly complicated models, as parameter estimates from complex NPN derived models do not agree with LTS derived models.
+If using the NPN dataset to make inferences about plant physiology researchers should not to use overly complicated models, as parameter estimates from complex NPN derived models do not agree with LTS derived models.
 Whether this disagreement stems from the spatial variation in plant phenology requirements, or in the irregular sampling of the NPN dataset is an avenue for future study.
+
+The National Phenology Network dataset will likely be an important resource for studying phenology.
+The addition of numerous species and observations from across a range of conditions is important for informing large scale processes, as similar datasets have shown in Europe \citep{olsson2014, basler2016} and Asia \citep{xu2013, zhang2017}.
+Research from long term small scale studies has been invaluable and should no doubt be continued, but can now be complimented by using large scale NPN data (ie.
+\citep{jeong2013, melaas2016}).
+As long term phenology observations from the National Ecological Observatory Network \citep{elmendorf2016} become available in the coming years  ....
 

### 2018-01-17 manuscript/main.tex
Spatial model analysis (#35)
-add spatial models (MSB and M1) to figures
-add map showing all the sites
-update manuscript intro structure to ethans suggestions

@@
 
-%% \title{Testing the usability of large scale citizen science data for phenology modeling}
 \title{Comparison of large scale citizen science data and long term study data for phenology modeling}
+
 \author[1]{Shawn D. Taylor}
@@
 
-Citizen science data is becoming increasingly important for ecological research, with the potential for significant impact in many fields \citep{dickinson2010, tulloch2013, kelling2009}.
-Citizen science efforts have helped document range shifts \citep{hitch2007}, study the effect of landscape fragmentation on amphibians \citep{cosentino2014}, and informed ecological theory of community composition \citep{locey2013}.
-A primary benefit of citizen science data is observations made at numerous sites across regional to continental scales.
-This helps researchers explain large scale patterns across natural gradients, but can also introduce new sources error because of observer skill, variation in sampling effort, and spatial bias \citep{dickinson2010}.
-These issues can be accounted for in models but first must be identified and quantified.
-As new citizen science datasets are made available, they must be evaluated for their own biases and limitations.
-
-The National Phenology Network (NPN) is an organization which accepts observations of phenology from volunteers throughout North America and makes the data freely available \citep{schwartz2012a}.
-Plant phenology is important for studies of ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}.
+Plant phenology affects many fields of ecological research such as ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}.
 Robust phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant and animal communities.
 At the largest scale variation in the timing of spring leaf out and fall senescence account for a significant amount of error in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long term forecasts \citep{richardson2012}.
-At local scales, because of species specific responses to temperature and precipitation, changes in phenology can alter the flower community available to pollinators and seed predators throughout the growing season \citep{diez2012, caradonna2014, ogilvie2017, theobald2017}.
+At local scales, because of species specific responses to temperature and precipitation, changes in phenology can alter the flower community available to pollinators and seed predators throughout the growing season, potentially causing a cascade of affects on abundance, species richness, and interspecific interactions.
+\citep{diez2012, caradonna2014, ogilvie2017, theobald2017}.
 
+Many phenological studies in the past have relied upon datasets collected at a single location over a long time-frame by a single group \citep{cook2012, roberts2015, iler2013, wolkovich2012}.
+These datasets have high temporal coverage, with some approaching several decades, and represent the species and climate variability at their respective sites extremely well.
+Yet it is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}.
+Phenology models are species specific, thus this lack of transferability across space can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995,diez2012}.
+Models fit to only a single site can easily be overfit, and being able to generalize across a species range will be important for improving phenology forecasts \citep{richardson2013}.
+To accomplish this phenology observations will be needed from as much of the full range of species as possible.
+
+The National Phenology Network (NPN) is an organization which accepts phenology observations from volunteers throughout North America and makes the data freely available \citep{schwartz2012a}.
+Citizen science data such as this is becoming increasingly important for ecological research, with the potential for significant impact in many fields \citep{dickinson2010, tulloch2013, kelling2009}.
+Citizen science efforts have helped document range shifts \citep{hitch2007}, study the effect of landscape fragmentation on amphibians \citep{cosentino2014}, and informed ecological theory of community composition \citep{locey2013}.
 The NPN dataset has already being used to study, among many things, variation in oak phenology \citep{gerst2017}, large scale community phenology models \citep{melaas2016}, and forecasting long term phenology trends \citep{jeong2013}.
-While there has been several studies looking at best practices when using this dataset \citep{crimmins2017, gerst2016} no study to date has looked a the underlying validity of the data.
+A primary benefit of citizen science data is observations made at numerous sites across regional to continental scales.
+This helps researchers explain large scale patterns across natural gradients, but can also introduce new sources error from observer skill, variation in sampling effort, and spatial bias \citep{dickinson2010}.
+These issues can be accounted for in models but first must be identified and quantified.
+While there has been several studies looking at best practices when using the NPN dataset \citep{crimmins2017, gerst2016} no study to date has looked a the underlying validity of the data.
 With hundreds of observers across North America, the observer and sampling bias for species identification and the true date of phenological events is potentially high.
-Though volunteers can be very accurate at distinguishing different phenophases \citep{fuccillo2015} observations can still be made sporadically throughout the growing season, some years skipped, or visits to a particular site discontinued entirely.
-These biases could adversely affect process based phenology models, which are species specific and depend on high precision in the data \citep{chuine2000}.
+Though volunteers can be very accurate at distinguishing different phenophases \citep{fuccillo2015}, observations can still be made sporadically throughout the growing season, some years skipped, or visits to a particular site discontinued entirely.
 
-The NPN dataset is complicated further by the spatial variation of the temperature response within species.
-Phenology models currently have trouble capturing the spatial variation in warming requirements \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}.
-Most phenology models are process based and designed to reflect physological processes driving phenological events.
-Yet models parameterized using observations from one site rarely transfer to other sites, and models using pooled data from many sites do not transfer well to individual sites \citep{garcia-mozo2008, basler2016}.
-This variation within species can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995,diez2012}.
-While this complicates further any testing of dataset validity, it's worth it because there is a large opportunity with this dataset for further study of phenology across spatial gradients.
-
-Many phenological studies in the past have relied upon datasets with observations collected at a single location over a long timeframe by a single group \citep{cook2012, roberts2015, clark2014}.
-These datasets are assumed to not suffer from the potential biases mentioned above, and many also have high temporal coverage of several decades.
-They are disadvantageous as the inference from the models can only be applied to species located at the site, and only to local climate variability.
-These datasets have been invaluable for ecology for several decades, but datasets with a larger spatial extent and more complete species pool will be needed to model the complex phenology dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}.
-The NPN datasets has the potential to meet these needs, and to date no study has assessed how well NPN data compares to long term study phenology datasets.
+Data from long term studies have been invaluable for phenology research for several decades, but datasets with a larger spatial extent and more complete species pool will be needed to model the complex phenology dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}.
+Such datasets from China and Europe have already contributed considerably to phenological research \citep{olsson2014, basler2016, xu2013, zhang2017}, and the NPN dataset has the potential to meet these needs for North America.
+To date no study has assessed how well the NPN dataset compares to long term study phenology datasets.
 Here we will compare process based phenology models built using only the NPN data with models built with data from Long Term Study (LTS) datasets.
@@
 Three major ecosystem types and 24 species are represented in these datasets.
-As all four have been collected by single groups we make the assumption that they are free of any bias in species identification or sampling effort.
 Observation metrics varied among these four datasets due to different protocols, so were first converted to binary "yes" and "no" observations for each phenophase (see supplement for details).
@@
 - March) temperature as the only independent variable and DOY as the response variable.
+Finally two models attempt to explain the spatial variation in phenological requirements.
+The first spatial model, M1, is an extension of the GDD model which adds a correction in the required forcing using the photoperiod ($L$).
+The second, the Macroscale Species-specific Budburst model (MSB), uses the mean spring temperature as a linear correction on the total forcing required in the Alternating model.
+Since there is little to no spatial variation in the LTS datasets, the two spatial models were fit using only the NPN dataset.
+The resulting parameters, estimates, and errors for the NPN derived M1 and MSB models were compared to the LTS GDD and Alternating models, respectively, for each species and phenophase.
 
@@
     Naive & \( \overline{DOY} \) & - & 1 & - \\
-    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(x_{t})\geq F* $  & $R_{f}(x_{t}) = min(T, 0)$ & 1 & - \\
+    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = min(T_{i}, 0)$ & 1 & - \\
     Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
-    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(x_{t})\geq F* $ & $ R_{f}(x_{t}) = max(T - T^{*}, 0) $  & 3 & - \\
-    Alternating & $\sum_{t=0}^{DOY}R_{f}(x_{t})\geq a + be^{cNCD(t)} $ & $R_{f}(x_{t}) = min(T, 0) $ & 3 & \citep{cannell1983} \\
-    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(x_{t})\geq F* $ & $ R_{f}(x_{t}) = \frac{1}{1 + e^{b(x-c)}} $ & 4 & \citep{chuine2000} \\
+    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T^{*}, 0) $  & 3 & - \\
+    M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T^{*}, 5) $  & 4 & \citep{blumel2012} \\
+    Alternating & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD(t)} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 3 & \citep{cannell1983} \\
+    MSB & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD_{i}} +dT_{mean} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 4 & \citep{jeong2013} \\
+    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = \frac{1}{1 + e^{b(T_{i}-c)}} $ & 4 & \citep{chuine2000} \\
 
@@
 
-Parameterization was done using differential evolution to minimize the root mean square error (RMSE) of the estimated day of year of the phenological event.
+All models were parameterized using differential evolution to minimize the root mean square error (RMSE) of the estimated day of year of the phenological event.
 Differential evolution is a global optimization algorithm similar to simulated annealing, which has traditionally been used to fit complex phenology models \citep{storn1997, chuine2000}.
@@
 A random 20\% of all data was held out from model fitting for use in later evaluation.
+Predictions were made by taking the mean DOY from the 250 bootstrapped models.
 
@@
 
-As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set parameterized using only LTS data.
-The distribution for each parameter in each model was then compared between the two sets using a Mann-Whitney test.
+As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set parameterized using only LTS data (with the exception of the M1 and MSB models, see above).
+The distribution for each parameter in each species/phenophase model was then compared between the two sets using a Mann-Whitney test.
 This was chosen due to the majority of parameters having non-normal distributions derived from the bootstrapped estimates.
-Statistically significant differences in these parameters, which are designed to represent biological mechanisms, would imply that the LTS and NPN datasets do not agree on
 
-Even models with different structures can produce similar estimates of phenological events \citep{basler2016}.
-Thus we also compared estimates between LTS and NPN derived models using the root mean square difference (RMSD) between the two estimates:
+Even models with entirely different structures can produce similar estimates of phenological events \citep{basler2016}.
+Thus we also compared estimates of the phenological events between LTS and NPN derived models using the root mean square difference (RMSD) between the two estimates:
 
@@
 where $n$ is the number of estimates produced for a single species and phenophase, and $LTS_{i}$ and $NPN_{i}$ are DOY estimates produced from the respective models.
-A RMSD value close to 0 implies that the LTS and NP derived models produced very similar estimates.
-For each species and phenophase 12 RMSD values were calculated from the 6 phenology models as well as 2 spatial extents.
-The first spatial extent used estimates from a the localities of the respective species in the NPN dataset, thus representing a large spatial extent.
-The second used estimates from the localities of the LTS datasets, thus representing a single site.
+A RMSD value close to 0 implies that the LTS and NPN derived models produced very similar estimates.
+For each species and phenophase 12 RMSD values were calculated from the 6 phenology models and 2 scales.
+We calculated the RMSD for estimates at the first scale of the extent of the NPN observations, thus representing a large spatial extent.
+The second scale used estimates from the localities of the LTS datasets, thus representing a single site.
 The RMSD does not consider actual observations as it is meant to infer the correspondence of the NPN and LTS datasets beyond parameter comparison.
 
-For all models the actual error was calculated using the using the RMSE:
+For all models the observation error of the held out data was calculated using the using the RMSE:
 
@@
 
-For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, held out NPN and LTS observations, and the six phenology models.
+For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, the six phenology models, and the two scales described above.
 
-Pairwise model error comparison
+TODO: Pairwise model error comparison
 
@@
 
+TODO: figure 4 results
+
 \begin{figure}
@@
 \end{figure}
-
--Even with different parameter values, some species/models had extremely similar
-estimates, suggesting these models had
 
@@
 While the values of additional parameters may be incorrect in the above sense, they nonetheless provide more predictive power.
-Across the two spatial scales the GDD, Uniforc, and Alternating model were generally the best performing (Fig.
+Across the two scales the GDD, Uniforc, and Alternating model were generally the best performing (Fig.
 3).
-Considering only NPN derived models predicting out of sample NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1).
-One goal of ecological forecasting is to inform decision makers about the future state of systems (dietz?).
+Considering only NPN derived models predicting NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1).
+One goal of ecological forecasting is to inform decision makers about the future state of systems \citep{dietz2017}.
 Thus the best performing model should be used in these cases to give the best possible forecast.
-The GDD and Uniforc models have three and four parameters, respectively.
-Thus the simpler GDD model shows promise for building models with the NPN dataset and making forecasts across large scales.
+The GDD and Uniforc models have three and four parameters, respectively, therefore the simpler GDD model shows promise for building models with the NPN dataset and making forecasts across large scales.
 
@@
 
-The spatially corrected NPN models were essentially identical to the non-corrected ones in terms of parameters, estimates of phenological events, and out of sample error rates.
-Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, yet models which attempt to model this variation did not significatnly improve estimates
+The models which incorporate a spatial component did not improve over their base models.
+Correspondence between parameter estimates, estimates of phenological events, and out of sample error rates for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively.
+The MSB and M1 models had the lowest error in 9 out of 39 occasions, but never by more than a single day over other models (Fig S1).
+Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, yet here models which attempt to account for this variation did not improve estimates overall.
+This could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors.
+\cite{basler2016} used the M1 model to predict budburst on six species across Europe and found it was usually among the best models in terms of RMSE, albeit never by more than a single day.
+Their result was strengthened by a 40 year timeseries across a large region.
 
 \section*{Conclusion}
-Despite being collected across a large spatial extend and by hundreds of observers the NPN dataset can be used to build phenology models which make comparable predictions to LTS datasets.
-Indeed, if making predictions across a large spatial extent the NPN derived models are preferable to LTS derived models.
-If using the NPN dataset to make inferences about plant physiology researchers should not to use overly complicated models, as parameter estimates from complex NPN derived models do not agree with LTS derived models.
-Whether this disagreement stems from the spatial variation in plant phenology requirements, or in the irregular sampling of the NPN dataset is an avenue for future study.
+
+Currently the best phenology model for a single location will always be those fit using data from only that location, yet these single site models do not transfer well to large scales.
+Making large scale phenology forecasts will require incorporating large scale observations such as those from the NPN dataset.
+Here we have shown that the NPN dataset is capable of parameterizing phenology models which make comparable predictions to LTS datasets.
+Indeed, if making forecasts across a large spatial extent the NPN derived models are preferable to LTS derived models.
+If using the NPN dataset to make inferences about plant physiology researchers should not use overly complicated models, as parameter estimates from complex NPN derived models do not agree with LTS derived models.
+Whether this disagreement stems from the spatial variation in plant phenology requirements, the irregular sampling of the NPN dataset, or the non-identifiability of models is an avenue for future study.
 
@@
 \citep{jeong2013, melaas2016}).
-As long term phenology observations from the National Ecological Observatory Network \citep{elmendorf2016} become available in the coming years  ....
 
@@
 
+\setcounter{figure}{0}    
+\section*{Suppliment}
+\begin{figure}
+	\centering
+		\includegraphics[width=1\textwidth]{fig_s1_best_npn_models.png}
+	\caption{Model error for specific species and phenophases of the NPN dataset. X marks the best performing models for the respective data type.}
+\end{figure}
+
+\begin{figure}
+	\centering
+		\includegraphics[width=1\textwidth]{fig_s3_site_map.png}
+	\caption{Locations of National Phenology Network sites used (black points) and Long Term Study sites (labeled circles).}
+\end{figure}
+
 \end{document}

### 2018-01-17 analysis/compare_parameters.R

@@ -30,7 +30,34 @@ lts_sample_sizes = all_parameters %>%
   summarize(n_species = n_distinct(species))
 print(paste0('total species/phenophase comparisons: ',sum(lts_sample_sizes$n_species)))
 
+############################################################################
+# Organize the two spatial models, MSB and M1
+# These models add a correction to the Alternating and GDD models, respectivaely,
+# to hopefully account for spatial variation. Since fitting these models to LTS
+# sites doesn't make sense (as there is no spatial variation in them ) I'll
+# compare them to LTS datasets fitted to the original Alternating and GDD models.
 
+# Remove spatial models fit to LTS data
+all_parameters = all_parameters %>%
+  filter(!(dataset!='npn' & model %in% c('msb','m1')))
+
+# Copy the LTS GDD and Alternating model to compare with the
+# corrected NPN ones.
+lts_models_parameters = all_parameters %>%
+  filter(dataset!='npn', model %in% c('gdd','alternating'))
+lts_models_parameters$model = with(lts_models_parameters, ifelse(model=='gdd','m1',
+                                                                 ifelse(model=='alternating','msb','unk')))
+if(any(lts_models_parameters$model=='unk')){stop('unknown model in lts subset')}  
+
+all_parameters = all_parameters %>%
+  bind_rows(lts_models_parameters)
+
+# Remove the additional parameters which are only in the corrected models
+all_parameters = all_parameters %>%
+  filter(!(model=='msb' & parameter_name=='d')) %>%
+  filter(!(model=='m1' & parameter_name=='k'))
+
+rm(lts_models_parameters)
 ############################################################################
 #The distribution of all parameters derived using bootstrapping
 make_parameter_histograms = FALSE
@@ -122,7 +149,7 @@ alternating=ggplot(filter(parameter_means, model=='alternating'), aes(x=npn, y=p
   facet_wrap(~parameter_name, scales='free', nrow=1) + 
   theme_bw() +
   theme(legend.position = "none") +
-  labs(y = "Alternating", x='') +
+  labs(y = "Alternating\n", x='') +
   common_plot_theme 
[…]

### 2018-01-17 analysis/observation_error_analysis.R

@@ -35,9 +35,25 @@ predictions = predictions %>%
   rename(phenophase = phenophase_type)
 
 #########################################################
-# # Leave out the new models for now
+# Account for spatially corrected models. See note in
+# compare_parameters.R
+
+# Remove spatial models fit to LTS data
 predictions = predictions %>%
-  filter(!model_name %in% c('m1','msb'))
+  filter(!(parameter_source!='npn' & model_name %in% c('msb','m1')))
+
+# Copy the LTS GDD and Alternating model to compare with the
+# corrected NPN ones.
+lts_models = predictions %>%
+  filter(parameter_source!='npn', model_name %in% c('gdd','alternating'))
+lts_models$model_name = with(lts_models, ifelse(model_name=='gdd','m1',
+                                                ifelse(model_name=='alternating','msb','unk')))
+if(any(lts_models$model_name=='unk')){stop('unknown model in lts subset')}  
+
+predictions = predictions %>%
+  bind_rows(lts_models)
+
+rm(lts_models)
 
 
 ##########################################################
@@ -51,8 +67,8 @@ model_errors = predictions %>%
   mutate(is_lts_model = parameter_source!='npn', is_lts_obs = observation_source != 'npn')
 
 # Apply more pleasing names to everything for figures
-model_names = c('gdd','gdd_fixed','linear_temp','naive','uniforc','alternating','m1','msb')
-pretty_model_names = c('GDD','Fixed GDD','Linear','Naive','Uniforc','Alternating','M1','MSB')
+model_names = c('gdd','m1','gdd_fixed','linear_temp','naive','alternating','msb','uniforc')
+pretty_model_names = c('GDD','M1','Fixed GDD','Linear','Naive','Alternating','MSB','Uniforc')
 datasets = c('harvard','hjandrews','hubbard','jornada','npn')
 pretty_dataset_names = c('Harvard Forest','H.J. Andrews','Hubbard Brook','Jornada','NPN')
 
@@ -128,7 +144,7 @@ bottom_row_text = 'B. Model error for observations in LTS datasets'
 fig3 = plot_grid(top_row, bottom_row, legend_alone, ncol=1, labels=c(top_row_text, bottom_row_text, ''), 
           rel_heights = c(1,1,0.2), hjust=-0.07, vjust=2.7, label_size=12)
 
-ggsave(fig3, filename = 'manuscript/fig_3_error_compare.png', width = 48, height = 24, units = 'cm')
[…]

### 2018-01-17 analysis/observation_estimate_comparison.R

@@ -32,10 +32,25 @@ predictions = predictions %>%
   rename(phenophase = phenophase_type)
 
 #########################################################
-# # Leave out the new models for now
-predictions = predictions %>%
-  filter(!model_name %in% c('m1','msb'))
+# Account for spatially corrected models. See note in
+# compare_parameters.R
 
+# Remove spatial models fit to LTS data
+predictions = predictions %>%
+  filter(!(parameter_source!='npn' & model_name %in% c('msb','m1')))
+
+# Copy the LTS GDD and Alternating model to compare with the
+# corrected NPN ones.
+lts_models = predictions %>%
+  filter(parameter_source!='npn', model_name %in% c('gdd','alternating'))
+lts_models$model_name = with(lts_models, ifelse(model_name=='gdd','m1',
+                                                ifelse(model_name=='alternating','msb','unk')))
+if(any(lts_models$model_name=='unk')){stop('unknown model in lts subset')}  
+
+predictions = predictions %>%
+  bind_rows(lts_models)
+
+rm(lts_models)
 ############################################################
 
 npn_model_estimates = predictions %>%
@@ -66,8 +81,8 @@ estimate_differences_npn_observations = observation_estimate_comparison %>%
 # Graph of RMSD model, phenophase, lts dataset
 
 # Apply more pleasing names to everything for figures
-model_names = c('gdd','gdd_fixed','linear_temp','naive','alternating','uniforc','m1','msb')
-pretty_model_names = c('GDD','Fixed GDD','Linear','Naive','Alternating','Uniforc','M1','MSB')
+model_names = c('gdd','m1','gdd_fixed','linear_temp','naive','alternating','msb','uniforc')
+pretty_model_names = c('GDD','M1','Fixed GDD','Linear','Naive','Alternating','MSB','Uniforc')
 datasets = c('harvard','hjandrews','hubbard','jornada','npn')
 pretty_dataset_names = c('Harvard Forest','H.J. Andrews','Hubbard Brook','Jornada','NPN')
 
@@ -86,6 +101,8 @@ common_theme_elements = theme(axis.text = element_text(size=18),
                               axis.title.x = element_text(size=20),
                               panel.grid.major.y = element_line(colour = "grey80", size=0.5),
                               panel.grid.minor.y = element_line(colour = "grey85", size=0.5))
+
+set.seed(7)
 
[…]

### 2018-01-17 manuscript_map.R

@@ -52,6 +52,6 @@ site_map = ggplot() +
         panel.border = element_blank(),
         legend.position = 'none')
 print(site_map)
-ggsave('manuscript/site_map.png', plot=site_map, height=20, width=50, units = 'cm')
+ggsave('manuscript/fig_s3_site_map.png', plot=site_map, height=20, width=50, units = 'cm')
 
 

### 2018-01-26 manuscript/main.tex
first draft sent to coauthors

@@
 \author[1]{Shawn D. Taylor}
-\author[2]{Ethan P. White}
-\affil[1]{UFL}
-\affil[2]{UFL}
+\author[2]{Joan M. Meiners}
+\author[3]{Michael Orr}
+\author[4]{Kristina Riemer}
+\author[5]{Ethan P. White}
+
+\affil[1,2]{UFL SNRE}
+\affil[3]{???}
+\affil[4,5]{UFL WEC}
+\affil[5]{UFL Bioinformatics}
 \corrauthor[1]{Shawn D. Taylor}{shawntaylor@weecology.org}
@@
 Plant phenology affects many fields of ecological research such as ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}.
-Robust phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant and animal communities.
+Robust plant phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant and animal communities.
 At the largest scale variation in the timing of spring leaf out and fall senescence account for a significant amount of error in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long term forecasts \citep{richardson2012}.
@@
 
-Many phenological studies in the past have relied upon datasets collected at a single location over a long time-frame by a single group \citep{cook2012, roberts2015, iler2013, wolkovich2012}.
+Many plant phenological studies in the past have relied upon datasets collected at a single location over a long time-frame by a single group \citep{cook2012, roberts2015, iler2013, wolkovich2012}.
 These datasets have high temporal coverage, with some approaching several decades, and represent the species and climate variability at their respective sites extremely well.
@@
 Data from long term studies have been invaluable for phenology research for several decades, but datasets with a larger spatial extent and more complete species pool will be needed to model the complex phenology dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}.
-Such datasets from China and Europe have already contributed considerably to phenological research \citep{olsson2014, basler2016, xu2013, zhang2017}, and the NPN dataset has the potential to meet these needs for North America.
+Such datasets from China and Europe have already contributed considerably to phenological research \citep{olsson2014, basler2016, xu2013, zhang2017}, and the NPN dataset has the potential to meet these needs for North America plant species and communities.
 To date no study has assessed how well the NPN dataset compares to long term study phenology datasets.
@@
 The Uniforc model is similar to the GDD model but has forcing units derived a sigmoid function.
-
+ \citep{chuine2000}
 The naive model uses only the mean DOY from all prior observations as the estimated DOY.
@@
 Finally two models attempt to explain the spatial variation in phenological requirements.
-The first spatial model, M1, is an extension of the GDD model which adds a correction in the required forcing using the photoperiod ($L$).
-The second, the Macroscale Species-specific Budburst model (MSB), uses the mean spring temperature as a linear correction on the total forcing required in the Alternating model.
+The first spatial model, M1, is an extension of the GDD model which adds a correction in the required forcing using the photoperiod ($L$) \citep{blumel2012}.
+The second, the Macroscale Species-specific Budburst model (MSB), uses the mean spring temperature as a linear correction on the total forcing required in the Alternating model \citep{jeong2013}.
 Since there is little to no spatial variation in the LTS datasets, the two spatial models were fit using only the NPN dataset.
@@
 As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set parameterized using only LTS data (with the exception of the M1 and MSB models, see above).
-The distribution for each parameter in each species/phenophase model was then compared between the two sets using a Mann-Whitney test.
+The distribution for each parameter in each species/phenophase model was compared between the two sets using a Mann-Whitney test.
 This was chosen due to the majority of parameters having non-normal distributions derived from the bootstrapped estimates.
@@
 
-For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, the six phenology models, and the two scales described above.
+For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, the eight phenology models, and the two scales described above.
 
-TODO: Pairwise model error comparison
+Finally we explored which set of models, either LTS or NPN derived, performed better overall at each scale.
+We calculated the pairwise difference of RMSE between LTS and NPN derived models for each model type, species, and phenophase at the two different scales.
+We plotted the density function
 
-$$ RMSE_{difference} = \widehat{RMSE_{NPN,s}} - RMSE_{LTS,s} $$
+$$ f(x) = RMSE_{NPN} - RMSE_{LTS} $$
 
-for all $$S = s_{1},...,s_{39}$$ species
+for each species and phenophases  $$\{ f(x) | s_{1},...,s_{39}\} $$
+
+for each model type and scale.
 
@@
 Each point is, for a single species/phenophase, the root mean square difference between LTS and NPN derived estimates for a specified model.
-A shows the RMSD over the extent of the NPN dataset, B shows the RMSD at the single sites represented by the LTS datasets.}
+A shows the RMSD over the extent of the NPN dataset (a large scale), B shows the RMSD at the single sites represented by the LTS datasets (a small scale).}
 \end{figure}
@@
 
-Figure 3 shows the errors for all different model types when tested against the held out data.
+Figure 3 shows the errors for all different model types using the held out data.
 Models predicting NPN observations (Fig.
@@
 
-TODO: figure 4 results
+Figure 4 shows the pairwise comparison of specific species/phenophase combinations for held out data.
+Overall NPN derived models performed best on held out NPN observations, while LTS derived models performed best on held out LTS observations.
+The naive and linear models had the largest differences between the two models sets, while the Fixed GDD model had relatively similar errors when evaluated on both NPN and LTS held out observations.
 
@@
 Despite having statistically different parameters to LTS derived models, many NPN derived models were able to produce similar estimates for phenological events.
-This is (likely) due to models having a non-linear parameter space and being non-identifiable.
+This is likely due to models having a non-linear parameter space and being non-identifiable.
 For example two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 can produce nearly identical results in many, but not all, scenarios.
@@
 Considering only NPN derived models predicting NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1).
-One goal of ecological forecasting is to inform decision makers about the future state of systems \citep{dietz2017}.
+One goal of ecological forecasting is to inform decision makers about the future state of systems \citep{dietz2018}.
 Thus the best performing model should be used in these cases to give the best possible forecast.
-The GDD and Uniforc models have three and four parameters, respectively, therefore the simpler GDD model shows promise for building models with the NPN dataset and making forecasts across large scales.
+The GDD and Uniforc models show promise for building models with the NPN dataset and making plant phenology forecasts across large scales.
 
@@
 The models which incorporate a spatial component did not improve over their base models.
-Correspondence between parameter estimates, estimates of phenological events, and out of sample error rates for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively.
+Correspondence between parameter estimates (Fig.
+1), estimates of phenological events (Fig.
+2), and out of sample error rates (Fig.
+3 & 4) for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively.
 The MSB and M1 models had the lowest error in 9 out of 39 occasions, but never by more than a single day over other models (Fig S1).
@@
 Their result was strengthened by a 40 year timeseries across a large region.
+\cite{chuine2017} noted that incorporating the spatial vaiation in warming requirements as a key issue in future phenology research.
+We feel the NPN dataset, as well as other large scale phenology datasets, will be key in addressing this and other phenological research needs.
 

### 2018-04-09 manuscript/main.tex
manuscript update
2nd large draft sent to co-authors. reworked analysis/into just
3 figures. incorporated ethans suggestions on large/small scale
data discussion

@@
 
-\title{Comparison of large scale citizen science data and long term study data for phenology modeling}
+\title{Comparison of large-scale citizen science data and long-term study data for phenology modeling}
 
@@
 \author[2]{Joan M. Meiners}
-\author[3]{Michael Orr}
-\author[4]{Kristina Riemer}
+\author[3]{Kristina Riemer}
+\author[4]{Michael C. Orr}
 \author[5]{Ethan P. White}
@@
 \affil[1,2]{UFL SNRE}
-\affil[3]{???}
-\affil[4,5]{UFL WEC}
+\affil[4]{Key Laboratory of Zoological Systematics and Evolution, Institute of Zoology, Chinese Academy of Sciences, Beijing 100101, P.R. China}
+\affil[3,5]{UFL WEC}
 \affil[5]{UFL Bioinformatics}
@@
 
-Plant phenology affects many fields of ecological research such as ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}.
-Robust plant phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant and animal communities.
-At the largest scale variation in the timing of spring leaf out and fall senescence account for a significant amount of error in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long term forecasts \citep{richardson2012}.
-At local scales, because of species specific responses to temperature and precipitation, changes in phenology can alter the flower community available to pollinators and seed predators throughout the growing season, potentially causing a cascade of affects on abundance, species richness, and interspecific interactions.
-\citep{diez2012, caradonna2014, ogilvie2017, theobald2017}.
+Robust plant phenology models are necessary for forecasting the future state of the ecological systems at several scales, from global climate to impacts on plant and animal communities.
+Plant phenology performs an important role in many fields of ecological research such as ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}.
+At the largest scale, variation in the timing of spring leaf out and fall senescence account for a significant amount of uncertainty in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long-term forecasts \citep{richardson2012}.
+At local scales, because of species specific responses to temperature and precipitation, changes in flowering times can alter the flower community throughout the growing season \citep{caradonna2014, diez2012, theobald2017}.
+This change in resource availability can affect the abundance and richness of pollinators \citep{ogilvie2017a, ogilvie2017b}, and higher trophic levels \citep{tylianakis2008}.
+Accurately accounting for phenology in these and other scenarios will be critical for forecasting environmental change.
 
+To best forecast plant phenology researchers will need models which account for the full range of variation across a species range \citep{richardson2013, chuine2017}.
 Many plant phenological studies in the past have relied upon datasets collected at a single location over a long time-frame by a single group \citep{cook2012, roberts2015, iler2013, wolkovich2012}.
@@
 Yet it is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}.
-Phenology models are species specific, thus this lack of transferability across space can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995,diez2012}.
-Models fit to only a single site can easily be overfit, and being able to generalize across a species range will be important for improving phenology forecasts \citep{richardson2013}.
-To accomplish this phenology observations will be needed from as much of the full range of species as possible.
+This lack of transferability across space can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995, diez2012}.
+Future models which account for this variability will require data from as much of a species' range as possible.
+Further, data from a single location may not be adequate for ecosystem level phenology modelling due to a potentially limited species pool.
+Thus, new sources of data beyond traditional single site studies will be needed.
 
-The National Phenology Network (NPN) is an organization which accepts phenology observations from volunteers throughout North America and makes the data freely available \citep{schwartz2012a}.
-Citizen science data such as this is becoming increasingly important for ecological research, with the potential for significant impact in many fields \citep{dickinson2010, tulloch2013, kelling2009}.
+Citizen science data is becoming increasingly important for ecological research, with the potential for significant impact in many fields \citep{dickinson2010, tulloch2013, kelling2009}.
 Citizen science efforts have helped document range shifts \citep{hitch2007}, study the effect of landscape fragmentation on amphibians \citep{cosentino2014}, and informed ecological theory of community composition \citep{locey2013}.
-The NPN dataset has already being used to study, among many things, variation in oak phenology \citep{gerst2017}, large scale community phenology models \citep{melaas2016}, and forecasting long term phenology trends \citep{jeong2013}.
-A primary benefit of citizen science data is observations made at numerous sites across regional to continental scales.
-This helps researchers explain large scale patterns across natural gradients, but can also introduce new sources error from observer skill, variation in sampling effort, and spatial bias \citep{dickinson2010}.
+A relatively new citizen science group, The National Phenology Network (NPN) is an organization which accepts phenology observations from volunteers throughout North America and makes the data freely available \citep{schwartz2012a}.
+The NPN dataset has already been used to study, among many things, variation in oak phenology \citep{gerst2017}, large-scale community phenology models \citep{melaas2016}, and forecasting long-term phenology trends \citep{jeong2013}.
+The regional to continental scale of citizen science data helps researchers explain large-scale patterns across natural gradients, but can also introduce new sources of error from observer skill, variation in sampling effort, and spatial bias \citep{dickinson2010}.
 These issues can be accounted for in models but first must be identified and quantified.
-While there has been several studies looking at best practices when using the NPN dataset \citep{crimmins2017, gerst2016} no study to date has looked a the underlying validity of the data.
 With hundreds of observers across North America, the observer and sampling bias for species identification and the true date of phenological events is potentially high.
@@
 
-Data from long term studies have been invaluable for phenology research for several decades, but datasets with a larger spatial extent and more complete species pool will be needed to model the complex phenology dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}.
-Such datasets from China and Europe have already contributed considerably to phenological research \citep{olsson2014, basler2016, xu2013, zhang2017}, and the NPN dataset has the potential to meet these needs for North America plant species and communities.
-To date no study has assessed how well the NPN dataset compares to long term study phenology datasets.
-Here we will compare process based phenology models built using only the NPN data with models built with data from Long Term Study (LTS) datasets.
-The specific research questions are:
-\begin{enumerate}
-\item Do parameter estimates of species level models differ when they are built with the NPN dataset versus an LTS dataset?
-\item Do estimates of phenological events differ between the NPN and LTS derived models?
-\item How well do models built with the NPN dataset transfer to LTS observations, and vice verse?
-\end{enumerate}
+While there have been several studies looking at best practices when using the NPN dataset \citep{crimmins2017, gerst2016}, no study to date has compared it directly with data from long-term studies to test its adequacy for plant phenology models.
+Data from long-term studies (LTS) have been invaluable for phenology research for several decades, but datasets with a larger spatial extent and more complete species pools will be needed to model the complex phenological dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}.
+Such datasets from China and Europe have already contributed considerably to phenological research \citep{olsson2014, basler2016, xu2013, zhang2017}, and the NPN dataset has the potential to meet these needs for North American plant species and communities.
+Here, we fit seven different plant phenology models to the budburst and first flowering phenophases of XX plant species.
+The models were fit twice, once using data from the NPN dataset, and again for the same species using data from an LTS dataset.
+We then compared three outputs between the two sets of models built using NPN and LTS data: 1) the models' parameter estimates; 2) the models' estimates of phenophase timing; and 3) the models' errors using held out testing data.
 
@@
 
-The National Phenology Network collections began in 2009 and continue to the present.
-Observations are based on status-based monitoring, where observers answer “yes”, “no”, or “unsure” when asked if a plant has a specific phenological phase present \citep{denny2014}.
-We downloaded all NPN observations from 2009-2016 for the phenophases Breaking Leaf Buds, Breaking Needle Buds, Emerging Needles, and Open Flowers.
-These phenophases apply to deciduous broadleafs, evergreen conifers, pines, and all angiosperms, respectively.
-Hereafter we will refer to these as either Flowers for the Open Flower phenophase, or Leaves for all other phenophases.
+The National Phenology Network observations are based on status-based monitoring, where observers answer yes, no, or unsure when asked if an individual plant has a specific phenophase present \citep{denny2014}.
+Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves.
+We downloaded all NPN observations from 2009, when collections began, to 2016 for the phenophases Breaking Leaf Buds, Breaking Needle Buds, Emerging Needles, and Open Flowers.
+The first three phenophases apply to the leaf out phase for deciduous broadleafs, evergreen conifers, and pines, respectively.
+The Open Flowers phenophase indicates fully open flowers and applies to all angiosperms.
+Hereafter we will refer to these as either Flowers for the Open Flower phenophase, or Budburst for all other phenophases.
+
 We used moderately strict subsetting of NPN phenology observations as outlined in \cite{crimmins2017}.
 First, "yes" observations for individual plants were only kept if they were preceded by a "no" observation within 30 days.
-Observations for Leaves past day of year (DOY) 172 and for Flowers past DOY 213 were dropped.
-Finally only species that had greater than 30 total observations were kept.
-\cite{crimmins2017} only kept observations that were preceded by a "no" within 15 days, and also grouped multiple observations at single sites in a year to a single observations.
+Observations for Budburst past day of year (DOY) 172 and for Flowers past DOY 213 were dropped to minimize any influence from outliers.
+Finally, only species that had greater than 30 total observations were kept.
+\cite{crimmins2017} only kept observations that were preceded by a "no" within 15 days, and also grouped multiple individuals at single sites to a single observation.
 We used 30 days to allow for a greater number of species to be compared.
@@
 
-To compare against the NPN dataset we used four LTS datasets from North America (Table 1).
-Three major ecosystem types and 24 species are represented in these datasets.
+To compare against the NPN dataset we used four LTS datasets from North America (Table 1), representing three major ecosystem types and 24 species.
 Observation metrics varied among these four datasets due to different protocols, so were first converted to binary "yes" and "no" observations for each phenophase (see supplement for details).
-As with the NPN dataset we infer the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation.
+As with the NPN dataset, we infer the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation.
 
@@
 
-For each species/phenophase in each of the five datasets we fit a range of phenology models from a naive mean day of year estimate to a more complex two phase forcing model (Table 2).
-With the exception of two models (naive and linear) the general model form predicts that a phenological event will happen when sufficient forcing units, \(F^{*}\), accumulate from a particular start day of the year.
-The day forcing accumulation begins, $t_{1}$, can either be estimated or fixed.
-For the growing degree day model (GDD) forcing units are the degrees in Celsius above an estimated threshold.
-The Fixed GDD model uses the same form but has fixed values for start day (Jan 1) and temperature threshold (0C).
-The Alternating model has a variable number of required forcing units defined as function of the total number of days below 0C (NCD).
-The Uniforc model is similar to the GDD model but has forcing units derived a sigmoid function.
- \citep{chuine2000}
+Plant species all have unique phenological requirements, thus it is common to fit several types of models to find the one which best represents a specific species \citep{chuine2013}.
+For each species/phenophase in each of the five datasets, we fit a range of phenology models from a naive mean day of year estimate to a more complex two phase forcing model (Table 2).
 The naive model uses only the mean DOY from all prior observations as the estimated DOY.
 The linear model uses a regression with the mean spring (Jan.
-- March) temperature as the only independent variable and DOY as the response variable.
-Finally two models attempt to explain the spatial variation in phenological requirements.
+1 - March 31) temperature as the only independent variable and DOY as the response variable.
+
+For the remaining models the general form form predicts that a phenological event will happen when sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$).
+The start day can either be estimated or fixed.
+For the growing degree day model (GDD) forcing units are the total degrees above the threshold $T$.
+The Fixed GDD model uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ( $T$ = 0C).
+The Alternating model has a variable number of required forcing units defined as a function of the total number of days below 0C since Jan.
+1 ($NCD$).
+The Uniforc model is similar to the GDD model but has forcing units transformed via a sigmoid function \citep{chuine2000}.
+
+Finally, two models attempt to explain the spatial variation in phenological requirements.
 The first spatial model, M1, is an extension of the GDD model which adds a correction in the required forcing using the photoperiod ($L$) \citep{blumel2012}.
@@
 All models were parameterized using differential evolution to minimize the root mean square error (RMSE) of the estimated day of year of the phenological event.
-Differential evolution is a global optimization algorithm similar to simulated annealing, which has traditionally been used to fit complex phenology models \citep{storn1997, chuine2000}.
-Initial testing showed that differential evolution produced more consistent results than simulated annealing and other similar algorithms.
+Differential evolution is a global optimization algorithm which uses a population of randomly initialized models which converge set of parameters which minimize the RMSE \citep{storn1997}.
 Confidence intervals for parameters were obtained by bootstrapping, where individual models were re-fit 250 times using a random sample, with replacement, of the data.
@@
 
-As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set parameterized using only LTS data (with the exception of the M1 and MSB models, see above).
-The distribution for each parameter in each species/phenophase model was compared between the two sets using a Mann-Whitney test.
-This was chosen due to the majority of parameters having non-normal distributions derived from the bootstrapped estimates.
+As described above, two sets of models were produced for each species and phenophase: one set of models parameterized using only NPN data, and one set parameterized using only LTS data (with the exception of the M1 and MSB models, see above).
+The distribution for each parameter in each species/phenophase model was compared between the two sets (LTS and NPN) using a Mann-Whitney test.
+This was chosen due to the majority of parameters having non-normal distributions.
+Using the mean value of each bootstrapped parameter we also calculated the coefficient of determination ($R^2$) between LTS and NPN derived models among the 39 species-phenophases.
 
-Even models with entirely different structures can produce similar estimates of phenological events \citep{basler2016}.
-Thus we also compared estimates of the phenological events between LTS and NPN derived models using the root mean square difference (RMSD) between the two estimates:
+Even models with different parameter values, even entirely different model structure, can still produce similar estimates for the date of phenological events \citep{basler2016}.
+Thus, we also compared the estimated DOY between LTS and NPN derived models.
+For estimates made at the four LTS sites, across all NPN sites, and for each of the 6 models we calculated the coefficient of determination ($R^2$) between LTS and NPN derived estimates.
 
-$$ RMSD = \sqrt[]{ \frac{\sum_{i=1}^{n}(\widehat{DOY_{NPN,i}} - \widehat{DOY_{LTS,i}})^{2}}{n}} $$
-
-where $n$ is the number of estimates produced for a single species and phenophase, and $LTS_{i}$ and $NPN_{i}$ are DOY estimates produced from the respective models.
-A RMSD value close to 0 implies that the LTS and NPN derived models produced very similar estimates.
-For each species and phenophase 12 RMSD values were calculated from the 6 phenology models and 2 scales.
-We calculated the RMSD for estimates at the first scale of the extent of the NPN observations, thus representing a large spatial extent.
-The second scale used estimates from the localities of the LTS datasets, thus representing a single site.
-The RMSD does not consider actual observations as it is meant to infer the correspondence of the NPN and LTS datasets beyond parameter comparison.
-
-For all models the observation error of the held out data was calculated using the using the RMSE:
-
-$$ RMSE = \sqrt[]{ \frac{\sum_{i=1}^{n}(\widehat{DOY_{i}} - DOY_{i})^{2}}{n}} $$
-
-For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, the eight phenology models, and the two scales described above.
-
-Finally we explored which set of models, either LTS or NPN derived, performed better overall at each scale.
-We calculated the pairwise difference of RMSE between LTS and NPN derived models for each model type, species, and phenophase at the two different scales.
-We plotted the density function
-
-$$ f(x) = RMSE_{NPN} - RMSE_{LTS} $$
-
-for each species and phenophases  $$\{ f(x) | s_{1},...,s_{39}\} $$
-
-for each model type and scale.
+To evaluate model performance we outlined four scenarios: A) LTS derived models used to predict LTS observations, B) NPN derived models used to predict LTS observations, C) LTS derived models used to predict NPN observations, and D) NPN derived models used to predict NPN observations.
+Within each of these scenarios we calculated the RMSE of the held out observations for each species, phenophase, and  model type.
+Due to varying sample sizes the RMSE values cannot be easily compared directly, thus we calculated two simple difference metrics for the relative performance of LTS and NPN derived models.
+Using RMSE values from the four scenarios metric 1, $RMSE_{A} - RMSE_{B}$ , compared models at the four LTS sites and metric 2, $RMSE_{C} - RMSE_{D}$, compares models at all NPN sites, representing a large spatial scale.
+In both metrics negative values indicate LTS derived models perform better while positive values indicate the NPN derived model performed better.
+Absolute RMSE values are provided in the supplement (Fig.
+S1-S3).
 
@@
 Each point represents a parameter value for a specific species/phenophase, and is the mean value from 250 bootstrapped  models.
-The black line is a 1:1 line.}
+The black line is a 1:1 line, and $R^2$ is the coefficient of determination.}
 \end{figure}
@@
 
-All parameters had statistically different distributions between LTS and NPN derived estimates when tested with a Mann-Whitney U test.
-Visually comparing the mean values of parameters among all species and phenophases showed a correspondence among the simple models with 1-2 parameters (Fig.
+The Fixed GDD model, followed by the Linear model had the highest $R^2$ between parameters of NPN and LTS derived models, while parameters for all other models had $R^2$ values < 0 (Fig.
 1).
-As model complexity increased this correspondence between the two datasets decreased.
-The Naive model showed a distinct late bias in the mean DOY for phenologcial events, likely resulting from the LTS datasets being mostly in the northern United States.
-Though it still has a small bias the Fixed GDD model had the best correspondence between the two datasets (see suppliment for zoomed in figure).
+The variation around each parameter was high and uneven between LTS and NPN derived models, as all parameters had statistically different (p < 0.001) distributions between NPN and LTS derived models according to the Mann-Whitney U test.
+The Naive model showed a distinct late bias in the mean DOY for phenological events, likely resulting from the LTS datasets being mostly in the northern United States.
+Though it still has a small bias, the Fixed GDD model had the best correlation ($R^2=0.49$ for the single parameter) between the two datasets.
 The outlier for this model, \textit{Larrea tridentata}, has phenology largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
@@
 		\includegraphics[width=1\textwidth]{fig_2_estimate_compare.png}
-	\caption{Difference between NPN and LTS derived estimates.
-Each point is, for a single species/phenophase, the root mean square difference between LTS and NPN derived estimates for a specified model.
-A shows the RMSD over the extent of the NPN dataset (a large scale), B shows the RMSD at the single sites represented by the LTS datasets (a small scale).}
+	\caption{Comparison of estimates between NPN and LTS derived models, made only at LTS sites (top panel) or among all NPN sites (bottom panel).
+Each point is a single observation.
+Different colors indicate observations for a single species/phenophase.}
 \end{figure}
 
-When comparing estimates of phenological events between the two sets of models, many NPN models produced similar estimates to LTS models (Fig.
+When comparing estimates of phenological events between the two sets of models, many NPN and LTS models produced similar estimates (Fig.
 2).
-When estimating at local LTS sites, NPN models produced estimates within 10 days, on average, of LTS models except when using the Naive model (Fig.
-2b).
-When estimating across the spatial extent of NPN data, except when using the Linear and Naive model, most NPN and LTS models produced estimates within 20 days of each other (Fig.
-2a).
+The Fixed GDD model had the highest correlation between the two models sets at NPN sites ($R^2 = 0.82$), while the GDD, M1, and Uniforc models had the highest correlation at LTS sites ($R^2 = 0.54$).
+The MSB (an extension of the alternating model which adds a spatial correction based on mean spring temperature, see Table 2 and methods) improved the correlation between the two datasets over the Alternating model.
+The MSB model improved the $R^2$ from 0.38 to 0.45 at LTS sites, and from -0.2 to -0.13 at NPN sites.
+The M1 model (an extension of the GDD model which adds a spatial correction based on daylength) did not improve the correlation between datasets over the GDD model.
 
-Figure 3 shows the errors for all different model types using the held out data.
-Models predicting NPN observations (Fig.
-3a) have higher error rates than predictions for LTS observations (Fig.
-3b).
-For the former models derived from NPN data made the best predictions, while LTS derived models performed best for the latter.
-Figure 4 shows the pairwise comparison of LTS and NPN errors.
-Similar to Figure 3, the NPN models performed best on NPN observations and LTS models performed best on LTS observations.
+NPN derived models perform better predicting held out NPN observations, while LTS derived models perform better predicting held out LTS observations (Fig.
+3).
+The Naive and Linear models had the largest differences between the two model sets, while the Fixed GDD model had relatively similar errors when evaluated on both NPN and LTS held out observations.
+Though the Fixed GDD model had the highest agreement in errors between NPN and LTS derived models, it was not the best performing model overall.
+The GDD and Uniforc were commonly the most best performing, having the lowest RMSE in 22\% and 40\% of cases among NPN derived models, and 42\% and 32\% of cases among LTS derived models, respectively (Fig.
+S1 & S2).
 
@@
 	\centering
-		\includegraphics[width=1\textwidth]{fig_3_error_compare.png}
-	\caption{Root mean square error of model predictions compared to observations.
-The top plot is for all held out observations in the NPN dataset.
-The bottom plot is for all held out observations in the LTS datasets.}
-\end{figure}
-
-Figure 4 shows the pairwise comparison of specific species/phenophase combinations for held out data.
-Overall NPN derived models performed best on held out NPN observations, while LTS derived models performed best on held out LTS observations.
-The naive and linear models had the largest differences between the two models sets, while the Fixed GDD model had relatively similar errors when evaluated on both NPN and LTS held out observations.
-
-\begin{figure}
-	\centering
-		\includegraphics[width=1\textwidth]{fig_4_pairwise_error.png}
-	\caption{Density plots for pairwise RMSE comparison. The x axis shows the distribution for $RMSE_{NPN} - RMSE_{LTS}$  for the  39 species/phenophase combinations.}
+		\includegraphics[width=1\textwidth]{rmse_metrics_density_plot.png}
+	\caption{Density plots for the two different metrics.
+Each plot represents the metric value for the 39 species/phenophase comparisons.
+The metric represents the difference in RMSE values between the difference scenarios (A - B, or C - D).}
 \end{figure}
@@
 
-Despite having statistically different parameters to LTS derived models, many NPN derived models were able to produce similar estimates for phenological events.
-This is likely due to models having a non-linear parameter space and being non-identifiable.
-For example two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 can produce nearly identical results in many, but not all, scenarios.
-As model complexity increases the ability to produce similar estimates increases.
-Indeed, the Uniforic models with NPN derived parameters still had estimates within 10 days of LTS derived parameters for most species (Fig.
-2) despite having very different parameters (Fig.
-1).
+\subsection{Discussion intro}
 
-This makes the interpretation of these parameters questionable, which is troublesome as they are designed to be biologically relevant.
+Here we compared two datasets which are typical in ecology.
+One is intensively studied with a long time series from several locations, the other has a large spatial coverage of hundreds of locations, but with no single location having a long timeseries.
+We show that they vary in parameter estimates and observation estimates in all but the simplest models, and regardless of the model they each predict their own out of sample data better than the other.
+Uses of each type of dataset thus depends on the desired research goals.
+Analyzing trends in response to changing climate will likely require using a long term study dataset, but extrapolation of those results to other locations must be done with care.
+Studying floral resource availability across a spatial gradient would be best done with NPN as it has a large number of species available, but attributing
+
+Here we emphasize the goal of large scale forecasting, where models must be built to the best estimate of the desired event, they must have independent/predictor variables which are currently available or forecastable (ie.
+temperature), and they must be able to extrapolate to new locations with no prior observations.
+While models built using the LTS datasets are able to give the best estimates for their locations, they do not perform better than NPN models on large scale data.
+This decrease in model performance with distance in single site phenology models is well known \citep{basler2016, xu2013, garcia-mozo2008}.
+
+The NPN dataset will be ideal in making large scale forecasts, but the ideal situation will be to integrate all types of data into analysis and forecasts.
+Datasets which have a long time series contain valuable information on the inter annual variation, the PHENOCAM network has a large spatial extent as well as fine temporal grain \citep{richardson2018}, and phenology derived from satellite imagery has worldwide coverage.
+Here straightforward models require all data to be from the same source, and give equal weight to all observations.
+More complex data integration methods can use data from all different sources to give the best estimate for forecasts \citep{ogle2015} and continuously update forecasts as new real time data becomes available \citep{luo2011}.
+
+\subsection{Details of the differences seen in the analysis}
+
+Despite having statistically different parameters to LTS derived models, many NPN derived models were able to produce similar estimates for phenological events when using a single parameter Fixed GDD model.
+This is likely due to models being non-identifiable.
+For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 can produce nearly identical results in many, but not all, scenarios.
+This makes the interpretation of these parameters difficult, which is troublesome as they are designed to be biologically relevant.
+For example the total forcing required, $F^*$, reflects the amount of heating in the spring required for a phenological even to occur.
+Here we show that there is little agreement when these parameters are estimated for the same species using two datasets.
+
 \cite{chuine2016} found similar issues, where models estimating endodormancy break using only observed budbreak did not match direct measurements of endodormancy break, suggesting that phenology models which are fit using only the observed phenophase cannot accurately infer the internal mechanisms described in the models.
 \cite{basler2016} suggests that the thermal forcing in models is responsible for most of the accuracy, while any additional parameters are only fit to residuals and sensitive to over fitting.
-Here our simplest model, the Fixed GDD model which uses only a warming component, had reasonable (thought statistically different) correspondence between LTS and NPN datasets.
-This correspondence decreases as more parameters are added to models, suggesting that if the NPN dataset is being used to make inferences about plant physology simple models could give a more consistent results.
+Here, our simplest model, the Fixed GDD model which uses only a warming component, had the highest correlation of it's single parameter between LTS and NPN datasets.
+This correlation decreases as more parameters are added to models, suggesting that if the NPN dataset is being used to make inferences about plant physiology, simple models could give more consistent results.
 Parameter estimates of more complex models do not agree with models from the same species in LTS datasets, and it is possible even the LTS datasets are over fitting these models.
 
-Even with similar parameters, the Fixed GDD model was not the best performing model when tested on out of sample data.
-While the values of additional parameters may be incorrect in the above sense, they nonetheless provide more predictive power.
-Across the two scales the GDD, Uniforc, and Alternating model were generally the best performing (Fig.
-3).
-Considering only NPN derived models predicting NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1).
 One goal of ecological forecasting is to inform decision makers about the future state of systems \citep{dietz2018}.
-Thus the best performing model should be used in these cases to give the best possible forecast.
-The GDD and Uniforc models show promise for building models with the NPN dataset and making plant phenology forecasts across large scales.
+Thus, the best performing model should be used in these cases to give the best possible forecast.
+In terms of lowest RMSE the GDD and Uniforc The GDD and Uniforc models show promise for building models with the NPN dataset and making plant phenology forecasts across large scales as they were the best performing models for the majority of species and phenophases (fig.
+S1).
+Even with similar parameters, the Fixed GDD model was not the best performing model when tested on held out data.
+While the values of additional parameters may be incorrect in the sense of the prior paragraph, they nonetheless provide more predictive power.
 
 Other models show promise for forecasting under different scenarios.
-For example if using data from a large spatial extent to make predictions at single sites our results show the Fixed GDD model is the best option, as performance under this scenario is very similar to performance of Fixed GDD models built from LTS data (Fig.
+For example, if using data from a large spatial extent to make predictions at single sites (sensu.
+Scenario B) our results show that the Alternating model is the best option, as performance under this scenario is very similar to performance of Alternating models built from LTS data (Fig.
+3).
+In the opposite scenario of making large-scale predictions using only data from a single site (sensu.
+Scenario C) the Fixed GDD model is likely the best option, as most LTS derived Fixed GDD models have a RMSE only 5 days greater, on average, than the corresponding NPN model (Fig.
 4).
-In the opposite scenario of making large scale predictions using only data from a single site the Alternating model is likely the best option, as most LTS derived models have a RMSE only 5 days greater on average than the corresponding NPN model when tested on out of sample NPN data (Fig.
-4).
-The Alternating model is also one of the best performing model overall for large scale NPN data (Fig.
+The Alternating model is also one of the best performing model overall for large-scale NPN data (Fig.
 3a).
@@
 Their result was strengthened by a 40 year timeseries across a large region.
-\cite{chuine2017} noted that incorporating the spatial vaiation in warming requirements as a key issue in future phenology research.
-We feel the NPN dataset, as well as other large scale phenology datasets, will be key in addressing this and other phenological research needs.
+\cite{chuine2017} noted that incorporating the spatial variation in warming requirements as a key issue in future phenology research.
+We feel the NPN dataset, as well as other large-scale phenology datasets, will be key in addressing this and other phenological research needs.
 
@@
 
-Currently the best phenology model for a single location will always be those fit using data from only that location, yet these single site models do not transfer well to large scales.
-Making large scale phenology forecasts will require incorporating large scale observations such as those from the NPN dataset.
+Currently the best phenology model for a single location will be one fit using data from only that location, yet these single site models do not transfer well to large scales.
+Making large-scale phenology forecasts will require the incorporation of large-scale observations such as those from the NPN dataset.
 Here we have shown that the NPN dataset is capable of parameterizing phenology models which make comparable predictions to LTS datasets.
@@
 The National Phenology Network dataset will likely be an important resource for studying phenology.
-The addition of numerous species and observations from across a range of conditions is important for informing large scale processes, as similar datasets have shown in Europe \citep{olsson2014, basler2016} and Asia \citep{xu2013, zhang2017}.
-Research from long term small scale studies has been invaluable and should no doubt be continued, but can now be complimented by using large scale NPN data (ie.
+The addition of numerous species and observations from across a range of conditions is important for informing large-scale processes, as similar datasets have shown in Europe \citep{olsson2014, basler2016} and Asia \citep{xu2013, zhang2017}.
+Research from long-term small scale studies has been invaluable and should no doubt be continued, but can now be complimented by using large-scale NPN data (ie.
 \citep{jeong2013, melaas2016}).
@@
 \setcounter{figure}{0}    
-\section*{Suppliment}
+\section*{Supplement}
 \begin{figure}
 	\centering
-		\includegraphics[width=1\textwidth]{fig_s1_best_npn_models.png}
-	\caption{Model error for specific species and phenophases of the NPN dataset. X marks the best performing models for the respective data type.}
+		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
+	\caption{RMSE for specific species and phenophases of the NPN dataset. X marks the best performing models for the respective data type.}
+\end{figure}
+
+\begin{figure}
+	\centering
+		\includegraphics[width=1\textwidth]{supplement_best_lts_models.png}
+	\caption{RMSE for specific species and phenophases of the LTS datasets. X marks the best performing models for the respective data type.}
+\end{figure}
+
+\begin{figure}
+	\centering
+		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
+	\caption{RMSE of all species and phenophases of the four scenarios described in the text. These values are calculated using held out test data.}
 \end{figure}

### 2018-05-09 manuscript/main.tex
first major revision from ethan

@@
 \author[1]{Shawn D. Taylor}
-\author[2]{Joan M. Meiners}
-\author[3]{Kristina Riemer}
-\author[4]{Michael C. Orr}
-\author[5]{Ethan P. White}
+\author[1]{Joan M. Meiners}
+\author[2]{Kristina Riemer}
+\author[3]{Michael C. Orr}
+\author[2,4]{Ethan P. White}
 
-\affil[1,2]{UFL SNRE}
-\affil[4]{Key Laboratory of Zoological Systematics and Evolution, Institute of Zoology, Chinese Academy of Sciences, Beijing 100101, P.R. China}
-\affil[3,5]{UFL WEC}
-\affil[5]{UFL Bioinformatics}
+\affil[1]{School of Natural Resources and Environment, University of Florida Gainesville, FL, United States}
+\affil[2]{Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, FL, United States}
+\affil[3]{Key Laboratory of Zoological Systematics and Evolution, Institute of Zoology, Chinese Academy of Sciences, Beijing 100101, P.R. China}
+\affil[4]{Informatics Institute, University of Florida, Gainesville, FL, United States}
 \corrauthor[1]{Shawn D. Taylor}{shawntaylor@weecology.org}
@@
 
-Robust plant phenology models are necessary for forecasting the future state of the ecological systems at several scales, from global climate to impacts on plant and animal communities.
-Plant phenology performs an important role in many fields of ecological research such as ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}.
-At the largest scale, variation in the timing of spring leaf out and fall senescence account for a significant amount of uncertainty in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long-term forecasts \citep{richardson2012}.
-At local scales, because of species specific responses to temperature and precipitation, changes in flowering times can alter the flower community throughout the growing season \citep{caradonna2014, diez2012, theobald2017}.
-This change in resource availability can affect the abundance and richness of pollinators \citep{ogilvie2017a, ogilvie2017b}, and higher trophic levels \citep{tylianakis2008}.
-Accurately accounting for phenology in these and other scenarios will be critical for forecasting environmental change.
+Plant phenology, the timing of biological events such as flowering, plays an important role in many fields of ecological research extending from local to global scales \citep{cleland2007, richardson2013, tang2016}.
+At large scales, variation in the timing of spring leaf out and fall senescence influence the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long-term climate forecasts \citep{richardson2012}.
+At local scales, species-specific responses to temperature and precipitation can alter flower communities \citep{diez2012, caradonna2014, theobald2017} and affect the abundance and richness of both pollinators \citep{ogilvie2017a, ogilvie2017b} and higher trophic levels \citep{tylianakis2008}.
+Robust cross-scale plant phenology models are needed for understanding the processes governing these key timings of biological events and forecasting the impacts of changes in these timings in response to environmental change.
 
-To best forecast plant phenology researchers will need models which account for the full range of variation across a species range \citep{richardson2013, chuine2017}.
-Many plant phenological studies in the past have relied upon datasets collected at a single location over a long time-frame by a single group \citep{cook2012, roberts2015, iler2013, wolkovich2012}.
-These datasets have high temporal coverage, with some approaching several decades, and represent the species and climate variability at their respective sites extremely well.
-Yet it is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}.
-This lack of transferability across space can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995, diez2012}.
-Future models which account for this variability will require data from as much of a species' range as possible.
-Further, data from a single location may not be adequate for ecosystem level phenology modelling due to a potentially limited species pool.
-Thus, new sources of data beyond traditional single site studies will be needed.
+Many plant phenology studies use intensively collected datasets collected at a single location over a long time-period by a single research group \citep{cook2012, wolkovich2012, iler2013, roberts2015}.
+These datasets have regular sampling and large numbers of samples over extensive periods of time.
+As a result they represent the biological and climatic variability at that site extremely well.
+However, it is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}.
+This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, and differences in plant age and population density \citep{kramer1995, diez2012}.
+In addition, data from a single location may not be adequate for larger scale phenology modelling due to limited species pools.
+Therefore it has been suggested that accurately understanding and forecasting phenology at larger scales will require models that account for the full range of variation across a species range \citep{richardson2013, chuine2017}.This will necessitate the use of data sources beyond traditional single site studies.
 
-Citizen science data is becoming increasingly important for ecological research, with the potential for significant impact in many fields \citep{dickinson2010, tulloch2013, kelling2009}.
-Citizen science efforts have helped document range shifts \citep{hitch2007}, study the effect of landscape fragmentation on amphibians \citep{cosentino2014}, and informed ecological theory of community composition \citep{locey2013}.
-A relatively new citizen science group, The National Phenology Network (NPN) is an organization which accepts phenology observations from volunteers throughout North America and makes the data freely available \citep{schwartz2012a}.
-The NPN dataset has already been used to study, among many things, variation in oak phenology \citep{gerst2017}, large-scale community phenology models \citep{melaas2016}, and forecasting long-term phenology trends \citep{jeong2013}.
-The regional to continental scale of citizen science data helps researchers explain large-scale patterns across natural gradients, but can also introduce new sources of error from observer skill, variation in sampling effort, and spatial bias \citep{dickinson2010}.
-These issues can be accounted for in models but first must be identified and quantified.
-With hundreds of observers across North America, the observer and sampling bias for species identification and the true date of phenological events is potentially high.
-Though volunteers can be very accurate at distinguishing different phenophases \citep{fuccillo2015}, observations can still be made sporadically throughout the growing season, some years skipped, or visits to a particular site discontinued entirely.
+Data from citizen science projects is becoming increasingly important for ecological research \citep{kelling2009, dickinson2010, tulloch2013}.
+Because this data is often collected by large numbers of volunteers it is possible to gather data at much larger scales than with individual research teams.
+A relatively new citizen science project, The National Phenology Network (NPN) collects phenology observations from volunteers throughout North America and makes the data openly available \citep{schwartz2012a}.
+Data from this project has already been used to study continental-scale variation in oak phenology \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}.
+Large scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the NPN dataset has the potential to meet these needs for North American plant species and communities.
+However, the features that allow citizen science projects to collect data at large-scales can also introduce sources of error including spatial biases toward cities and easily accessible areas and variation in sampling effort and observer skill \citep{dickinson2010}.
+With hundreds of participants across North America, the potential for variation among observers for species identification and dating phenological events is high.
+While volunteers have been shown to be accurate at distinguishing different phenophases \citep{fuccillo2015}, observations are sometimes made sporadically within seasons, across years, and among locations.
+This means that the quality of the data at a specific site will typically be lower than for intensive long-term studies.
 
-While there have been several studies looking at best practices when using the NPN dataset \citep{crimmins2017, gerst2016}, no study to date has compared it directly with data from long-term studies to test its adequacy for plant phenology models.
-Data from long-term studies (LTS) have been invaluable for phenology research for several decades, but datasets with a larger spatial extent and more complete species pools will be needed to model the complex phenological dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}.
-Such datasets from China and Europe have already contributed considerably to phenological research \citep{olsson2014, basler2016, xu2013, zhang2017}, and the NPN dataset has the potential to meet these needs for North American plant species and communities.
-Here, we fit seven different plant phenology models to the budburst and first flowering phenophases of XX plant species.
-The models were fit twice, once using data from the NPN dataset, and again for the same species using data from an LTS dataset.
-We then compared three outputs between the two sets of models built using NPN and LTS data: 1) the models' parameter estimates; 2) the models' estimates of phenophase timing; and 3) the models' errors using held out testing data.
+In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large scale citizen science projects influence both our inferences about biological processes and our ability to predict future phenology events.
+Here, we fit a suite of plant phenology models for the budburst and first flowering phenophases of 24 plant species to data from both the NPN and a set of intensive long-term studies.
+We compare the resulting models based on both inference about models and parameters and predictions for unobserved events.
+We use this comparison to assess the best methods for both local and large scale phenology modeling and to point the way forward for integrating large scale and local scale data to obtain the best possible models across scales.
 
@@
 
-The National Phenology Network observations are based on status-based monitoring, where observers answer yes, no, or unsure when asked if an individual plant has a specific phenophase present \citep{denny2014}.
+The National Phenology Network protocol uses status-based monitoring, where observers answer yes, no, or unsure when asked if an individual plant has a specific phenophase present \citep{denny2014}.
 Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves.
@@
 
-To compare against the NPN dataset we used four LTS datasets from North America (Table 1), representing three major ecosystem types and 24 species.
-Observation metrics varied among these four datasets due to different protocols, so were first converted to binary "yes" and "no" observations for each phenophase (see supplement for details).
-As with the NPN dataset, we infer the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation.
+To represent long-term intensive phenology studies we used four datasets from North America (Table 1), representing three major ecosystem types and 24 species.
+All four long-term studies are located in the U.S.A and are part of the Long Term Ecological Research network.
+The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S.A.
+and are dominated by deciduous broadleaf species.
+The H.J.
+Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S.A.
+The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.A.
+Observation metrics varied among these four datasets due to different protocols, so we converted all metrics to binary "yes" and "no" observations for each phenophase (see supplement for details).
+As with the NPN data, we inferred the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation and only kept species x phenophases which had at least 30 total observations.
 
@@
 
-Plant species all have unique phenological requirements, thus it is common to fit several types of models to find the one which best represents a specific species \citep{chuine2013}.
-For each species/phenophase in each of the five datasets, we fit a range of phenology models from a naive mean day of year estimate to a more complex two phase forcing model (Table 2).
-The naive model uses only the mean DOY from all prior observations as the estimated DOY.
+Plant species all have unique phenological requirements, thus it is common to fit multiple models to find the one that best represents a specific species and phenophase \citep{chuine2013}.
+For each species and phenophase in each of the five datasets (NPN and four LTS datasets), we fit a range of phenology models ranging from a naive mean day of year estimate to more complex two phase forcing models (Table 2).
+The naive model uses the mean DOY from prior observations as the estimated DOY.
 The linear model uses a regression with the mean spring (Jan.
-1 - March 31) temperature as the only independent variable and DOY as the response variable.
-
-For the remaining models the general form form predicts that a phenological event will happen when sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$).
+1 - March 31) temperature as the independent variable and DOY as the response variable.
+For the remaining models the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$).
 The start day can either be estimated or fixed.
 For the growing degree day model (GDD) forcing units are the total degrees above the threshold $T$.
-The Fixed GDD model uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ( $T$ = 0C).
+The Fixed GDD model uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0C).
 The Alternating model has a variable number of required forcing units defined as a function of the total number of days below 0C since Jan.
 1 ($NCD$).
-The Uniforc model is similar to the GDD model but has forcing units transformed via a sigmoid function \citep{chuine2000}.
+The Uniforc model is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}.
 
-Finally, two models attempt to explain the spatial variation in phenological requirements.
+We also fit two models that attempt to capture spatial variation in phenological requirements.
 The first spatial model, M1, is an extension of the GDD model which adds a correction in the required forcing using the photoperiod ($L$) \citep{blumel2012}.
 The second, the Macroscale Species-specific Budburst model (MSB), uses the mean spring temperature as a linear correction on the total forcing required in the Alternating model \citep{jeong2013}.
-Since there is little to no spatial variation in the LTS datasets, the two spatial models were fit using only the NPN dataset.
-The resulting parameters, estimates, and errors for the NPN derived M1 and MSB models were compared to the LTS GDD and Alternating models, respectively, for each species and phenophase.
+Since there is little to no spatial variation in the LTS datasets, the two spatial models were only fit to data from the NPN.
+The resulting parameters, estimates, and errors for the NPN derived M1 and MSB models were compared to their non-spatial analogs (the GDD and Alternating models, respectively) for each species and phenophase in the LTS data.
 
@@
 All models were parameterized using differential evolution to minimize the root mean square error (RMSE) of the estimated day of year of the phenological event.
-Differential evolution is a global optimization algorithm which uses a population of randomly initialized models which converge set of parameters which minimize the RMSE \citep{storn1997}.
+Differential evolution is a global optimization algorithm which uses a population of randomly initialized models to find the set of parameters that minimize the RMSE \citep{storn1997}.
 Confidence intervals for parameters were obtained by bootstrapping, where individual models were re-fit 250 times using a random sample, with replacement, of the data.
-A random 20\% of all data was held out from model fitting for use in later evaluation.
+A random 20\% of all data was held out from model fitting for model evaluation.
 Predictions were made by taking the mean DOY from the 250 bootstrapped models.
@@
 As described above, two sets of models were produced for each species and phenophase: one set of models parameterized using only NPN data, and one set parameterized using only LTS data (with the exception of the M1 and MSB models, see above).
-The distribution for each parameter in each species/phenophase model was compared between the two sets (LTS and NPN) using a Mann-Whitney test.
-This was chosen due to the majority of parameters having non-normal distributions.
+To compare the inferences about process made by the two datasets we compared the distribution for each parameter in each species x phenophase model between LTS and NPN using a Mann-Whitney test (since the majority of parameters having non-normal distributions).
 Using the mean value of each bootstrapped parameter we also calculated the coefficient of determination ($R^2$) between LTS and NPN derived models among the 39 species-phenophases.
+In the three cases where the same species x phenophase combination occurred in two LTS sites, the two instances were compared separately to the NPN data.
 
-Even models with different parameter values, even entirely different model structure, can still produce similar estimates for the date of phenological events \citep{basler2016}.
-Thus, we also compared the estimated DOY between LTS and NPN derived models.
-For estimates made at the four LTS sites, across all NPN sites, and for each of the 6 models we calculated the coefficient of determination ($R^2$) between LTS and NPN derived estimates.
+Models with different parameter values, and even entirely different model structures, can produce similar estimates for the date of phenological events \citep{basler2016}.
+Therefore, to compare the predictions and potential forecasts for models fit to the different datasets we compared the estimated DOY predicted by the LTS and NPN derived models.
+We calculated the coefficient of determination ($R^2$) between LTS and NPN derived estimates for estimates made at the four LTS sites, and across all NPN sites, for each of the 6 models.
 
-To evaluate model performance we outlined four scenarios: A) LTS derived models used to predict LTS observations, B) NPN derived models used to predict LTS observations, C) LTS derived models used to predict NPN observations, and D) NPN derived models used to predict NPN observations.
+We also directly evaluated model performance using four combinations of models and observed data: A) LTS derived models predicting LTS observations, B) NPN derived models predicting LTS observations, C) LTS derived models predicting NPN observations, and D) NPN derived models predicting NPN observations.
 Within each of these scenarios we calculated the RMSE of the held out observations for each species, phenophase, and model type.
-Due to varying sample sizes the RMSE values cannot be easily compared directly, thus we calculated two simple difference metrics for the relative performance of LTS and NPN derived models.
-Using RMSE values from the four scenarios metric 1, $RMSE_{A} - RMSE_{B}$ , compared models at the four LTS sites and metric 2, $RMSE_{C} - RMSE_{D}$, compares models at all NPN sites, representing a large spatial scale.
-In both metrics negative values indicate LTS derived models perform better while positive values indicate the NPN derived model performed better.
+Using the RMSE values we calculated two difference metrics to compare the performance of LTS and NPN derived models on different data types.
+The first metric focuses on local scale prediction by comparing the fits of LTS and NPN derived models on LTS observations: $RMSE_{A} - RMSE_{B}$.
+The second metric focuses on large scale prediction by comparing the fits of LTS and NPN derived models on the NPN data: $RMSE_{C} - RMSE_{D}$.
+Negative values indicate LTS derived models perform better while positive values indicate the NPN derived model performed better.
+In the three cases where the same species x phenophase combination occurred in two LTS sites, the LTS-LTS comparison (scenario A) was made within each site, not across sites, to focus on local scale prediction when LTS data is available.
 Absolute RMSE values are provided in the supplement (Fig.
@@
 		\includegraphics[width=1\textwidth]{fig_1_param_comparison.png}
-	\caption{Comparisons of parameters between NPN and LTS derived models.
-Each point represents a parameter value for a specific species/phenophase, and is the mean value from 250 bootstrapped  models.
-The black line is a 1:1 line, and $R^2$ is the coefficient of determination.}
+	\caption{Comparisons of parameter estimates between NPN and LTS derived models.
+Each point represents a parameter value for a specific species and phenophase, and is the mean value from 250 bootstrapped models.
+The black line is the 1:1 line.
+The $R^2$ is the coefficient of determination, which can be negative if the relationship between the two parameter sets is worse than no relationship but with the same mean values.}
 \end{figure}
@@
 
-The Fixed GDD model, followed by the Linear model had the highest $R^2$ between parameters of NPN and LTS derived models, while parameters for all other models had $R^2$ values < 0 (Fig.
+The best matches between parameter estimates based on NPN and LTS data were the Fixed GDD model ($R^2$ = 0.49) and the Linear model ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
+The parameters for all other models had $R^2$ values < 0 indicating that the relationship was worse than no relationship between the parameters but with matching mean parameter values across the two sets of models (Fig.
 1).
-The variation around each parameter was high and uneven between LTS and NPN derived models, as all parameters had statistically different (p < 0.001) distributions between NPN and LTS derived models according to the Mann-Whitney U test.
-The Naive model showed a distinct late bias in the mean DOY for phenological events, likely resulting from the LTS datasets being mostly in the northern United States.
-Though it still has a small bias, the Fixed GDD model had the best correlation ($R^2=0.49$ for the single parameter) between the two datasets.
-The outlier for this model, \textit{Larrea tridentata}, has phenology largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
+The Naive model showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTS datasets being mostly in the northern United States.
+The one major outlier for the Fixed GDD model, \textit{Larrea tridentata}, has phenology largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
+While the Fixed GDD and Linear models showed reasonable correspondence between parameter estimates, all parameters for individual species x phenophase combinations were statistically different between NPN and LTS derived models according to the Mann-Whitney U test (p < 0.001).
 
@@
 		\includegraphics[width=1\textwidth]{fig_2_estimate_compare.png}
-	\caption{Comparison of estimates between NPN and LTS derived models, made only at LTS sites (top panel) or among all NPN sites (bottom panel).
-Each point is a single observation.
-Different colors indicate observations for a single species/phenophase.}
+	\caption{Comparison of predicted day of year (DOY) of all phenological events between NPN and LTS derived models.
+Top panels show comparisons at LTS sites and bottom panels show comparisons at NPN sites.
+Each point is an estimate for a single held out observation.
+Colors indicate observations for a single species x phenophase combination.}
 \end{figure}
@@
 The Fixed GDD model had the highest correlation between the two models sets at NPN sites ($R^2 = 0.82$), while the GDD, M1, and Uniforc models had the highest correlation at LTS sites ($R^2 = 0.54$).
-The MSB (an extension of the alternating model which adds a spatial correction based on mean spring temperature, see Table 2 and methods) improved the correlation between the two datasets over the Alternating model.
+Comparing models with spatial corrections to the non-spatial alternatives, the MSB (an extension of the alternating model with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the Alternating model.
 The MSB model improved the $R^2$ from 0.38 to 0.45 at LTS sites, and from -0.2 to -0.13 at NPN sites.
-The M1 model (an extension of the GDD model which adds a spatial correction based on daylength) did not improve the correlation between datasets over the GDD model.
+The M1 model (an extension of the GDD model with a spatial correction based on daylength) did not improve the correlation between datasets over the GDD model.
 
-NPN derived models perform better predicting held out NPN observations, while LTS derived models perform better predicting held out LTS observations (Fig.
+When comparing the prediction accuracy on held out data NPN derived models make more accurate predictions for held out NPN observations, and LTS derived models perform better on held out LTS observations (Fig.
 3).
 The Naive and Linear models had the largest differences between the two model sets, while the Fixed GDD model had relatively similar errors when evaluated on both NPN and LTS held out observations.
-Though the Fixed GDD model had the highest agreement in errors between NPN and LTS derived models, it was not the best performing model overall.
-The GDD and Uniforc were commonly the most best performing, having the lowest RMSE in 22\% and 40\% of cases among NPN derived models, and 42\% and 32\% of cases among LTS derived models, respectively (Fig.
-S1 & S2).
+Though the Fixed GDD model had the highest agreement in accuracy between NPN and LTS derived models, it was not the best performing model overall.
+The GDD and Uniforc models commonly made the best predictions, having the lowest RMSE in 22\% and 40\% of cases among NPN derived models, and 42\% and 32\% of cases among LTS derived models, respectively (Fig.
+S1 \& S2).
 
@@
 		\includegraphics[width=1\textwidth]{rmse_metrics_density_plot.png}
-	\caption{Density plots for the two different metrics.
-Each plot represents the metric value for the 39 species/phenophase comparisons.
-The metric represents the difference in RMSE values between the difference scenarios (A - B, or C - D).}
+	\caption{Differences in prediction error between NPN and LTS derived models.
+Density plots for comparisons of predictions on LTS data (top row) and NPN data (bottom row).
+Each plot represents the difference between the RMSE for LTS derived model and the NPN derived model, meaning that values less than zero indicate more accurate prediction by LTS derived models and values greater than zero indicate more accurate prediction by NPN derived models.
+Differences are calculated pairwise for the 39 species/phenophase comparisons.}
 \end{figure}
@@
 
-\subsection{Discussion intro}
+Data for building phenology models typically falls into two categories: intensive long-term data with long time-series at a small number of locations (e.g., the LTS data in this study) and large scale data with less intensive data at hundreds of locations (e.g., the NPN data).
+This data scenario, a small amount of intensive data and a large amount of less intensive data, is common in many areas of science, and results in the need to understand how to choose between, or combine, data sources.
+We explored this issue for phenology modeling at local and continental scales for making predictions and inferring process from models.
+For inference, we found that models based on different data sources resulted in different parameter estimates for all but the simplest models.
+For prediction we found that models fit to different data sources tended to make similar predictions, but that models better predicted out of sample data from the data type they were fit to.
+As a result the best choice of both data and models depends on the desired research goals.
+Understanding and making predictions for the phenology of a single location is best served by intensive local scale data, if it is available, but large scale datasets work better for extrapolating phenology predictions across a species range.
+These results are consistent with other research showing that phenology model performance decreases when transferring single site models to other locations \citep{garcia-mozo2008, xu2013, basler2016} and with the call for models that better incorporate spatial variation in phenology requirements \citep{richardson2013, chuine2017}.
 
-Here we compared two datasets which are typical in ecology.
-One is intensively studied with a long time series from several locations, the other has a large spatial coverage of hundreds of locations, but with no single location having a long timeseries.
-We show that they vary in parameter estimates and observation estimates in all but the simplest models, and regardless of the model they each predict their own out of sample data better than the other.
-Uses of each type of dataset thus depends on the desired research goals.
-Analyzing trends in response to changing climate will likely require using a long term study dataset, but extrapolation of those results to other locations must be done with care.
-Studying floral resource availability across a spatial gradient would be best done with NPN as it has a large number of species available, but attributing
+In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process oriented model: the Fixed GDD (Fig.
+1).
+These differences may be caused by a variety of factors that have different implications for interpreting process oriented models and their parameters.
+First, the differences could result from limitations in the sampling of the NPN dataset, leading to less accurate parameter estimates.
+If this is the case it would suggest using local scale data when it is available for making inferences about plant physiology and focusing on the Fixed GDD model when large scale data is all that is available.
+Second, spatial variation in phenology requirements could drive these differences, because large scale data integrates over that spatial variation, while local scale data only estimates the phenological requirements for a specific site.
+In this case large scale would provide a better estimate of the general phenological requirements of a species, but local scale data would provide a more accurate understanding for a single site.
+The best solution to this issue would be the development of models that accurately incorporate spatial variation.
+Finally, these differences could result from issues with model identifiability, where different parameter values yield nearly identical fits to the data, which can result in fits to different datasets varying widely even if the underlying processes generating the data are the same.
 
-Here we emphasize the goal of large scale forecasting, where models must be built to the best estimate of the desired event, they must have independent/predictor variables which are currently available or forecastable (ie.
-temperature), and they must be able to extrapolate to new locations with no prior observations.
-While models built using the LTS datasets are able to give the best estimates for their locations, they do not perform better than NPN models on large scale data.
-This decrease in model performance with distance in single site phenology models is well known \citep{basler2016, xu2013, garcia-mozo2008}.
+Information about which of these issues may be causing the differences between datasets can be explored using our analyses.
+Despite the substantial differences in parameter estimates the LTS and NPN derived models produced similar estimates for phenological events in most cases (Fig.
+2).
+This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues.
+For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 produce nearly identical results in many scenarios.
+This possibility is supported by the fact that the the highest correlation between parameter estimates is seen in a models with only 1 or 2 parameters.
+In addition, the bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTS data (Supplementary Material).
+Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTS datasets (Supplementary Material).
+These results are consistent with research showing that models estimating endodormancy break from budbreak time-series failed to accurately infer the internal phenology described in the models \citep{chuine2016}.
+\cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over fitting.
+Here, our simplest model, the Fixed GDD model which uses only a warming component, had the highest correlation among parameters between LTS and NPN datasets.
+In combination with this previous research, our results warrant caution in interpreting the parameter estimates from complex phenology models regardless of the data source used for fitting the models.
 
-The NPN dataset will be ideal in making large scale forecasts, but the ideal situation will be to integrate all types of data into analysis and forecasts.
-Datasets which have a long time series contain valuable information on the inter annual variation, the PHENOCAM network has a large spatial extent as well as fine temporal grain \citep{richardson2018}, and phenology derived from satellite imagery has worldwide coverage.
-Here straightforward models require all data to be from the same source, and give equal weight to all observations.
-More complex data integration methods can use data from all different sources to give the best estimate for forecasts \citep{ogle2015} and continuously update forecasts as new real time data becomes available \citep{luo2011}.
+While complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information beyond the Fixed GDD model based on their ability to make out-of-sample predictions.
+Based on the RMSE, the GDD and Uniforc models produce the best out-of-sample predictions for the majority of species and phenophases at both large and local scales (Fig.
+S1).
+For local scale predictions This demonstrates that the more complex models are capturing additional information about phenology and that some of the differences between datasets result from differences in either the scales or the sampling of the data.
+Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, and in combination with our results showing observed differences in parameter estimates between LTS sites (Supplementary Material) this suggests that variation in phenological requirements across the the range is likely important.
+However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses.
+Correspondence between parameter estimates (Fig.
+1), estimates of phenological events (Fig.
+2), and out of sample error rates (Fig.
+3) for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively.
+This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors.
+\cite{basler2016} used the M1 model to predict budburst on six species across Europe and found it was usually among the best models in terms of RMSE, albeit never by more than a single day.
+Their result was strengthened by having a 40 year timeseries across a large region.
+\cite{chuine2017} noted that incorporating the spatial variation in warming requirements as a primary issue in future phenology research.
+Large-scale phenology datasets, like NPN, will be key in addressing this and other phenological research needs.
 
-\subsection{Details of the differences seen in the analysis}
-
-Despite having statistically different parameters to LTS derived models, many NPN derived models were able to produce similar estimates for phenological events when using a single parameter Fixed GDD model.
-This is likely due to models being non-identifiable.
-For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 can produce nearly identical results in many, but not all, scenarios.
-This makes the interpretation of these parameters difficult, which is troublesome as they are designed to be biologically relevant.
-For example the total forcing required, $F^*$, reflects the amount of heating in the spring required for a phenological even to occur.
-Here we show that there is little agreement when these parameters are estimated for the same species using two datasets.
-
-\cite{chuine2016} found similar issues, where models estimating endodormancy break using only observed budbreak did not match direct measurements of endodormancy break, suggesting that phenology models which are fit using only the observed phenophase cannot accurately infer the internal mechanisms described in the models.
-\cite{basler2016} suggests that the thermal forcing in models is responsible for most of the accuracy, while any additional parameters are only fit to residuals and sensitive to over fitting.
-Here, our simplest model, the Fixed GDD model which uses only a warming component, had the highest correlation of it's single parameter between LTS and NPN datasets.
-This correlation decreases as more parameters are added to models, suggesting that if the NPN dataset is being used to make inferences about plant physiology, simple models could give more consistent results.
-Parameter estimates of more complex models do not agree with models from the same species in LTS datasets, and it is possible even the LTS datasets are over fitting these models.
-
-One goal of ecological forecasting is to inform decision makers about the future state of systems \citep{dietz2018}.
-Thus, the best performing model should be used in these cases to give the best possible forecast.
-In terms of lowest RMSE the GDD and Uniforc The GDD and Uniforc models show promise for building models with the NPN dataset and making plant phenology forecasts across large scales as they were the best performing models for the majority of species and phenophases (fig.
-S1).
-Even with similar parameters, the Fixed GDD model was not the best performing model when tested on held out data.
-While the values of additional parameters may be incorrect in the sense of the prior paragraph, they nonetheless provide more predictive power.
-
+In addition to providing qualitative information about the source of differences between LTS and NPN data, our analysis provides guidance on how best to use phenology data and models currently available to fill the need for forecasts that inform decision makers and provide feedback for researchers \citep{clark2001, dietze2018}.
+Our results suggest that GDD and Uniforc are good choices for large scale forecasting based on NPN data (sensu.
+Scenario D) and that XXX and XXX are good choices for local scale forecasting using LTS data (Scenario A).
 Other models show promise for forecasting under different scenarios.
-For example, if using data from a large spatial extent to make predictions at single sites (sensu.
-Scenario B) our results show that the Alternating model is the best option, as performance under this scenario is very similar to performance of Alternating models built from LTS data (Fig.
+For example, if using data from a large spatial extent to make predictions at single sites (Scenario B) our results show that the Alternating model is the best option, as performance under this scenario is very similar to performance of Alternating models built from LTS data (Fig.
 3).
@@
 4).
-The Alternating model is also one of the best performing model overall for large-scale NPN data (Fig.
-3a).
 
-The models which incorporate a spatial component did not improve over their base models.
-Correspondence between parameter estimates (Fig.
-1), estimates of phenological events (Fig.
-2), and out of sample error rates (Fig.
-3 & 4) for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively.
-The MSB and M1 models had the lowest error in 9 out of 39 occasions, but never by more than a single day over other models (Fig S1).
-Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, yet here models which attempt to account for this variation did not improve estimates overall.
-This could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors.
-\cite{basler2016} used the M1 model to predict budburst on six species across Europe and found it was usually among the best models in terms of RMSE, albeit never by more than a single day.
-Their result was strengthened by a 40 year timeseries across a large region.
-\cite{chuine2017} noted that incorporating the spatial variation in warming requirements as a key issue in future phenology research.
-We feel the NPN dataset, as well as other large-scale phenology datasets, will be key in addressing this and other phenological research needs.
-
-\section*{Conclusion}
-
-Currently the best phenology model for a single location will be one fit using data from only that location, yet these single site models do not transfer well to large scales.
-Making large-scale phenology forecasts will require the incorporation of large-scale observations such as those from the NPN dataset.
-Here we have shown that the NPN dataset is capable of parameterizing phenology models which make comparable predictions to LTS datasets.
-Indeed, if making forecasts across a large spatial extent the NPN derived models are preferable to LTS derived models.
-If using the NPN dataset to make inferences about plant physiology researchers should not use overly complicated models, as parameter estimates from complex NPN derived models do not agree with LTS derived models.
-Whether this disagreement stems from the spatial variation in plant phenology requirements, the irregular sampling of the NPN dataset, or the non-identifiability of models is an avenue for future study.
-
-The National Phenology Network dataset will likely be an important resource for studying phenology.
-The addition of numerous species and observations from across a range of conditions is important for informing large-scale processes, as similar datasets have shown in Europe \citep{olsson2014, basler2016} and Asia \citep{xu2013, zhang2017}.
-Research from long-term small scale studies has been invaluable and should no doubt be continued, but can now be complimented by using large-scale NPN data (ie.
-\citep{jeong2013, melaas2016}).
+In conclusion, our results suggest that both intensive local scale data and large scale citizen science efforts provide valuable information on plant phenology.
+While identifying the most effective data sources for different types and scales of analysis is a useful first step, the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018}.
+Datasets with long, regularly sampled, time series contain valuable information on inter annual variation, while large scale data captures information on numerous species and how phenology varies across the range  \citep{xu2013, olsson2014, basler2016, zhang2017}.
+Our results suggest that methods that can weight LTS data heavily in nearby regions and use large scale data to explicitly capture spatial variation in phenological requirements will help improve both our understanding of phenology and our ability to make predictions.
+Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network, which has a large spatial extent as well as a daily temporal grain \citep{richardson2018}, and phenology information derived from satellite imagery.
+More complex data integration methods that use data from many sources to give the best estimate for forecasts \citep{ogle2015} and continuously update forecasts as new real time data becomes available \citep{luo2011, dietze2018}, provide the best opportunity for accurate inference about, and forecasting of, the timing of biological events.
 
@@
 
+\begin{figure}
+	\centering
+		\includegraphics[width=1\textwidth]{hubbard_harvard_comparison.png}
+	\caption{Distribution of model parameters for the three species common to the Hubbard Brook, Harvard, and NPN datasets.
+The phenophase is budburst for all three species.
+Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution.}
+\end{figure}
+
 \end{document}

### 2018-05-23 analysis/compare_parameters.R
figure edits
take out figure number from filenames
LTS -> LTER in all figures

@@ -223,7 +223,7 @@ legend = cowplot::get_legend(ggplot(filter(parameter_means, model=='uniforc'), a
                                      legend.title = element_text(size = 18),
                                     
                                      legend.key.size = unit(5, units = 'mm')) +
-                               labs(colour = "LTS Dataset", 
+                               labs(colour = "LTER Dataset", 
                                     shape = "Phenophase"))
 
 empty_space = grid::textGrob('')
@@ -238,11 +238,11 @@ complex_layout = rbind(c(2,1,7,7),
 
 #                                      1       2(1)     3(2)       4(3)  5(4)      6(3)         7        8(1)    9(3) 10(3)
 whole_plot=gridExtra::grid.arrange(empty_space,naive, linear_temp, gdd, uniforc, alternating, legend, gdd_fixed, m1, msb, layout_matrix=complex_layout,
-                        left = 'Long Term Study (LTS) Derived Parameter Estimates',
+                        left = 'LTER Derived Parameter Estimates',
                         bottom = 'NPN Derived Parameter Estimates')
 
 
 
-ggsave('manuscript/fig_1_param_comparison.png', plot=whole_plot, height=33, width=20, units = 'cm')
+ggsave('manuscript/figure_param_comparison.png', plot=whole_plot, height=33, width=20, units = 'cm')
 
 

### 2018-05-23 analysis/observation_estimate_comparison.R

@@ -73,7 +73,7 @@ observation_estimate_comparison = predictions %>%
 observation_estimate_comparison = observation_estimate_comparison %>%
   mutate(is_lts_obs = case_when(
     observation_source=='npn'~'At NPN Sites',
-    observation_source!='npn'~'At LTS Sites'
+    observation_source!='npn'~'At LTER Sites'
   ))
 
 model_names = c('gdd','m1','gdd_fixed','linear_temp','naive','alternating','msb','uniforc')
@@ -96,9 +96,9 @@ fig2=ggplot(observation_estimate_comparison, aes(x=doy_estimated_npn_model, y=do
   theme_bw() + 
   theme(legend.position = 'none') + 
   labs(x='Estimates from NPN derived models',
-       y='Estimates from LTS derived models')
+       y='Estimates from LTER derived models')
 
-ggsave(fig2, filename = 'manuscript/fig_2_estimate_compare.png', height = 14, width = 40, units = 'cm')
+ggsave(fig2, filename = 'manuscript/figure_estimate_compare.png', height = 14, width = 40, units = 'cm')
 ################################################################
 # Graph of root mean square different model, phenophase, lts dataset
 # Old graph that might not end up in manuscript

### 2018-05-23 analysis/compare_hubbard_harvard.R
figure edits
take out figure number from filenames
LTS -> LTER in all figures

@@ -3,7 +3,7 @@ library(cowplot)
 config = yaml::yaml.load_file('config.yaml')
 
 ###################################################
-# This produced the supplimentary figure comparing parameters
+# This produced the supplementary figure comparing parameters
 # of hubbard, harvard, and NPN
 #################################################
 
@@ -90,4 +90,4 @@ all_plots =
   (uniforc + no_legend) +
   plot_layout(ncol=1, heights=c(1,1,2,3,4,4))
 
-ggsave('manuscript/hubbard_harvard_comparison.png', plot=all_plots, height=50, width=30, units = 'cm')
+ggsave('manuscript/supplement_hubbard_harvard_comparison.png', plot=all_plots, height=50, width=30, units = 'cm')

### 2018-05-23 analysis/observation_error_analysis.R

@@ -98,6 +98,7 @@ scenarios_error_data = model_errors %>%
     is_npn_model & is_npn_obs ~ 'D'
   ))
 
+# Remove spatial models 
 
 # For the scenario comparison, mark hubbard species and the associated NPN species uniquely so they don't
 # get mixed up with harvard in the scenario comparisons
@@ -130,11 +131,11 @@ rmse_metrics = scenarios_error_data %>%
 
 # The annotated lines
 y_pos_line=0.15
-indicator_lines=data.frame(x=c(-10, 10), xend=c(-25, 25),
+indicator_lines=data.frame(x=c(-12, 12), xend=c(-27, 27),
                            y=c(y_pos_line,y_pos_line), yend=c(y_pos_line,y_pos_line),
                            is_lts_obs='NPN Observations', model_name='GDD')
 y_pos_text=0.12
-indicator_text=data.frame(x=c(-21, 20), y=y_pos_text, t=c('LTS Models\n Better','NPN Models\n Better'),
+indicator_text=data.frame(x=c(-19, 22), y=y_pos_text, t=c('LTER Models\n Better','NPN Models\n Better'),
                           is_lts_obs='NPN Observations', model_name='GDD')
 
 
@@ -155,7 +156,7 @@ rmse_metrics_figure=ggplot(rmse_metrics, aes(metric_value)) +
         axis.title.x = element_text(size=20),
         strip.background = element_rect(fill='grey95'))
 
-ggsave(rmse_metrics_figure, filename = 'manuscript/rmse_metrics_density_plot.png', width = 60, height = 15, units = 'cm')
+ggsave(rmse_metrics_figure, filename = 'manuscript/figure_rmse_metrics_density_plot.png', width = 60, height = 15, units = 'cm')
 
 ######################################################
 ######################################################
@@ -281,11 +282,12 @@ supplement_fig2 = ggplot(supplement_fig2_data, aes(x=model_name, y=error_value, 
   supplement_fig_12_theme
 
 
-ggsave(supplement_fig2, filename = 'manuscript/supplement_best_lts_models.png',
+ggsave(supplement_fig2, filename = 'manuscript/supplement_best_lter_models.png',
        width = 40, height = 50, units = 'cm')
 ############################################
 ############################################
-
+library(kableExtra)
+library(knitr)
 ###
 # Stats for manuscript
 
[…]

### 2018-05-23 manuscript_map.R

@@ -50,8 +50,10 @@ site_map = ggplot() +
         panel.grid = element_blank(),
         axis.ticks = element_blank(),
         panel.border = element_blank(),
-        legend.position = 'none')
+        legend.position = 'none',
+        plot.margin = unit(c(0,0,0,0),'mm'))
 print(site_map)
-ggsave('manuscript/fig_s3_site_map.png', plot=site_map, height=20, width=50, units = 'cm')
+# Note, I saved this manually using the export tool to get ride of uneeded whitespace in the figure.
+ggsave('manuscript/figure_site_map.png', plot=site_map, height=20, width=50, units = 'cm')
 
 

### 2018-05-23 manuscript/main.tex
first large edits from joan
-lots of restructuring and grammar fixes.
-dropped the LTS acronym
-added a table 3 to emphasize differences besides spatial scale
-added separate supplementary image and methods files
-moved all main images to the end for submission

@@
 
-\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
+\documentclass[fleqn,12pt,lineno]{wlpeerj} % for journal submissions
 % \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
+\usepackage{rotating}
+\usepackage{float}
+\usepackage{adjustbox}
 
@@
 
-% \keywords{Keyword1, Keyword2, Keyword3}
+% \keywords{LTER, forecasting, budburst, flowering}
+
+\newpage
 
@@
 \end{abstract}
+
+% This is the double space command
+\renewcommand{\baselinestretch}{1.6}
 
@@
 
+\newpage
 \section*{Introduction}
 
-Plant phenology, the timing of biological events such as flowering, plays an important role in many fields of ecological research extending from local to global scales \citep{cleland2007, richardson2013, tang2016}.
-At large scales, variation in the timing of spring leaf out and fall senescence influence the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long-term climate forecasts \citep{richardson2012}.
-At local scales, species-specific responses to temperature and precipitation can alter flower communities \citep{diez2012, caradonna2014, theobald2017} and affect the abundance and richness of both pollinators \citep{ogilvie2017a, ogilvie2017b} and higher trophic levels \citep{tylianakis2008}.
-Robust cross-scale plant phenology models are needed for understanding the processes governing these key timings of biological events and forecasting the impacts of changes in these timings in response to environmental change.
+Plant phenology, the timing of recurring biological events such as flowering, plays an important role in ecological research extending from local to global scales \citep{cleland2007, richardson2013, tang2016}.
+At large scales, uncertainty in the timing of spring leaf out and fall senescence influence the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long-term climate forecasts \citep{richardson2012}.
+At smaller scales, species-specific responses to temperature and precipitation can alter flower communities \citep{diez2012, caradonna2014, theobald2017} and affect the abundance and richness of both pollinators \citep{ogilvie2017a, ogilvie2017b} and organisms at higher trophic levels \citep{tylianakis2008}.
+Plant phenology models that are robust at multiple ecological scales, or deemed appropriate for a particular scale, are needed to better understand and forecast the timing of key biological events.
 
 Many plant phenology studies use intensively collected datasets collected at a single location over a long time-period by a single research group \citep{cook2012, wolkovich2012, iler2013, roberts2015}.
-These datasets have regular sampling and large numbers of samples over extensive periods of time.
-As a result they represent the biological and climatic variability at that site extremely well.
-However, it is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}.
-This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, and differences in plant age and population density \citep{kramer1995, diez2012}.
-In addition, data from a single location may not be adequate for larger scale phenology modelling due to limited species pools.
-Therefore it has been suggested that accurately understanding and forecasting phenology at larger scales will require models that account for the full range of variation across a species range \citep{richardson2013, chuine2017}.This will necessitate the use of data sources beyond traditional single site studies.
+These datasets have regular sampling and large numbers of samples over long periods of time.
+As a result, the biological and climatic variability at that site is well represented.
+It is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}.
+This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, or differences in plant age, population density, or species pools between sites \citep{kramer1995, diez2012}.
+For these reasons, data from a single location is not adequate for larger scale phenology modeling.
+Accurately forecasting phenology at larger scales will require models that account for the full range of variation across a species' range \citep{richardson2013, chuine2017}, which will necessitate the use of data sources beyond traditional single-site studies.
 
 Data from citizen science projects is becoming increasingly important for ecological research \citep{kelling2009, dickinson2010, tulloch2013}.
-Because this data is often collected by large numbers of volunteers it is possible to gather data at much larger scales than with individual research teams.
-A relatively new citizen science project, The National Phenology Network (NPN) collects phenology observations from volunteers throughout North America and makes the data openly available \citep{schwartz2012a}.
-Data from this project has already been used to study continental-scale variation in oak phenology \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}.
-Large scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the NPN dataset has the potential to meet these needs for North American plant species and communities.
-However, the features that allow citizen science projects to collect data at large-scales can also introduce sources of error including spatial biases toward cities and easily accessible areas and variation in sampling effort and observer skill \citep{dickinson2010}.
-With hundreds of participants across North America, the potential for variation among observers for species identification and dating phenological events is high.
-While volunteers have been shown to be accurate at distinguishing different phenophases \citep{fuccillo2015}, observations are sometimes made sporadically within seasons, across years, and among locations.
-This means that the quality of the data at a specific site will typically be lower than for intensive long-term studies.
+Because this data is often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams.
+A relatively new citizen science project started in 2009, The National Phenology Network (NPN), collects phenology observations from volunteers throughout North America and makes the data openly available \citep{schwartz2012a}.
+Data from this project has already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}.
+Large-scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the NPN dataset has the potential to meet these needs for North American plant species and communities.
+However, the features that allow citizen science projects to collect data at large scales can also introduce spatial biases toward cities and easily-accessible areas, and variation in sampling effort and observer skill \citep{dickinson2010}.
+With hundreds of participants across North America, the potential for variation among observers in their determination of species identification and dating of phenological events is high.
+While volunteers have been shown to be accurate at distinguishing different leaf and flower stages for plants, \citep{fuccillo2015}, observations are sometimes made sporadically across seasons, years, and locations.
+This means that the quality of data at a specific site will typically be lower for citizen science efforts than for intensive, long-term studies.
 
-In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large scale citizen science projects influence both our inferences about biological processes and our ability to predict future phenology events.
-Here, we fit a suite of plant phenology models for the budburst and first flowering phenophases of 24 plant species to data from both the NPN and a set of intensive long-term studies.
+In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes and our ability to predict future phenology events.
+Here, we fit a suite of plant phenology models for the budburst and first-flowering phenophases of 24 plant species to data from both the NPN and a set of intensive long-term studies from the Long Term Ecological Research (LTER) network.
 We compare the resulting models based on both inference about models and parameters and predictions for unobserved events.
-We use this comparison to assess the best methods for both local and large scale phenology modeling and to point the way forward for integrating large scale and local scale data to obtain the best possible models across scales.
+We then use this comparison to assess the best methods for both local and large-scale phenology modeling and to point the way forward for integrating large-scale and local-scale data to determine the best possible models across scales.
 
@@
 
-The National Phenology Network protocol uses status-based monitoring, where observers answer yes, no, or unsure when asked if an individual plant has a specific phenophase present \citep{denny2014}.
+The National Phenology Network protocol uses status-based monitoring, where via an phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}.
 Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves.
-We downloaded all NPN observations from 2009, when collections began, to 2016 for the phenophases Breaking Leaf Buds, Breaking Needle Buds, Emerging Needles, and Open Flowers.
-The first three phenophases apply to the leaf out phase for deciduous broadleafs, evergreen conifers, and pines, respectively.
-The Open Flowers phenophase indicates fully open flowers and applies to all angiosperms.
-Hereafter we will refer to these as either Flowers for the Open Flower phenophase, or Budburst for all other phenophases.
-
-We used moderately strict subsetting of NPN phenology observations as outlined in \cite{crimmins2017}.
-First, "yes" observations for individual plants were only kept if they were preceded by a "no" observation within 30 days.
-Observations for Budburst past day of year (DOY) 172 and for Flowers past DOY 213 were dropped to minimize any influence from outliers.
-Finally, only species that had greater than 30 total observations were kept.
-\cite{crimmins2017} only kept observations that were preceded by a "no" within 15 days, and also grouped multiple individuals at single sites to a single observation.
-We used 30 days to allow for a greater number of species to be compared.
-We did not group together multiple individuals at a single site to better incorporate intra-site variability.
-In addition to these steps we also inferred the date of each phenophase as the midpoint between each "yes" observation and the preceding "no" observation.
-
-To represent long-term intensive phenology studies we used four datasets from North America (Table 1), representing three major ecosystem types and 24 species.
-All four long-term studies are located in the U.S.A and are part of the Long Term Ecological Research network.
-The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S.A.
+Sites in the NPN datasets are located across the U.S.
+and generally clustered around populated areas (Fig 1.).
+To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig.
+1).
+All four long-term studies are located in the U.S.
+and are part of the Long Term Ecological Research network (LTER).
+The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S.
 and are dominated by deciduous broadleaf species.
 The H.J.
-Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S.A.
-The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.A.
-Observation metrics varied among these four datasets due to different protocols, so we converted all metrics to binary "yes" and "no" observations for each phenophase (see supplement for details).
-As with the NPN data, we inferred the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation and only kept species x phenophases which had at least 30 total observations.
+Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S.
+The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.
 
-\renewcommand{\arraystretch}{2}\tabcolsep=5pt
-\begin{center}
+We downloaded all NPN observations from 2009, when collections began, to 2016 for the following phenophases: Breaking Leaf Buds, Breaking Needle Buds, Emerging Needles, and Open Flowers.
+The first three phenophases apply to the 'leaf out' phase for deciduous broadleafs, evergreen conifers, and pines, respectively.
+The 'Open Flowers' phenophase refers to fully-open flowers and applies to all angiosperms.
+Hereafter, we will refer to these as either 'Flowers' for the Open Flower phenophase, or 'Budburst' for all other phenophases.
+We subset the NPN observations similar to methods outlined in \cite{crimmins2017}.
+First, 'yes' observations for individual plants were kept only if they were preceded by a 'no' observation within 30 days.
+Observations for 'Budburst' that were past day of year (DOY) 172, and for 'Flowers' that were past DOY 213 were dropped to minimize any influence from outliers.
+We inferred the observed DOY of each phenophase as the midpoint between each 'yes' observation and the preceding 'no' observation.
+Finally, only species that had greater than 30 total observations were kept.
+\cite{crimmins2017} only kept observations that were preceded by a 'no' within 15 days, and also grouped multiple individuals at single sites to a single observation.
+We used 30 days to allow for a greater number of species to be compared.
+We chose not to group multiple individuals at a single site to better incorporate intra-site variability.
+In addition to these steps,
+
+In the LTER datasets observation metrics varied widely due to different protocols.
+To match the NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods).
+As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophases combinations which had at least 30 total observations.
+After data processing there were 39 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table SX).
+
+\subsection*{Models}
+
+It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}.
+For each of the 39 species and phenophase combinations in the five datasets (NPN and four LTER datasets), we fit eight phenology models (Table 2).
+The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY.
+The \textit{Linear model} uses a regression with the mean spring (Jan.
+1 - March 31) temperature as the independent variable and DOY as the response variable.
+For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$).
+Forcing units are a transformation of the daily mean temperature.
+The start day can either be estimated or fixed.
+For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$.
+The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0C).
+The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan.
+1 ($NCD$).
+The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}.
+These models are some of the most commonly used in phenology research and serve as a suitable baseline for comparing the NPN and LTER datasets.
+
+We also fit two models that attempt to capture spatial variation in phenological requirements.
+The first spatial model, \textit{M1}, is an extension of the \textit{GDD model} which adds a correction in the required forcing using the photoperiod ($L$) \citep{blumel2012}.
+The second, the \textit{Macroscale Species-specific Budburst model (MSB)}, uses the mean spring temperature as a linear correction on the total forcing required in the \textit{Alternating model} \citep{jeong2013}.
+Since there is little to no spatial variation in the LTER datasets, we fit the two spatial models to data from the NPN only.
+We compared the resulting parameters, estimates, and errors for the NPN-derived \textit{M1} and \textit{MSB} models to their non-spatial analogs (the \textit{GDD} and \textit{Alternating models}, respectively) for each species and phenophase in the LTER data.
+
+We parameterized all models using differential evolution to minimize the root mean square error (RMSE) of the estimated DOY of the phenological event.
+Differential evolution is a global optimization algorithm which uses a population of randomly initialized models to find the set of parameters that minimize the RMSE \citep{storn1997}.
+Confidence intervals for parameters were obtained by bootstrapping, in which individual models were re-fit 250 times using a random sample, with replacement, of the data.
+We made predictions by taking the mean DOY estimated from the 250 bootstrapped iterations.
+A random subset consisting of 20\% of observations from each species and phenophase combination was held out from model fitting for later evaluation.
+
+\subsection*{Analysis}
+
+As described above, we fit two sets of models for each species and phenophase: one set of models parameterized using only NPN data, and one set parameterized using only LTER data (with the exception of the \textit{M1} and \textit{MSB} models, see above).
+To compare the inferences about process made by the two datasets, we compared the distribution of each parameter between LTER and NPN-derived models for each species and phenophase combination.
+Using the mean value of each bootstrapped parameter, we also calculated the coefficient of determination ($R^2$) between LTER and NPN-derived models among the 39 species-phenophases.
+In three cases where a species phenophase combination occurred in two LTER sites (Budburst for \textit{Acer saccharum}, \textit{Betula alleghaniensis}, and \textit{Fagus grandifolia} in the Harvard and Hubbard Brook datasets) they were compared separately to the NPN data.
+
+Models with different parameter values, and even entirely different structures, can produce similar estimates for the date of phenological events \citep{basler2016}.
+Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and NPN derived models.
+For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and NPN-derived estimates for estimates made at the four lTER sites and across all NPN sites.
+
+We also directly evaluated model performance using four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) NPN-derived models predicting LTER observations, C) LTER-derived models predicting NPN observations, and D) NPN-derived models predicting NPN observations.
+Within each of these scenarios, we calculated the RMSE of the held-out observations for each species, phenophase, and model type.
+Using the RMSE values, we calculated two different metrics to compare the performance of LTER and NPN-derived models on different data types.
+The first metric focuses on local-scale prediction by comparing the fits of LTER and NPN-derived models on LTER observations: $RMSE_{A} - RMSE_{B}$.
+The second metric focuses on large-scale prediction by comparing the fits of LTER and NPN-derived models on the NPN data: $RMSE_{C} - RMSE_{D}$.
+Negative values indicate that LTER-derived models perform better, while positive values indicate that the NPN-derived model performed better.
+In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison (scenario A) within each site, not across sites, to focus on local scale prediction when LTER data is available.
+Absolute RMSE values are provided in the supplement (Fig.
+S1-S3).
+
+\section*{Results}
+
+The best matches between parameter estimates based on NPN and LTER data were the Fixed GDD model ($R^2$ = 0.49) and the Linear model ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
+The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters but with matching mean parameter values across the two sets of models (Fig.
+2).
+The Naive model showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the NPN dataset (Fig.
+2).
+The large outlier for the Fixed GDD model is \textit{Larrea tridentata}.
+\textit{L.
+tridentata} flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
+While the Fixed GDD and Linear models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between NPN and LTER-derived models (Fig.
+SX).
+
+When comparing estimates of phenological events between the two sets of models, many NPN and LTER models produced similar estimates (Fig.
+3).
+The Fixed GDD model had the highest correlation between the two models sets at NPN sites ($R^2 = 0.82$), while the GDD, M1, and Uniforc models had the highest correlation at LTER sites ($R^2 = 0.54$).
+Comparing models with spatial corrections to the non-spatial alternatives, the MSB (an extension of the alternating model with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the Alternating model.
+The MSB model improved the $R^2$ from 0.38 to 0.45 at LTER sites, and from -0.2 to -0.13 at NPN sites.
+The M1 model (an extension of the GDD model with a spatial correction based on day length) did not improve the correlation between datasets over the GDD model.
+
+When comparing the prediction accuracy on held-out data, NPN-derived models made more accurate predictions for held-out NPN observations, and LTER-derived models performed better on held-out LTER observations (Fig.
+4).
+The Naive and Linear models had the largest differences between the two model sets, while the Fixed GDD model had relatively similar errors when evaluated on both NPN and LTER held-out observations.
+Although the Fixed GDD model had the highest agreement in accuracy between NPN and LTER-derived models, it was not the best performing model overall.
+The GDD and Uniforc models commonly made the best predictions, having the lowest RMSE in 22\% and 40\% of cases among NPN-derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig.
+S1 \& S2).
+
+\section*{Discussion}
+
+Data used to build phenology models typically falls into two categories: intensive long-term data with long time-series at a small number of locations (e.g., LTER data in this study), and large-scale data with less intensive data at hundreds of locations (e.g., NPN data)(Table 3).
+This data scenario--a small amount of intensive data and a large amount of less intensive data--is common in many areas of science and results in the need to understand how to choose between, or combine, data sources \citep{hanks2018}.
+We explored this issue for phenology modeling in realation to making predictions and inferring process from models.
+For inference we found that models based on different data sources resulted in different parameter estimates for all but the simplest models.
+For prediction we found that models fit to different data sources tended to make similar predictions, but that models better predicted out-of-sample data from the data type they were fit to.
+These results are consistent with other research showing that phenology model performance decreases when transferring single-site models to other locations \citep{garcia-mozo2008, xu2013, basler2016}, and with the call for models that better incorporate spatial variation in phenology requirements \citep{richardson2013, chuine2017}.
+Understanding and making predictions for the phenology of a single location is best served by intensive local-scale data, when available, but large-scale datasets work better for extrapolating phenology predictions across a species range.
+Thus, the best choice of both data and models depends on the desired research goals.
+
+In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the Fixed GDD (Fig.
+2).
+These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters.
+First, the differences could result from limitations in the sampling of the NPN dataset, leading to less accurate parameter estimates.
+If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the Fixed GDD model is best for making inferences when NPN data is all that is available.
+Second, spatial variation in phenology requirements could drive these differences, because NPN data integrates over that spatial variation, while LTER data only estimates the phenological requirements for a specific site.
+In this case NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site.
+The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}.
+Third, these differences could result from issues with model identifiability.
+Since different parameter values can yield nearly identical estimates of phenological events, fits to the data can result in parameters differing between datasets even when the underlying processes generating the data are the same.
+Information about which of these issues may be causing the differences between datasets can be explored using this analyses, as will be explained below.
+
+Despite substantial differences in parameter estimates, LTER and NPN-derived models produced similar estimates for phenological events in most cases (Fig.
+3).
+This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues.
+For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 produce nearly identical results in many scenarios.
+This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters.
+In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTER data (Supplementary Material).
+Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Supplementary Material).
+These results are consistent with research showing that models estimating the starting day of warming accumulation from budbreak time-series failed to accurately infer the internal phenology described in the models \citep{chuine2016}.
+\cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting.
+Here, our simplest model, the Fixed GDD model which uses only a warming component, had the highest correlation among parameters between LTER and NPN datasets.
+In combination with this previous research, our results warrant caution in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.
+
+While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information (beyond the Fixed GDD model) based on their ability to make out-of-sample predictions.
+Based on the RMSE, the GDD and Uniforc models produce the best out-of-sample predictions for the majority of species and phenophases at both NPN and LTER datasets (Fig.
+S1 \& S2).
+This demonstrates that the more complex models are capturing additional information about phenology, and that some of the differences between datasets result from differences in either the scales or the sampling of the data.
+Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}.
+In combination with our results showing observed differences in parameter estimates between LTS sites (Supplementary Material), this suggests that variation in phenological requirements across the the range is likely important.
+However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses.
+Correspondence between parameter estimates (Fig.
+2), estimates of phenological events (Fig.
+3), and out-of-sample error rates (Fig.
+4) for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively.
+This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors.
+\cite{basler2016} used the M1 model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day.
+Their result was strengthened by having a 40-year time-series across a large region.
+\cite{chuine2017} listed the incorporation of spatial variation in warming requirements in models as a primary issue in future phenology research.
+Large-scale phenology datasets, like NPN, will be key in addressing this and other phenological research needs.
+
+In conclusion, our results suggest that both LTER and NPN data both provide valuable information on plant phenology.
+The primary difference in the datasets is spatial scale, yet other differences exist in a dichotomy such that no single dataset can "have it all" (Table 3).
+These differences can be strengths or potential limitations.
+Low site fidelity can limit the ability of the NPN dataset to measure long term trends, which is arguably the largest strength of LTER data.
+Representation of only local species and communities limits the use of LTER data in species level predictive modeling.
+The high number of observers contributing to the NPN dataset can potentially cause issues without adequate processing \citep{gerst2016}.
+While identifying the most effective data sources for different types and scales of analysis is a useful first step, the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018, melaas2016}.
+Our results suggest that methods that can weight LTS data heavily in nearby regions and use large-scale data to explicitly capture spatial variation in phenological requirements will help improve both our understanding of phenology and our ability to make predictions.
+Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have a large spatial extent as well as a daily temporal grain \citep{richardson2018}.
+More complex data integration methods can use data from many sources to give the best estimate for forecasts \citep{ogle2015}, and continuously update forecasts as new real-time data becomes available \citep{luo2011, dietze2018}.
+These methods, as well as integration of multiple data sources, provide the best opportunity for accurate inference about, and forecasting of, the timing of biological events.
+
+\section*{Acknowledgments}
+
+\bibliography{refs}
+
+%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%
+%% Begin tables and figures
+%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%
+%% Table 1, the 4 LTER datasets
+%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+
+\begin{adjustbox}{angle=90}
+%%renewcommand{\arraystretch}{2}\tabcolsep=5pt
     \begin{tabular}{ | l | l | l | l |}
@@
     \end{tabular}
-\end{center}
+\end{adjustbox} \newline
+\textbf{Table 1}: LTER datasets used in the analysis
 
-\subsection*{Models}
+%%%%%%%%%%%%%%%%%%%%%%
+%% Table 2, the 8 phenology models
+%%%%%%%%%%%%%%%%%%%%%%
 
-Plant species all have unique phenological requirements, thus it is common to fit multiple models to find the one that best represents a specific species and phenophase \citep{chuine2013}.
-For each species and phenophase in each of the five datasets (NPN and four LTS datasets), we fit a range of phenology models ranging from a naive mean day of year estimate to more complex two phase forcing models (Table 2).
-The naive model uses the mean DOY from prior observations as the estimated DOY.
-The linear model uses a regression with the mean spring (Jan.
-1 - March 31) temperature as the independent variable and DOY as the response variable.
-For the remaining models the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$).
-The start day can either be estimated or fixed.
-For the growing degree day model (GDD) forcing units are the total degrees above the threshold $T$.
-The Fixed GDD model uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0C).
-The Alternating model has a variable number of required forcing units defined as a function of the total number of days below 0C since Jan.
-1 ($NCD$).
-The Uniforc model is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}.
-
-We also fit two models that attempt to capture spatial variation in phenological requirements.
-The first spatial model, M1, is an extension of the GDD model which adds a correction in the required forcing using the photoperiod ($L$) \citep{blumel2012}.
-The second, the Macroscale Species-specific Budburst model (MSB), uses the mean spring temperature as a linear correction on the total forcing required in the Alternating model \citep{jeong2013}.
-Since there is little to no spatial variation in the LTS datasets, the two spatial models were only fit to data from the NPN.
-The resulting parameters, estimates, and errors for the NPN derived M1 and MSB models were compared to their non-spatial analogs (the GDD and Alternating models, respectively) for each species and phenophase in the LTS data.
-
-\begin{center}
-{\def\arraystretch{2}\tabcolsep=5pt
+\newpage
+\begin{adjustbox}{angle=90}
+%%{\def\arraystretch{2}\tabcolsep=5pt
     \begin{tabular}{ | l | c | c | p{1.3cm} | l |}
@@
     Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = \frac{1}{1 + e^{b(T_{i}-c)}} $ & 4 & \citep{chuine2000} \\
-
     \hline
     \end{tabular}
-    }
-\end{center}
+\end{adjustbox} \newline 
+\textbf{Table 2}: Phenology models used in the analysis
 
-All models were parameterized using differential evolution to minimize the root mean square error (RMSE) of the estimated day of year of the phenological event.
-Differential evolution is a global optimization algorithm which uses a population of randomly initialized models to find the set of parameters that minimize the RMSE \citep{storn1997}.
-Confidence intervals for parameters were obtained by bootstrapping, where individual models were re-fit 250 times using a random sample, with replacement, of the data.
-A random 20\% of all data was held out from model fitting for model evaluation.
-Predictions were made by taking the mean DOY from the 250 bootstrapped models.
+%%%%%%%%%%%%%%%%%%%%%%
+%% Table 3, differences between LTER and NPN
+%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection*{Analysis}
+\newpage
 
-As described above, two sets of models were produced for each species and phenophase: one set of models parameterized using only NPN data, and one set parameterized using only LTS data (with the exception of the M1 and MSB models, see above).
-To compare the inferences about process made by the two datasets we compared the distribution for each parameter in each species x phenophase model between LTS and NPN using a Mann-Whitney test (since the majority of parameters having non-normal distributions).
-Using the mean value of each bootstrapped parameter we also calculated the coefficient of determination ($R^2$) between LTS and NPN derived models among the 39 species-phenophases.
-In the three cases where the same species x phenophase combination occurred in two LTS sites, the two instances were compared separately to the NPN data.
+\begin{tabular}{lll}
+                                & \underline{LTER}  & \underline{NPN}           \\
+Time-series length                          & \textbf{High} & \textbf{Low}  \\
+Spatial extent                              & Low           & \textbf{High} \\
+Local species representation                & High          & Low           \\
+Regional/Continental species representation & Low           & \textbf{High} \\
+Number of observers                         & \textbf{Low}  & \textbf{High} \\
+Site fidelity                               & High          & Low          
+\end{tabular}
 
-Models with different parameter values, and even entirely different model structures, can produce similar estimates for the date of phenological events \citep{basler2016}.
-Therefore, to compare the predictions and potential forecasts for models fit to the different datasets we compared the estimated DOY predicted by the LTS and NPN derived models.
-We calculated the coefficient of determination ($R^2$) between LTS and NPN derived estimates for estimates made at the four LTS sites, and across all NPN sites, for each of the 6 models.
+\textbf{Table 3}: Attributes of the two datasets used in this study. Bold text indicates an attribute is expected to increase over time.
 
-We also directly evaluated model performance using four combinations of models and observed data: A) LTS derived models predicting LTS observations, B) NPN derived models predicting LTS observations, C) LTS derived models predicting NPN observations, and D) NPN derived models predicting NPN observations.
-Within each of these scenarios we calculated the RMSE of the held out observations for each species, phenophase, and model type.
-Using the RMSE values we calculated two difference metrics to compare the performance of LTS and NPN derived models on different data types.
-The first metric focuses on local scale prediction by comparing the fits of LTS and NPN derived models on LTS observations: $RMSE_{A} - RMSE_{B}$.
-The second metric focuses on large scale prediction by comparing the fits of LTS and NPN derived models on the NPN data: $RMSE_{C} - RMSE_{D}$.
-Negative values indicate LTS derived models perform better while positive values indicate the NPN derived model performed better.
-In the three cases where the same species x phenophase combination occurred in two LTS sites, the LTS-LTS comparison (scenario A) was made within each site, not across sites, to focus on local scale prediction when LTS data is available.
-Absolute RMSE values are provided in the supplement (Fig.
-S1-S3).
+%%%%%%%%%%%%%%%%%%%%%%
+%% Figure 1, the map
+%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{figure}[]
+\newpage
+
+\textbf{Figure 1}: Locations of National Phenology Network sites used (black points) and Long Term Ecological Research sites (labeled circles).
+
+\newpage
+
+\begin{figure}[H]
 	\centering
-		\includegraphics[width=1\textwidth]{fig_1_param_comparison.png}
-	\caption{Comparisons of parameter estimates between NPN and LTS derived models.
+	\includegraphics[scale=0.6, angle=90]{figure_site_map.png}
+	\caption{.}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%
+%% Figure 2, parameter comparison
+%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+
+\textbf{Figure 2}: Comparisons of parameter estimates between NPN and LTS derived models.
 Each point represents a parameter value for a specific species and phenophase, and is the mean value from 250 bootstrapped models.
 The black line is the 1:1 line.
-The $R^2$ is the coefficient of determination, which can be negative if the relationship between the two parameter sets is worse than no relationship but with the same mean values.}
+The $R^2$ is the coefficient of determination, which can be negative if the relationship between the two parameter sets is worse than no relationship but with the same mean values.
+
+\newpage
+
+\begin{figure}[H]
+	\centering
+		\includegraphics[width=1\textwidth]{figure_param_comparison.png}
+    \caption{.}
 \end{figure}
 
-\section*{Results}
+%%%%%%%%%%%%%%%%%%%%%%
+%% Figure 3, estimate comparison (1:1 graphs w/ yellow and
+%% green points
+%%%%%%%%%%%%%%%%%%%%%%
 
-The best matches between parameter estimates based on NPN and LTS data were the Fixed GDD model ($R^2$ = 0.49) and the Linear model ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
-The parameters for all other models had $R^2$ values < 0 indicating that the relationship was worse than no relationship between the parameters but with matching mean parameter values across the two sets of models (Fig.
-1).
-The Naive model showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTS datasets being mostly in the northern United States.
-The one major outlier for the Fixed GDD model, \textit{Larrea tridentata}, has phenology largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
-While the Fixed GDD and Linear models showed reasonable correspondence between parameter estimates, all parameters for individual species x phenophase combinations were statistically different between NPN and LTS derived models according to the Mann-Whitney U test (p < 0.001).
+\newpage
 
-\begin{figure}
+\textbf{Figure 3}: Comparison of predicted day of year (DOY) of all phenological events between NPN and LTS-derived models.
+Top panels show comparisons at LTS sites and bottom panels show comparisons at NPN sites.
+Each point is an estimate for a single held-out observation.
+Colors indicate observations for a single species x phenophase combination.
+
+\newpage
+
+\begin{figure}[H]
 	\centering
-		\includegraphics[width=1\textwidth]{fig_2_estimate_compare.png}
-	\caption{Comparison of predicted day of year (DOY) of all phenological events between NPN and LTS derived models.
-Top panels show comparisons at LTS sites and bottom panels show comparisons at NPN sites.
-Each point is an estimate for a single held out observation.
-Colors indicate observations for a single species x phenophase combination.}
+		\includegraphics[scale=0.5, angle=90]{figure_estimate_compare.png}
+	\caption{.}
 \end{figure}
 
-When comparing estimates of phenological events between the two sets of models, many NPN and LTS models produced similar estimates (Fig.
-2).
-The Fixed GDD model had the highest correlation between the two models sets at NPN sites ($R^2 = 0.82$), while the GDD, M1, and Uniforc models had the highest correlation at LTS sites ($R^2 = 0.54$).
-Comparing models with spatial corrections to the non-spatial alternatives, the MSB (an extension of the alternating model with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the Alternating model.
-The MSB model improved the $R^2$ from 0.38 to 0.45 at LTS sites, and from -0.2 to -0.13 at NPN sites.
-The M1 model (an extension of the GDD model with a spatial correction based on daylength) did not improve the correlation between datasets over the GDD model.
+%%%%%%%%%%%%%%%%%%%%%%
+%% Figure 4, error comparison (2 rows of density plots)
+%%%%%%%%%%%%%%%%%%%%%%
 
-When comparing the prediction accuracy on held out data NPN derived models make more accurate predictions for held out NPN observations, and LTS derived models perform better on held out LTS observations (Fig.
-3).
-The Naive and Linear models had the largest differences between the two model sets, while the Fixed GDD model had relatively similar errors when evaluated on both NPN and LTS held out observations.
-Though the Fixed GDD model had the highest agreement in accuracy between NPN and LTS derived models, it was not the best performing model overall.
-The GDD and Uniforc models commonly made the best predictions, having the lowest RMSE in 22\% and 40\% of cases among NPN derived models, and 42\% and 32\% of cases among LTS derived models, respectively (Fig.
-S1 \& S2).
+\newpage
 
-\begin{figure}
+\textbf{Figure 4}: Differences in prediction error between NPN and LTS-derived models.
+Density plots for comparisons of predictions on LTS data (top row) and NPN data (bottom row).
+Each plot represents the difference between the RMSE for LTS-derived model and the NPN-derived model, meaning that values less than zero indicate more accurate prediction by LTS-derived models and values greater than zero indicate more accurate prediction by NPN-derived models.
+Differences are calculated pairwise for the 39 species/phenophase comparisons.
+
+\newpage
+
+\begin{figure}[H]
 	\centering
-		\includegraphics[width=1\textwidth]{rmse_metrics_density_plot.png}
-	\caption{Differences in prediction error between NPN and LTS derived models.
-Density plots for comparisons of predictions on LTS data (top row) and NPN data (bottom row).
-Each plot represents the difference between the RMSE for LTS derived model and the NPN derived model, meaning that values less than zero indicate more accurate prediction by LTS derived models and values greater than zero indicate more accurate prediction by NPN derived models.
-Differences are calculated pairwise for the 39 species/phenophase comparisons.}
-\end{figure}
-
-\section*{Discussion}
-
-Data for building phenology models typically falls into two categories: intensive long-term data with long time-series at a small number of locations (e.g., the LTS data in this study) and large scale data with less intensive data at hundreds of locations (e.g., the NPN data).
-This data scenario, a small amount of intensive data and a large amount of less intensive data, is common in many areas of science, and results in the need to understand how to choose between, or combine, data sources.
-We explored this issue for phenology modeling at local and continental scales for making predictions and inferring process from models.
-For inference, we found that models based on different data sources resulted in different parameter estimates for all but the simplest models.
-For prediction we found that models fit to different data sources tended to make similar predictions, but that models better predicted out of sample data from the data type they were fit to.
-As a result the best choice of both data and models depends on the desired research goals.
-Understanding and making predictions for the phenology of a single location is best served by intensive local scale data, if it is available, but large scale datasets work better for extrapolating phenology predictions across a species range.
-These results are consistent with other research showing that phenology model performance decreases when transferring single site models to other locations \citep{garcia-mozo2008, xu2013, basler2016} and with the call for models that better incorporate spatial variation in phenology requirements \citep{richardson2013, chuine2017}.
-
-In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process oriented model: the Fixed GDD (Fig.
-1).
-These differences may be caused by a variety of factors that have different implications for interpreting process oriented models and their parameters.
-First, the differences could result from limitations in the sampling of the NPN dataset, leading to less accurate parameter estimates.
-If this is the case it would suggest using local scale data when it is available for making inferences about plant physiology and focusing on the Fixed GDD model when large scale data is all that is available.
-Second, spatial variation in phenology requirements could drive these differences, because large scale data integrates over that spatial variation, while local scale data only estimates the phenological requirements for a specific site.
-In this case large scale would provide a better estimate of the general phenological requirements of a species, but local scale data would provide a more accurate understanding for a single site.
-The best solution to this issue would be the development of models that accurately incorporate spatial variation.
-Finally, these differences could result from issues with model identifiability, where different parameter values yield nearly identical fits to the data, which can result in fits to different datasets varying widely even if the underlying processes generating the data are the same.
-
-Information about which of these issues may be causing the differences between datasets can be explored using our analyses.
-Despite the substantial differences in parameter estimates the LTS and NPN derived models produced similar estimates for phenological events in most cases (Fig.
-2).
-This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues.
-For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 produce nearly identical results in many scenarios.
-This possibility is supported by the fact that the the highest correlation between parameter estimates is seen in a models with only 1 or 2 parameters.
-In addition, the bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTS data (Supplementary Material).
-Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTS datasets (Supplementary Material).
-These results are consistent with research showing that models estimating endodormancy break from budbreak time-series failed to accurately infer the internal phenology described in the models \citep{chuine2016}.
-\cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over fitting.
-Here, our simplest model, the Fixed GDD model which uses only a warming component, had the highest correlation among parameters between LTS and NPN datasets.
-In combination with this previous research, our results warrant caution in interpreting the parameter estimates from complex phenology models regardless of the data source used for fitting the models.
-
-While complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information beyond the Fixed GDD model based on their ability to make out-of-sample predictions.
-Based on the RMSE, the GDD and Uniforc models produce the best out-of-sample predictions for the majority of species and phenophases at both large and local scales (Fig.
-S1).
-For local scale predictions This demonstrates that the more complex models are capturing additional information about phenology and that some of the differences between datasets result from differences in either the scales or the sampling of the data.
-Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, and in combination with our results showing observed differences in parameter estimates between LTS sites (Supplementary Material) this suggests that variation in phenological requirements across the the range is likely important.
-However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses.
-Correspondence between parameter estimates (Fig.
-1), estimates of phenological events (Fig.
-2), and out of sample error rates (Fig.
-3) for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively.
-This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors.
-\cite{basler2016} used the M1 model to predict budburst on six species across Europe and found it was usually among the best models in terms of RMSE, albeit never by more than a single day.
-Their result was strengthened by having a 40 year timeseries across a large region.
-\cite{chuine2017} noted that incorporating the spatial variation in warming requirements as a primary issue in future phenology research.
-Large-scale phenology datasets, like NPN, will be key in addressing this and other phenological research needs.
-
-In addition to providing qualitative information about the source of differences between LTS and NPN data, our analysis provides guidance on how best to use phenology data and models currently available to fill the need for forecasts that inform decision makers and provide feedback for researchers \citep{clark2001, dietze2018}.
-Our results suggest that GDD and Uniforc are good choices for large scale forecasting based on NPN data (sensu.
-Scenario D) and that XXX and XXX are good choices for local scale forecasting using LTS data (Scenario A).
-Other models show promise for forecasting under different scenarios.
-For example, if using data from a large spatial extent to make predictions at single sites (Scenario B) our results show that the Alternating model is the best option, as performance under this scenario is very similar to performance of Alternating models built from LTS data (Fig.
-3).
-In the opposite scenario of making large-scale predictions using only data from a single site (sensu.
-Scenario C) the Fixed GDD model is likely the best option, as most LTS derived Fixed GDD models have a RMSE only 5 days greater, on average, than the corresponding NPN model (Fig.
-4).
-
-In conclusion, our results suggest that both intensive local scale data and large scale citizen science efforts provide valuable information on plant phenology.
-While identifying the most effective data sources for different types and scales of analysis is a useful first step, the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018}.
-Datasets with long, regularly sampled, time series contain valuable information on inter annual variation, while large scale data captures information on numerous species and how phenology varies across the range  \citep{xu2013, olsson2014, basler2016, zhang2017}.
-Our results suggest that methods that can weight LTS data heavily in nearby regions and use large scale data to explicitly capture spatial variation in phenological requirements will help improve both our understanding of phenology and our ability to make predictions.
-Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network, which has a large spatial extent as well as a daily temporal grain \citep{richardson2018}, and phenology information derived from satellite imagery.
-More complex data integration methods that use data from many sources to give the best estimate for forecasts \citep{ogle2015} and continuously update forecasts as new real time data becomes available \citep{luo2011, dietze2018}, provide the best opportunity for accurate inference about, and forecasting of, the timing of biological events.
-
-\section*{Acknowledgments}
-
-\bibliography{refs}
-
-\setcounter{figure}{0}    
-\section*{Supplement}
-\begin{figure}
-	\centering
-		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
-	\caption{RMSE for specific species and phenophases of the NPN dataset. X marks the best performing models for the respective data type.}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[width=1\textwidth]{supplement_best_lts_models.png}
-	\caption{RMSE for specific species and phenophases of the LTS datasets. X marks the best performing models for the respective data type.}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
-	\caption{RMSE of all species and phenophases of the four scenarios described in the text. These values are calculated using held out test data.}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[width=1\textwidth]{fig_s3_site_map.png}
-	\caption{Locations of National Phenology Network sites used (black points) and Long Term Study sites (labeled circles).}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[width=1\textwidth]{hubbard_harvard_comparison.png}
-	\caption{Distribution of model parameters for the three species common to the Hubbard Brook, Harvard, and NPN datasets.
-The phenophase is budburst for all three species.
-Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution.}
+		\includegraphics[scale=0.4, angle=90]{figure_rmse_metrics_density_plot.png}
+	\caption{.}
 \end{figure}

### 2018-05-23 manuscript/supplement_images.tex
first large edits from joan
-lots of restructuring and grammar fixes.
-dropped the LTS acronym
-added a table 3 to emphasize differences besides spatial scale
-added separate supplementary image and methods files
-moved all main images to the end for submission

\documentclass[a4paper,12pt]{article}

\usepackage{graphicx}
\usepackage{a4wide}
\usepackage{lineno}
\usepackage{setspace}
  \doublespacing
\usepackage[document]{ragged2e}

\begin{document}
\centering{\Large Paper title goes here \par}

\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White}

\textbf{\centering{\large Supplementary materials \par}} \newline
Supplementary images S1 - S4

%%%%%%%%%%%%%%%%%
%% Figure S1
%%%%%%%%%%%%%%%%%
\textbf{Figure S1}: RMSE for specific species and phenophases of the NPN dataset. X marks the best performing models for the respective data type.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
\end{center}

\newpage
%%%%%%%%%%%%%%%%%
%% Figure S2
%%%%%%%%%%%%%%%%%
\textbf{Figure S2}: RMSE for specific species and phenophases of the LTER datasets. X marks the best performing models for the respective data type.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_best_lter_models.png}
	\caption{Figure S2}
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S3
%%%%%%%%%%%%%%%%%

\newpage

\textbf{Figure S3}: RMSE of all species and phenophases of the four scenarios described in the text. These values are calculated using held out test data.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
	\caption{Figure S3}
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S4
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S4}: Distribution of model parameters for the three species common to the Hubbard Brook, Harvard, and NPN datasets. The phenophase is budburst for all three species. Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_hubbard_harvard_comparison.png}
	\caption{Figure S4}
\end{center}

\end{document}}

### 2018-05-23 manuscript/supplement_methods.tex
first large edits from joan
-lots of restructuring and grammar fixes.
-dropped the LTS acronym
-added a table 3 to emphasize differences besides spatial scale
-added separate supplementary image and methods files
-moved all main images to the end for submission

\documentclass[a4paper,12pt]{article}

\usepackage{graphicx}
\usepackage{a4wide}
\usepackage{lineno}
    \linenumbers
%\usepackage{setspace}
%  \doublespacing
\usepackage[document]{ragged2e}

\begin{document}
\begin{center}
{\huge Paper title goes here \par} \newline

\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White} \newline

{\Large \textbf{Supplementary materials}} \newline
Methods describing processing of the 4 LTER datasets
\end{center}

The four LTER datasets each had different protocols for recording phenology observations. Below are details for converting the data from each to a status based yes/no as used in the National Phenology Network. As in the NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation.

\textbf{Harvard Forest} \newline
Observations here were recorded as relative percentage of flowering or budburst for individual plants. "Yes" observations budburst and flower where set to the DOY when the percentage of each tree had was greater than or equal to 10\%

\textbf{H.J. Andrews Experimental Forest} \newline
"yes" observations for budburst was the first DOY when an individual was marked as "bud break". "yes" observations for flowering was was when an indiviudal was first marked as "Flowers open". Note that each species has slightly different ordinal codes to make each of these events.

\textbf{Jornada Experimental Range} \newline
Observations for this dataset represent, for each zone, the percent of plants for a particular species which is observed within each phenophase. "Yes" observations for flowers was the first DOY where the flower phenophase was 10\% or greater.

\textbf{Hubbard Brook} \newline
Observations for this dataset represent for each species the average, among 3 individuals, of an ordinal description of phenophase. \newline

0: winter conditions \newline
1: bud swelling \newline
2: small leaves or flowers \newline
3: leafs 1/2 of final length, leafs obscure half the sky as seen thru crown \newline
3.5: leaves 3/4 expanded, sky mostly obscured, crown not yet in summer condition \newline
4: fully expanded, canopy in summer conditions \newline

"Yes" observations for budburst were the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].

\end{document}}

### 2018-05-24 analysis/observation_estimate_comparison.R
some minor data fixes in the estimate comparison
put pine (phenophse 496) budburst into the estimate figure
exclude the 3 species common to harvard/hubbard

@@ -24,6 +24,7 @@ phenophase,phenophase_type
 371,Budburst
 480,Budburst
 488,Budburst
+496,Budburst
 501,Flowers')
 
 predictions = predictions %>%
@@ -65,6 +66,9 @@ npn_model_estimates = predictions %>%
 
 observation_estimate_comparison = predictions %>%
   filter(parameter_source != 'npn') %>%
+  # Exclude the 3 common species between hubbard and harvard in this comparison
+  filter(!(observation_source == 'hubbard' & parameter_source == 'harvard')) %>%
+  filter(!(observation_source == 'harvard' & parameter_source == 'hubbard')) %>%
   rename(doy_estimated_non_npn_model = doy_estimated, non_npn_parameter_source = parameter_source) %>%
   left_join(npn_model_estimates, by=c('model_name','observation_source','species','observation_id','phenophase')) %>%
   filter(complete.cases(.)) %>%

### 2018-05-24 analysis/compare_parameters.R
move the species stats to a separate file
these are the total species used, total comparisons,
and suppliment table 1 (the species list).

@@ -1,5 +1,4 @@
 library(tidyverse)
-library(broom)
 library(cowplot)
 config = yaml::yaml.load_file('config.yaml')
 
@@ -21,15 +20,6 @@ npn_species = all_parameters %>%
 
 all_parameters = all_parameters %>% 
   filter(paste(species,phenophase) %in% paste(npn_species$species,npn_species$phenophase) )
-
-# This is the final numbers put into the Table 1
-lts_sample_sizes = all_parameters %>%
-  filter(dataset!='npn') %>%
-  select(dataset, species, phenophase) %>%
-  distinct() %>% group_by(dataset, phenophase) %>%
-  summarize(n_species = n_distinct(species))
-print(paste0('total unique species: ',length(unique(all_parameters$species))))
-print(paste0('total species/phenophase comparisons: ',sum(lts_sample_sizes$n_species)))
 
 ############################################################################
 # Organize the two spatial models, MSB and M1
@@ -91,6 +81,7 @@ ggplot(x, aes(x=value, group=dataset, fill=dataset)) +
   facet_wrap(parameter_name~model~species~phenophase, scales = 'free')
 
 ###########################################################################
+# Mann Whitney and/or ks test for parameter distribution comparison.
 # Are these results robust to a sample size smaller than 250 bootstraps?
 # 
 # all_parameters_subset = all_parameters %>%
@@ -115,7 +106,7 @@ ggplot(x, aes(x=value, group=dataset, fill=dataset)) +
 ###############################################################################
 #scatter plots of npn vs long term datasets
 
-budburst_phenophases = c(371, 496, 488)
+budburst_phenophases = c(371, 496, 488, 480)
 flower_phenophases = c(501)
 
 parameter_means = all_parameters %>%

### 2018-05-24 analysis/species_statistics.R

@@ -1,0 +1,53 @@
+library(tidyverse)
+config = yaml::yaml.load_file('config.yaml')
+
+all_species = read_csv(config$model_parameter_file) %>%
+  filter(!parameter_name %in% c('run_time','num_iterations')) %>%
+  select(dataset, species) %>%
+  distinct()
+
+#Pull out phenophase
+all_species = all_species %>% 
+  mutate(phenophase = stringr::word(species,2,2, ' - '),
+         species = stringr::word(species,1,1,' - '))
+
+all_species$phenophase = as.numeric(all_species$phenophase)
+
+#Keep only species/phenophases that are present in NPN dataset
+npn_species = all_species %>%
+  filter(dataset == 'npn') %>%
+  select(species, phenophase) %>%
+  distinct()
+
+all_species = all_species %>% 
+  filter(paste(species,phenophase) %in% paste(npn_species$species,npn_species$phenophase) )
+
+# This is the final numbers put into the Table 1
+lts_sample_sizes = all_species %>%
+  filter(dataset!='npn') %>%
+  select(dataset, species, phenophase) %>%
+  distinct() %>% 
+  group_by(dataset, phenophase) %>%
+  summarize(n_species = n_distinct(species))
+print(paste0('total unique species: ',length(unique(all_species$species))))
+print(paste0('total species/phenophase comparisons: ',sum(lts_sample_sizes$n_species)))
+
+# Suppliment table of species names
+library(kableExtra)
+library(knitr)
+
+phenophase_types = read.table(header=TRUE, sep=',', stringsAsFactors = FALSE, text='
+phenophase,phenophase_type
+371,Budburst
+480,Budburst
+488,Budburst
+496,Budburst
+501,Flowers')
+
+suppliment_species_table = all_species %>%
+  filter(dataset!='npn') %>%
+  left_join(phenophase_types) %>%
+  mutate(present='x') %>%
+  spread(dataset, present,fill='')
+
+kable(suppliment_species_table, 'latex')

### 2018-05-25 analysis/observation_error_analysis.R
cleanup extraneous stuff

@@ -98,7 +98,6 @@ scenarios_error_data = model_errors %>%
     is_npn_model & is_npn_obs ~ 'D'
   ))
 
-# Remove spatial models 
 
 # For the scenario comparison, mark hubbard species and the associated NPN species uniquely so they don't
 # get mixed up with harvard in the scenario comparisons

### 2018-05-25 analysis/observation_estimate_comparison.R

@@ -103,80 +103,3 @@ fig2=ggplot(observation_estimate_comparison, aes(x=doy_estimated_npn_model, y=do
        y='Estimates from LTER derived models')
 
 ggsave(fig2, filename = 'manuscript/figure_estimate_compare.png', height = 14, width = 40, units = 'cm')
-################################################################
-# Graph of root mean square different model, phenophase, lts dataset
-# Old graph that might not end up in manuscript
-
-# estimate_differences_lts_observations = observation_estimate_comparison %>%
-#   group_by(model_name, observation_source, non_npn_parameter_source, species, phenophase) %>%
-#   summarise(rmsd = sqrt(mean(model_estimate_difference^2))) %>%
-#   ungroup() %>%
-#   filter(observation_source == non_npn_parameter_source)
-# 
-# estimate_differences_npn_observations = observation_estimate_comparison %>%
-#   group_by(model_name, observation_source, non_npn_parameter_source, species, phenophase) %>%
-#   summarise(rmsd = sqrt(mean(model_estimate_difference^2))) %>%
-#   ungroup() %>%
-#   filter(observation_source == 'npn')
-
-# Apply more pleasing names to everything for figures
-# model_names = c('gdd','m1','gdd_fixed','linear_temp','naive','alternating','msb','uniforc')
-# pretty_model_names = c('GDD','M1','Fixed GDD','Linear','Naive','Alternating','MSB','Uniforc')
-# datasets = c('harvard','hjandrews','hubbard','jornada','npn')
-# pretty_dataset_names = c('Harvard Forest','H.J. Andrews','Hubbard Brook','Jornada','NPN')
-# 
-# estimate_differences_lts_observations$model_name = factor(estimate_differences_lts_observations$model_name, levels=model_names, labels=pretty_model_names)
-# estimate_differences_npn_observations$model_name = factor(estimate_differences_npn_observations$model_name, levels=model_names, labels=pretty_model_names)
-# estimate_differences_lts_observations$non_npn_parameter_source = factor(estimate_differences_lts_observations$non_npn_parameter_source,
-#                                                                            levels=datasets, labels=pretty_dataset_names)
-# estimate_differences_npn_observations$non_npn_parameter_source = factor(estimate_differences_npn_observations$non_npn_parameter_source,
[…]

### 2018-05-25 analysis/compare_hubbard_harvard.R
split hubbard/harvard figure into 2

@@ -53,7 +53,7 @@ single_model_plot = function(model_name, plot_title){
     summarise(mean_value = mean(value), median_value=median(value)) %>%
     ungroup() %>%
     gather(summary_metric, summary_value, mean_value, median_value)
-  p=ggplot(plot_data, aes(value, fill=dataset)) +
+  p=ggplot(plot_data, aes(x=value, fill=dataset)) +
     geom_density(position = position_identity(), alpha=0.7) +
     geom_vline(data=summary_lines, aes(xintercept=summary_value, color=dataset, linetype=summary_metric), size=1) +
     scale_fill_manual(values=color_pallete) +
@@ -81,13 +81,17 @@ plot(linear)
 
 no_legend = theme(legend.position = 'none')
 
-all_plots = 
+hubbard_hubbard_1 = 
   (naive + no_legend)+
   (gdd_fixed + no_legend) + 
-  (linear + no_legend) + 
-  gdd + 
-  (alternating + no_legend) + 
-  (uniforc + no_legend) +
-  plot_layout(ncol=1, heights=c(1,1,2,3,4,4))
+  (linear) + 
+  (gdd + no_legend) + 
+  plot_layout(ncol=1, heights=c(1,1,2,3))
 
-ggsave('manuscript/supplement_hubbard_harvard_comparison.png', plot=all_plots, height=50, width=30, units = 'cm')
+hubbard_hubbard_2 = 
+  (alternating) + 
+  (uniforc + no_legend) +
+  plot_layout(ncol=1, heights=c(1,1))
+
+ggsave('manuscript/supplement_hubbard_harvard_comparison1.png', plot=hubbard_hubbard_1, height=40, width=30, units = 'cm')
+ggsave('manuscript/supplement_hubbard_harvard_comparison2.png', plot=hubbard_hubbard_2, height=40, width=30, units = 'cm')

### 2018-05-25 analysis/plot_select_species_parameters.R
select species parameter comparison
this makes the supplementary figures 6 & 7, and should
hopefully be easy for other people make figures for
other species.

@@ -1,0 +1,95 @@
+library(tidyverse)
+library(cowplot)
+config = yaml::yaml.load_file('config.yaml')
+
+###################################################
+# This produced the supplementary figure comparing parameters
+# of 4 species. Adjust the ones below to create different plots. 
+#################################################
+
+select_species = c('acer circinatum','populus tremuloides','acer rubrum','vaccinium parvifolium')
+select_phenophase = c(371)
+#models_of_interest = c('gdd_fixed','gdd','alternating')
+
+all_parameters = read_csv(config$model_parameter_file) %>%
+  filter(!parameter_name %in% c('run_time','num_iterations')) 
+
+#Pull out phenophase
+all_parameters = all_parameters %>% 
+  mutate(phenophase = stringr::word(species,2,2, ' - '),
+         species = stringr::word(species,1,1,' - '))
+
+all_parameters$phenophase = as.numeric(all_parameters$phenophase)
+
+select_species_params = all_parameters %>%
+  filter(species %in% select_species, 
+         phenophase %in% select_phenophase)
+
+########################################################################
+datasets = c('harvard','hubbard','npn','hjandrews','jornada')
+pretty_dataset_names = c('Harvard Forest','Hubbard Brook','NPN','H.J. Andrews','Jornada')
+select_species_params$dataset = factor(select_species_params$dataset, levels = datasets, labels = pretty_dataset_names)
+############################################################################
+
+common_plot_theme = theme(strip.text = element_text(size=10),
+                          strip.background = element_rect(fill='grey95'),
+                          axis.text = element_text(size=12),
+                          axis.title.y = element_text(size=18))
+
+point_size=4
+point_shapes = c(17,13)
+color_pallete=c("grey42", "#56B4E9", "#009E73")
+
+
+single_model_plot = function(model_name, plot_title, num_column){
+  plot_data = select_species_params %>%
+    filter(model==model_name)
+  summary_lines = plot_data %>%
+    group_by(dataset, parameter_name, species) %>%
+    summarise(mean_value = mean(value), median_value=median(value)) %>%
+    ungroup() %>%
+    gather(summary_metric, summary_value, mean_value, median_value)
[…]

### 2018-05-30 manuscript/main.tex
final revisions from coauthors
mostly minor edits from ethan,joan,michael, and kristina
added abstract
formated using a better latex template

@@
 
-\documentclass[fleqn,12pt,lineno]{wlpeerj} % for journal submissions
-% \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
+\documentclass[fleqn,12pt,lineno]{article} 
+\usepackage{Ecology}
 \usepackage{rotating}
@@
 \usepackage{adjustbox}
+\RequirePackage{natbib}
+\bibliographystyle{apalike}
 
@@
 \affil[4]{Informatics Institute, University of Florida, Gainesville, FL, United States}
-\corrauthor[1]{Shawn D. Taylor}{shawntaylor@weecology.org}
 
-% \keywords{LTER, forecasting, budburst, flowering}
-
+\date{}
 \newpage
-
-\begin{abstract}
-\end{abstract}
 
@@
 \maketitle
+Corresponding author: Shawn D. Taylor - shawntaylor@weecology.org
+
 \thispagestyle{empty}
+\newpage
+
+% Limit 350 words
+\begin{abstract}
+Large-scale observational data from citizen science efforts are becoming increasingly common in ecology, and researchers often choose between these and data from intensive local-scale studies for their analyses.
+This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability.
+Here we explore this issue with phenology by comparing models built using data from the large-scale, citizen science National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites.
+We built process based phenology models for species common to each dataset.
+From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both NPN and LTER data.
+We found that model parameter estimates for the same species are most similar between the two datasets when using simple models, but parameter estimates vary widely as model complexity increases.
+Despite this, estimates for the date of phenological events and out-of-sample errors are similar, regardless of the model chosen.
+Predictions for NPN data had the lowest error when using models built from the NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built.
+Accordingly, the choice of dataset depends on the research question.
+Inferences about species-specific phenological requirements are best made with LTER data, and if NPN or similar data are all that is available, then analyses should be limited to simple models.
+Large-scale predictive modeling is best done with the larger-scale NPN data, which has high spatial representation and a large regional species pool.
+LTER datasets, on the other hand, have high site fidelity and thus characterize inter-annual variability extremely well.
+Future research aimed at forecasting phenology events for particular species over larger scales should develop models which integrate the strengths of both datasets.
+\end{abstract}
+
+\keywords{LTER, NPN, forecasting, budburst, flowering, data integration, scale}
 
@@
 
-Many plant phenology studies use intensively collected datasets collected at a single location over a long time-period by a single research group \citep{cook2012, wolkovich2012, iler2013, roberts2015}.
+Many plant phenology studies use intensively collected datasets from a single location over a long time-period by a single research group \citep{cook2012, wolkovich2012, iler2013, roberts2015}.
 These datasets have regular sampling and large numbers of samples over long periods of time.
@@
 It is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}.
-This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, or differences in plant age, population density, or species pools between sites \citep{kramer1995, diez2012}.
+This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, or differences in plant age or population density \citep{kramer1995, diez2012}.
 For these reasons, data from a single location is not adequate for larger scale phenology modeling.
@@
 Because this data is often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams.
-A relatively new citizen science project started in 2009, The National Phenology Network (NPN), collects phenology observations from volunteers throughout North America and makes the data openly available \citep{schwartz2012a}.
+A relatively new citizen science project started in 2009, The National Phenology Network (NPN), collects phenology observations from volunteers throughout North America and makes the data openly available \citep{npncitation}.
 Data from this project has already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}.
@@
 While volunteers have been shown to be accurate at distinguishing different leaf and flower stages for plants, \citep{fuccillo2015}, observations are sometimes made sporadically across seasons, years, and locations.
-This means that the quality of data at a specific site will typically be lower for citizen science efforts than for intensive, long-term studies.
+This means that the quanity and quality of data at a specific site will typically be more variable for citizen science efforts than for intensive, long-term studies.
 
-In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes and our ability to predict future phenology events.
+In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology and our ability to predict future phenology events.
 Here, we fit a suite of plant phenology models for the budburst and first-flowering phenophases of 24 plant species to data from both the NPN and a set of intensive long-term studies from the Long Term Ecological Research (LTER) network.
 We compare the resulting models based on both inference about models and parameters and predictions for unobserved events.
-We then use this comparison to assess the best methods for both local and large-scale phenology modeling and to point the way forward for integrating large-scale and local-scale data to determine the best possible models across scales.
+We then use this comparison to assess the best methods for both local- and large-scale phenology modeling and to point the way forward for integrating large-scale and local-scale data to determine the best possible models across scales.
 
@@
 Sites in the NPN datasets are located across the U.S.
-and generally clustered around populated areas (Fig 1.).
+and generally clustered around populated areas (Fig.
+1).
 To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig.
@@
 We chose not to group multiple individuals at a single site to better incorporate intra-site variability.
-In addition to these steps,
 
@@
 As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophases combinations which had at least 30 total observations.
-After data processing there were 39 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table SX).
+After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table S1).
 
@@
 It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}.
-For each of the 39 species and phenophase combinations in the five datasets (NPN and four LTER datasets), we fit eight phenology models (Table 2).
+For each of the 38 species and phenophase combinations in the five datasets (NPN and four LTER datasets), we fit eight phenology models (Table 2).
 The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY.
@@
 
+We extracted corresponding daily mean temperature for all NPN and LTER observations from the gridded PRISM dataset \citep{prismdata}.
 We parameterized all models using differential evolution to minimize the root mean square error (RMSE) of the estimated DOY of the phenological event.
@@
 To compare the inferences about process made by the two datasets, we compared the distribution of each parameter between LTER and NPN-derived models for each species and phenophase combination.
-Using the mean value of each bootstrapped parameter, we also calculated the coefficient of determination ($R^2$) between LTER and NPN-derived models among the 39 species-phenophases.
+Using the mean value of each bootstrapped parameter, we also calculated the coefficient of determination ($R^2$) between LTER and NPN-derived models among the 38 species-phenophases.
 In three cases where a species phenophase combination occurred in two LTER sites (Budburst for \textit{Acer saccharum}, \textit{Betula alleghaniensis}, and \textit{Fagus grandifolia} in the Harvard and Hubbard Brook datasets) they were compared separately to the NPN data.
@@
 Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and NPN derived models.
-For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and NPN-derived estimates for estimates made at the four lTER sites and across all NPN sites.
+For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and NPN-derived estimates for estimates made at the four LTER sites and across all NPN sites.
 
@@
 
+We performed all analysis using both the R and Python programming languages \citep{rcitation, python}.
+Primary R packages used in the analysis included dplyr \citep{dplyr}, tidyr \citep{tidyr}, ggplot2 \citep{ggplot2}, lubridate \citep{lubridate}, prism \citep{prismR}, raster \citep{rasterR}, and sp \citep{sp1}.
+Primary Python packages included SciPy \citep{scipy}, NumPy \citep{numpy}, Pandas \citep{pandas}, and MPI for Python \citep{mpi4py}.
+Code to fully reproduce this analysis is available on GitHub (https://github.com/sdtaylor/phenology\_dataset\_study) and archived on Zenodo (XXXX)
+
 \section*{Results}
 
-The best matches between parameter estimates based on NPN and LTER data were the Fixed GDD model ($R^2$ = 0.49) and the Linear model ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
-The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters but with matching mean parameter values across the two sets of models (Fig.
+The best matches between parameter estimates based on NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
+The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Fig.
 2).
-The Naive model showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the NPN dataset (Fig.
+The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the NPN dataset (Fig.
 2).
-The large outlier for the Fixed GDD model is \textit{Larrea tridentata}.
-\textit{L.
-tridentata} flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
-While the Fixed GDD and Linear models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between NPN and LTER-derived models (Fig.
-SX).
+The large outlier for the \textit{Fixed GDD model} is \textit{Larrea tridentata}; this species' flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
+While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between NPN and LTER-derived models (Fig.
+S6-S7).
 
@@
 3).
-The Fixed GDD model had the highest correlation between the two models sets at NPN sites ($R^2 = 0.82$), while the GDD, M1, and Uniforc models had the highest correlation at LTER sites ($R^2 = 0.54$).
-Comparing models with spatial corrections to the non-spatial alternatives, the MSB (an extension of the alternating model with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the Alternating model.
-The MSB model improved the $R^2$ from 0.38 to 0.45 at LTER sites, and from -0.2 to -0.13 at NPN sites.
-The M1 model (an extension of the GDD model with a spatial correction based on day length) did not improve the correlation between datasets over the GDD model.
+The \textit{Fixed GDD model} had the highest correlation between the two models sets at NPN sites ($R^2 = 0.82$), while the \textit{GDD}, \textit{M1}, and \textit{Uniforc} models had the highest correlation at LTER sites ($R^2 =$ 0.51, 0.52, and 0.51, respectively).
+Comparing models with spatial corrections to the non-spatial alternatives, the \textit{MSB} (an extension of the \textit{Alternating model} with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the \textit{Alternating model}.
+The \textit{MSB model} improved the $R^2$ from 0.36 to 0.45 at LTER sites, and from -0.23 to -0.15 at NPN sites.
+The \textit{M1 model} (an extension of the \textit{GDD model} with a spatial correction based on day length) did not improve the correlation between datasets over the \textit{GDD model}.
 
@@
 4).
-The Naive and Linear models had the largest differences between the two model sets, while the Fixed GDD model had relatively similar errors when evaluated on both NPN and LTER held-out observations.
-Although the Fixed GDD model had the highest agreement in accuracy between NPN and LTER-derived models, it was not the best performing model overall.
-The GDD and Uniforc models commonly made the best predictions, having the lowest RMSE in 22\% and 40\% of cases among NPN-derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig.
+The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both NPN and LTER held-out observations.
+Although the \textit{Fixed GDD model} had the highest agreement in accuracy between NPN and LTER-derived models, it was not the best performing model overall.
+The \textit{GDD} and \textit{Uniforc} models commonly made the best predictions, having the lowest RMSE in 23\% and 40\% of cases among NPN-derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig.
 S1 \& S2).
@@
 
-Data used to build phenology models typically falls into two categories: intensive long-term data with long time-series at a small number of locations (e.g., LTER data in this study), and large-scale data with less intensive data at hundreds of locations (e.g., NPN data)(Table 3).
-This data scenario--a small amount of intensive data and a large amount of less intensive data--is common in many areas of science and results in the need to understand how to choose between, or combine, data sources \citep{hanks2018}.
-We explored this issue for phenology modeling in realation to making predictions and inferring process from models.
+Data used to build phenology models typically falls into two categories: intensive long-term data with long time-series at a small number of locations (e.g., LTER data in this study), and large-scale data with less intensive sampling at hundreds of locations (e.g., NPN data) (Table 3).
+This data scenario--a small amount of intensive data and a large amount of less intensive data--is common in many areas of science and makes it necessary to understand how to choose between, or combine, data sources \citep{hanks2018}.
+We explored this issue for phenology modeling in relation to making predictions and inferring process from models.
 For inference we found that models based on different data sources resulted in different parameter estimates for all but the simplest models.
-For prediction we found that models fit to different data sources tended to make similar predictions, but that models better predicted out-of-sample data from the data type they were fit to.
+For prediction we found that models fit to different data sources tended to make similar predictions, but that models better predicted out-of-sample data from the data type to which they were fit.
 These results are consistent with other research showing that phenology model performance decreases when transferring single-site models to other locations \citep{garcia-mozo2008, xu2013, basler2016}, and with the call for models that better incorporate spatial variation in phenology requirements \citep{richardson2013, chuine2017}.
@@
 Second, spatial variation in phenology requirements could drive these differences, because NPN data integrates over that spatial variation, while LTER data only estimates the phenological requirements for a specific site.
-In this case NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site.
+In this case, NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site.
 The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}.
 Third, these differences could result from issues with model identifiability.
-Since different parameter values can yield nearly identical estimates of phenological events, fits to the data can result in parameters differing between datasets even when the underlying processes generating the data are the same.
-Information about which of these issues may be causing the differences between datasets can be explored using this analyses, as will be explained below.
+Since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same.
+Information about which of these issues may be causing the differences between datasets can be explored using these analyses, as will be explained below.
 
@@
 
-While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information (beyond the Fixed GDD model) based on their ability to make out-of-sample predictions.
+While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the Fixed GDD model, based on their ability to make out-of-sample predictions.
 Based on the RMSE, the GDD and Uniforc models produce the best out-of-sample predictions for the majority of species and phenophases at both NPN and LTER datasets (Fig.
@@
 However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses.
-Correspondence between parameter estimates (Fig.
+Specifically, correspondence between parameter estimates (Fig.
 2), estimates of phenological events (Fig.
@@
 
-In conclusion, our results suggest that both LTER and NPN data both provide valuable information on plant phenology.
-The primary difference in the datasets is spatial scale, yet other differences exist in a dichotomy such that no single dataset can "have it all" (Table 3).
+In conclusion, our results suggest that both LTER and NPN data provide valuable information on plant phenology.
+Models built using both data sources yield effective predictions for phenological events, but parameter estimates from the two data sources differ and models from each source best predict that data source's phenology events.
+The primary difference in the datasets is spatial scale, but due to tradeoffs in data collection efforts, the larger scale NPN data has shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3).
 These differences can be strengths or potential limitations.
-Low site fidelity can limit the ability of the NPN dataset to measure long term trends, which is arguably the largest strength of LTER data.
-Representation of only local species and communities limits the use of LTER data in species level predictive modeling.
-The high number of observers contributing to the NPN dataset can potentially cause issues without adequate processing \citep{gerst2016}.
-While identifying the most effective data sources for different types and scales of analysis is a useful first step, the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018, melaas2016}.
-Our results suggest that methods that can weight LTS data heavily in nearby regions and use large-scale data to explicitly capture spatial variation in phenological requirements will help improve both our understanding of phenology and our ability to make predictions.
-Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have a large spatial extent as well as a daily temporal grain \citep{richardson2018}.
-More complex data integration methods can use data from many sources to give the best estimate for forecasts \citep{ogle2015}, and continuously update forecasts as new real-time data becomes available \citep{luo2011, dietze2018}.
-These methods, as well as integration of multiple data sources, provide the best opportunity for accurate inference about, and forecasting of, the timing of biological events.
+Observers sampling opportunistically allows the NPN dataset to have a large spatial scale, but also leads to low site fidelity which limits the ability to measure long-term trends at local scales \citep{gerst2016}.
+Tracking long-term trends is the major strength of LTER data, but having a relatively small species pool limits its use in species-level predictive modeling.
+Due to these differences, the best data source for making predictions depends on the scale at which the predictions are being made.
+Identifying the most effective data sources for different types and scales of analysis is a useful first step, but the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018, melaas2016}.
+Our results suggest that methods that can learn from the intensive information available in LTER data in regions where it is available, and simultaneously use large-scale data to capture spatial variation in phenological requirements will help improve our ability to understand and predict phenology.
+Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have both a large spatial extent and high temporal resolution \citep{richardson2018}.
+Data integration provides the potential to use data from many sources to produce the best opportunity for accurate inference about, and forecasting of, the timing of biological events.
 
 \section*{Acknowledgments}
+
+This research was supported by the Gordon and Betty Moore Foundation’s Data-Driven Discovery Initiative through Grant GBMF4563 to E.P.
+White.
+We thank the developers and providers of the data that made this research possible including: the USA National Phenology Network and the many participants who contribute to its Nature’s Notebook program; the H.J.
+Andrews Experimental Forest research program, funded by the National Science Foundation's (NSF) LTER Program (DEB-1440409), US Forest Service Pacific Northwest Research Station, and Oregon State University; the Jornada Basin LTER project (NSF Grant DEB-1235828); the Hubbard Brook Experimental Forest, which is operated and maintained by the USDA Forest Service, Northern Research Station, Newtown Square, PA.; the Harvard Forest LTER; and the PRISM Climate Group at Oregon State University.
 
@@
     Dataset Name & Habitat &  Phenological Event\newline (Num. Species) & Reference \\ \hline
-    Harvard & N.E. Deciduous\newline Forest & Budburst (16)\newline Flowering (15) & \citep{okeefe2000} \\
-    Jornada & Chihuahuan Desert & Flowering (2) & ... \\
-    HJ Andrews \newline Experimental Forest & N.W. Wet Coniferous \newline Forest & Budburst (6)\newline Flowering (6) & ... \\
-    Hubbard Brook & N.E. Deciduous \newline Forest & Budburst (3) & ... \\
+    Harvard Forest & N.E. Deciduous\newline Forest & Budburst (16)\newline Flowering (15) & \citep{okeefe2015} \\
+    Jornada Experimental Range & Chihuahuan Desert & Flowering (2) &  \\
+    H.J. Andrews Experimental Forest & N.W. Wet Coniferous \newline Forest & Budburst (6)\newline Flowering (6) & \citep{schulze2017} \\
+    Hubbard Brook & N.E. Deciduous \newline Forest & Budburst (3) & \citep{bailey2018} \\
     \hline
@@
 
-\textbf{Figure 1}: Locations of National Phenology Network sites used (black points) and Long Term Ecological Research sites (labeled circles).
+\textbf{Figure 1}: Locations of National Phenology Network sites used (black points) and Long Term Ecological Research sites (labeled circles), with greyscale showing elevation.
 
@@
 \textbf{Figure 2}: Comparisons of parameter estimates between NPN and LTS derived models.
-Each point represents a parameter value for a specific species and phenophase, and is the mean value from 250 bootstrapped models.
+Each point represents a parameter value for a specific species and phenophase, and is the mean value from 250 bootstrap iterations.
 The black line is the 1:1 line.
@@
 	\centering
-		\includegraphics[width=1\textwidth]{figure_param_comparison.png}
+		\includegraphics[scale=0.65]{figure_param_comparison.png}
     \caption{.}
@@
 Each point is an estimate for a single held-out observation.
-Colors indicate observations for a single species x phenophase combination.
+Colors indicate observations for a single species and phenophase combination.
 
@@
 Each plot represents the difference between the RMSE for LTS-derived model and the NPN-derived model, meaning that values less than zero indicate more accurate prediction by LTS-derived models and values greater than zero indicate more accurate prediction by NPN-derived models.
-Differences are calculated pairwise for the 39 species/phenophase comparisons.
+Differences are calculated pairwise for the 38 species/phenophase comparisons.
 

### 2018-05-30 manuscript/supplement_methods.tex
final revisions from coauthors
mostly minor edits from ethan,joan,michael, and kristina
added abstract
formated using a better latex template

@@
 \begin{document}
-\begin{center}
-{\huge Paper title goes here \par} \newline
+
+{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par} \newline
 
@@
 Methods describing processing of the 4 LTER datasets
-\end{center}
 
@@
 As in the NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation.
+The years used for each datasets were all years available at the time of analysis.
 
 \textbf{Harvard Forest} \newline
+We used observations from 1990-2014.
 Observations here were recorded as relative percentage of flowering or budburst for individual plants.
-"Yes" observations budburst and flower where set to the DOY when the percentage of each tree had was greater than or equal to 10\%
+We set "Yes" observations budburst and flower to the DOY when the percentage of each tree had was greater than or equal to 10\%
 
 \textbf{H.J. Andrews Experimental Forest} \newline
-"yes" observations for budburst was the first DOY when an individual was marked as "bud break".
-"yes" observations for flowering was was when an indiviudal was first marked as "Flowers open".
-Note that each species has slightly different ordinal codes to make each of these events.
+We used observations from 2009-2015.
+We set "yes" observations for budburst to the first DOY when an individual was marked as "bud break" and "yes" observations for flowering were when an indiviudal was first marked as "Flowers open".
+Note that each species has slightly different ordinal codes to mark each of these events.
 
 \textbf{Jornada Experimental Range} \newline
+We used observations from 1992-2009.
 Observations for this dataset represent, for each zone, the percent of plants for a particular species which is observed within each phenophase.
-"Yes" observations for flowers was the first DOY where the flower phenophase was 10\% or greater.
+We set "Yes" observations for flowers to the first DOY where the flower phenophase was 10\% or greater.
 
 \textbf{Hubbard Brook} \newline
-Observations for this dataset represent for each species the average, among 3 individuals, of an ordinal description of phenophase. \newline
+We used observations from 1989-2015. Observations for this dataset represent for each species the average, among 3 individuals, of an ordinal description of phenophase. \newline
 
@@
 
-"Yes" observations for budburst were the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].
+We set "Yes" observations for budburst to the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].
 

### 2018-05-30 manuscript/supplement_images.tex
final revisions from coauthors
mostly minor edits from ethan,joan,michael, and kristina
added abstract
formated using a better latex template

@@
 \begin{document}
-\centering{\Large Paper title goes here \par}
+{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par}
 
@@
 
-\textbf{\centering{\large Supplementary materials \par}} \newline
-Supplementary images S1 - S4
+\textbf{\large Supplementary materials \par} \newline
+Supplementary images S1 - S7 and Table S1
 
@@
 %%%%%%%%%%%%%%%%%
-\textbf{Figure S1}: RMSE for specific species and phenophases of the NPN dataset. X marks the best performing models for the respective data type.
+\newpage
+
+\textbf{Figure S1}: RMSE for specific species and phenophases of the NPN dataset. X marks the best performing models for each respective data type.
 
@@
 		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
+	\caption{Figure S1}
 \end{center}
@@
 %%%%%%%%%%%%%%%%%
-\textbf{Figure S2}: RMSE for specific species and phenophases of the LTER datasets. X marks the best performing models for the respective data type.
+\textbf{Figure S2}: RMSE for specific species and phenophases of the LTER datasets. X marks the best performing models for each respective data type.
 
@@
 
-\textbf{Figure S3}: RMSE of all species and phenophases of the four scenarios described in the text. These values are calculated using held out test data.
+\textbf{Figure S3}: RMSE of all species and phenophases of the four scenarios described in the text. These values were calculated using held out test data.
 
@@
 
-\textbf{Figure S4}: Distribution of model parameters for the three species common to the Hubbard Brook, Harvard, and NPN datasets.
+\textbf{Figure S4}: Distribution of parameters of the Naive, GDD, Fixed GDD, and Linear models for the three species common to the Hubbard Brook, Harvard, and NPN datasets.
 The phenophase is budburst for all three species.
 Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution.
+Note the heading for each sub figure.
 
@@
 	\centering
-		\includegraphics[width=1\textwidth]{supplement_hubbard_harvard_comparison.png}
+		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison1.png}
 	\caption{Figure S4}
@@
 
+%%%%%%%%%%%%%%%%%
+%% Figure S5
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S5}: As in Figure S4, but for the Alternating and Uniforc models.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison2.png}
+	\caption{Figure S5}
+\end{center}
+
+%%%%%%%%%%%%%%%%%
+%% Figure S6
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S6}: As in Figure S4, but for 4 selected species to show the difference in parameter distributions between LTER and NPN derived models.
+The phenophase for the four species is budburst.
+These 4 species are representative of the analysis, and for the remaining comparisons the reader is pointed to the script 'analysis/plot\_select\_species\_parameters.R' in the code repository to generate additional figures.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_select_species_param_comparison1.png}
+	\caption{Figure S6}
+\end{center}
+
+%%%%%%%%%%%%%%%%%
+%% Figure S7
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S7}: As in Figure S6, but for the Alternating and Uniforc models.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_select_species_param_comparison2.png}
+	\caption{Figure S7}
+\end{center}
+%%%%%%%%%%%%%%%%%
+%% Table S1
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\begin{table}
+\footnotesize
+\begin{tabular}{l|r|l|l|l|l|l}
+\hline
+Species & PhenophaseID & Desc. & Harvard & H.J. Andrews & Hubbard & Jornada\\
+\hline
+acer circinatum & 371 & Budburst &  & x &  & \\
+\hline
+acer circinatum & 501 & Flowers &  & x &  & \\
+\hline
+acer pensylvanicum & 371 & Budburst & x &  &  & \\
+\hline
+acer rubrum & 371 & Budburst & x &  &  & \\
+\hline
+acer rubrum & 501 & Flowers & x &  &  & \\
+\hline
+acer saccharum & 371 & Budburst & x &  & x & \\
+\hline
+betula alleghaniensis & 371 & Budburst & x &  & x & \\
+\hline
+betula alleghaniensis & 501 & Flowers & x &  &  & \\
+\hline
+betula lenta & 371 & Budburst & x &  &  & \\
+\hline
+betula papyrifera & 371 & Budburst & x &  &  & \\
+\hline
+betula papyrifera & 501 & Flowers & x &  &  & \\
+\hline
+fagus grandifolia & 371 & Budburst & x &  & x & \\
+\hline
+fraxinus americana & 371 & Budburst & x &  &  & \\
+\hline
+fraxinus americana & 501 & Flowers & x &  &  & \\
+\hline
+ilex verticillata & 371 & Budburst & x &  &  & \\
+\hline
+larrea tridentata & 501 & Flowers &  &  &  & x\\
+\hline
+nyssa sylvatica & 371 & Budburst & x &  &  & \\
+\hline
+pinus strobus & 496 & Budburst & x &  &  & \\
+\hline
+populus tremuloides & 371 & Budburst & x &  &  & \\
+\hline
+populus tremuloides & 501 & Flowers & x &  &  & \\
+\hline
+prosopis glandulosa & 501 & Flowers &  &  &  & x\\
+\hline
+prunus serotina & 371 & Budburst & x &  &  & \\
+\hline
+pseudotsuga menziesii & 480 & Budburst &  & x &  & \\
+\hline
+quercus alba & 371 & Budburst & x &  &  & \\
+\hline
+quercus rubra & 371 & Budburst & x &  &  & \\
+\hline
+quercus rubra & 501 & Flowers & x &  &  & \\
+\hline
+quercus velutina & 371 & Budburst & x &  &  & \\
+\hline
+rhododendron macrophyllum & 371 & Budburst &  & x &  & \\
+\hline
+rhododendron macrophyllum & 501 & Flowers &  & x &  & \\
+\hline
+trillium ovatum & 488 & Budburst &  & x &  & \\
+\hline
+trillium ovatum & 501 & Flowers &  & x &  & \\
+\hline
+vaccinium corymbosum & 371 & Budburst & x &  &  & \\
+\hline
+vaccinium corymbosum & 501 & Flowers & x &  &  & \\
+\hline
+vaccinium parvifolium & 371 & Budburst &  & x &  & \\
+\hline
+vaccinium parvifolium & 501 & Flowers &  & x &  & \\
+\hline
+\end{tabular}
+\caption{Species used in the analysis.}
+\end{table}
+
 \end{document}}

### 2018-05-31 analysis_and_figures.sh
some final cleanup
add parameter and predictions results
add README
add NPN data (but only on zenodo since it's so big)
update run all analysis script

@@ -3,3 +3,5 @@
 Rscript analysis/compare_parameters.R
 Rscript analysis/observation_estimate_comparison.R
 Rscript analysis/observation_error_analysis.R
+Rscript analysis/compare_hubbard_harvard.R
+Rscript analysis/plot_select_species_parameters.R

### 2018-05-31 data_preprocessing/extract_prism_temps.R

@@ -44,7 +44,7 @@ options(prism.path = "~/data/prism_daily")
 #get_prism_dailys(type = 'tmean', minDate = '1981-01-01', maxDate = '2005-12-31', keepZip = FALSE)
 
 #NPN Site coordinates
-site_info = read_csv('~/data/phenology/npn_core/ancillary_site_data.csv') %>%
+site_info = read_csv('raw_data/npn/ancillary_site_data.csv') %>%
   dplyr::select(Site_ID, Latitude, Longitude) %>%
   dplyr::distinct() %>%
   mutate(dataset='npn', start_year=2009, end_year=2016)

### 2018-05-31 data_preprocessing/compile_npn_data.R
some final cleanup
add parameter and predictions results
add README
add NPN data (but only on zenodo since it's so big)
update run all analysis script

@@ -17,7 +17,7 @@ subset_npn_data = function(this_species, this_phenophase){
 }
 
 ##################################################################################
-data_dir = '~/data/phenology/npn_core/'
+data_dir = 'raw_data/npn/'
 
 
 #Some species are in > dataset, so only get distinct species/phenophases

### 2018-05-31 data_preprocessing/compile_rmbl.R

@@ -1,109 +1,0 @@
-library(tidyverse)
-source('./data_preprocessing/processing_utils.R')
-
-rmbl = readxl::read_excel('./raw_data/rmbl/phenology data_1973_2012 from Amy Iler.xlsx', sheet='All DATA') %>%
-  select(-HABITAT, plot=PLOT, species=SPECIES, year=YEAR, flower_count=FLOWER_COUNT, doy=DOY) %>%
-  filter(year>=1974) %>%
-  mutate(species = tolower(species))
-
-
-#Counting flowers means that observations of 0 flowers go unrecorded. Here I infer those
-#by using all obsevations from each plot in each year. The plots are 2mx2m so it seems safe to assume this. 
-#The first observation at each site in each year
-all_observations = rmbl %>%
-  select(plot,year,doy) %>%
-  distinct()
-
-all_plot_species_year_observations = rmbl %>%
-  select(plot,year,species) %>%
-  distinct() %>%
-  right_join(all_observations, by=c('plot','year'))
-
-#Threshold for first flowering is 10% of the maximum count of flowers for each site,species,year
-#Same as Iler et al. 2013
-max_yearly_count = rmbl %>%
-  group_by(plot,species,year) %>%
-  filter(flower_count == max(flower_count)) %>%
-  filter(doy==min(doy)) %>%
-  ungroup() %>%
-  mutate(first_flower_threshold = flower_count*0.1) %>%
-  select(-flower_count, -doy)
-
-
-#takes a vector of whether a count is past the first flowering threshold
-#and return the first day which this is true
-get_past_threshold_doy = function(doy, thresholds){
-  position = which.max(thresholds)
-  if(position < 1){
-    return(NA)
-  } else {
-    return(doy[position])
-  }
-}
-#takes a vector of whether a count is past the first flowering threshold
-#and return the first day which this is true
-get_prior_past_threshold_doy = function(doy, thresholds){
-  position = which.max(thresholds)-1
-  if(position < 1){
-    return(NA)
-  } else {
-    return(doy[position])
-  }
-}
-
-
-#First flowering date is the  midpoint between the 2 dats it passes the threshold
-#again from Iler et al. 2013.
-#This dataset is a organized differently than the others because it has flower counts,
-#so I'm not using the common code in processing_utils.R
-first_flowering = rmbl %>%
-  right_join(all_plot_species_year_observations, by=c('species','plot','year','doy')) %>%
[…]

### 2018-05-31 data_preprocessing/compile_zumwalt_data.R

@@ -1,26 +1,0 @@
-library(tidyverse)
-library(lubridate)
-
-data_dir = '~/data/phenology/zumwalt/'
-
-observations = read_csv(paste0(data_dir,'status_intensity_observation_data.csv')) 
-
-
-obs_dates = observations %>% 
-  group_by(date) %>% 
-  summarise(num_species = n_distinct(tree.id)) %>% 
-  mutate(year=lubridate::year(date)) %>%
-  arrange(date)
-
-obs_dates$days_since_last = NA
-for(i in 2:nrow(obs_dates)){
-  obs_dates$days_since_last[i] = obs_dates$date[i] - obs_dates$date[i-1]
-}
-
-
-%>%
-  filter(Phenophase_ID == 371, Phenophase_Status>=0) %>%
-  select(date=Observation_Date, Site_ID, species, Phenophase_Status, Individual_ID) %>%
-  mutate(year = year(date), doy=yday(date)) %>%
-  filter(Site_ID %in% sites_with_env_data$Site_ID)
-

### 2018-05-31 compile_npn_species_sampling_data.R
remove old unused script

@@ -1,90 +1,0 @@
-library(tidyverse)
-library(sp)
-library(geosphere)
-
-#############################################################
-#Get the number of NPN observers for each species observation
-data_dir = '~/data/phenology/npn_core/'
-
-#The raw npn data
-all_observations = read_csv(paste0(data_dir,'status_intensity_observation_data.csv')) %>%
-  select(Site_ID, individual_id = Individual_ID, observer_id = ObservedBy_Person_ID, Phenophase_ID, 
-         Observation_Date, status = Phenophase_Status, intensity_id = Intensity_Category_ID,
-         intensity = Intensity_Value, Genus, Species) %>%
-  mutate(species= tolower(paste(Genus,Species,sep=' ')), 
-         year   = lubridate::year(Observation_Date),
-         doy    = lubridate::yday(Observation_Date))
-
-#########################################################################
-unique_observers_for_each_species = all_observations %>%
-  group_by(species) %>%
-  summarize(num_observers = n_distinct(observer_id)) %>%
-  ungroup()
-
-#########################################################################
-#Get the total area observed for each species (~range size, but not quite)
-
-#Get the area of a species distrubtion by drawing a convex hull over points
-#75% sure this is calculating as expected. 
-get_area=function(lat,lon){
-  if(length(lat)<3){return(NA)}
-  coords = matrix(c(lon,lat), ncol=2)
-  hull=grDevices::chull(coords)
-  hull=c(hull, hull[1])
-  hull_coord=coords[hull,]
-  
-  hull_polygon=SpatialPolygons(list(Polygons(list(Polygon(hull_coord)), ID=1)))
-  
-  #convert area to square km
-  return(geosphere::areaPolygon(hull_polygon)/(1000^2))
-}
-
-site_info = read_csv(paste0(data_dir,'ancillary_site_data.csv')) %>%
-  select(lon=Longitude, lat=Latitude, Site_ID)
-
-unique_sites_for_each_species = all_observations %>%
-  select(species, Site_ID) %>%
-  distinct() %>%
-  left_join(site_info, by='Site_ID')
-
-area_sampled_for_each_species = unique_sites_for_each_species %>%
-  group_by(species) %>%
-  do(area_sampled = get_area(lat=.$lat, lon=.$lon)) %>%
-  ungroup()
-area_sampled_for_each_species$area_sampled = unlist(area_sampled_for_each_species$area_sampled)
-###################################################################
[…]

### 2018-05-31 manuscript/main.tex
final submitted version of manuscript
including images and rendered PDF's

@@
 This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability.
-Here we explore this issue with phenology by comparing models built using data from the large-scale, citizen science National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites.
+Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites.
 We built process based phenology models for species common to each dataset.
 From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both NPN and LTER data.
-We found that model parameter estimates for the same species are most similar between the two datasets when using simple models, but parameter estimates vary widely as model complexity increases.
-Despite this, estimates for the date of phenological events and out-of-sample errors are similar, regardless of the model chosen.
+We found that model parameter estimates for the same species were most similar between the two datasets when using simple models, but parameter estimates varied widely as model complexity increased.
+Despite this, estimates for the date of phenological events and out-of-sample errors were similar, regardless of the model chosen.
 Predictions for NPN data had the lowest error when using models built from the NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built.
@@
 For these reasons, data from a single location is not adequate for larger scale phenology modeling.
-Accurately forecasting phenology at larger scales will require models that account for the full range of variation across a species' range \citep{richardson2013, chuine2017}, which will necessitate the use of data sources beyond traditional single-site studies.
+Accurately forecasting phenology at larger scales will require models that account for the full range of variation across a species' range \citep{richardson2013, tang2016, chuine2017}, which will necessitate the use of data sources beyond traditional single-site studies.
 
@@
 
-The National Phenology Network protocol uses status-based monitoring, where via an phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}.
+The National Phenology Network protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}.
 Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves.
@@
 As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophases combinations which had at least 30 total observations.
-After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table S1).
+After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table 1 \& S1).
 
@@
 For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$.
-The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0C).
+The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0$^{\circ}$C).
 The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan.
@@
 Models with different parameter values, and even entirely different structures, can produce similar estimates for the date of phenological events \citep{basler2016}.
-Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and NPN derived models.
+Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and NPN derived models for all held out observations.
 For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and NPN-derived estimates for estimates made at the four LTER sites and across all NPN sites.
@@
 The second metric focuses on large-scale prediction by comparing the fits of LTER and NPN-derived models on the NPN data: $RMSE_{C} - RMSE_{D}$.
+These metrics were calculated for each of the model types and 38 species-phenophase combinations.
 Negative values indicate that LTER-derived models perform better, while positive values indicate that the NPN-derived model performed better.
@@
 Primary Python packages included SciPy \citep{scipy}, NumPy \citep{numpy}, Pandas \citep{pandas}, and MPI for Python \citep{mpi4py}.
-Code to fully reproduce this analysis is available on GitHub (https://github.com/sdtaylor/phenology\_dataset\_study) and archived on Zenodo (XXXX)
+Code to fully reproduce this analysis is available on GitHub  (https://github.com/sdtaylor/phenology\_dataset\_study) and archived on Zenodo  (https://doi.org/10.5281/zenodo.1256705)
 
@@
 The \textit{MSB model} improved the $R^2$ from 0.36 to 0.45 at LTER sites, and from -0.23 to -0.15 at NPN sites.
-The \textit{M1 model} (an extension of the \textit{GDD model} with a spatial correction based on day length) did not improve the correlation between datasets over the \textit{GDD model}.
+The \textit{M1 model} (an extension of the \textit{GDD model} with a spatial correction based on day length) improved the correlation over the \textit{GDD model} only slightly at LTER sites (from 0.51 to 0.52) and did not improve the correlation at NPN sites.
 
@@
 This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues.
-For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 produce nearly identical results in many scenarios.
+For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T^*$=0 and $t_{1}$=5, $F$=5, $T^*$=0 produce nearly identical estimates in many scenarios.
 This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters.
-In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTER data (Supplementary Material).
-Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Supplementary Material).
+In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTER data (Fig.
+S6-S7).
+Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig.
+S4-S5).
 These results are consistent with research showing that models estimating the starting day of warming accumulation from budbreak time-series failed to accurately infer the internal phenology described in the models \citep{chuine2016}.
 \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting.
-Here, our simplest model, the Fixed GDD model which uses only a warming component, had the highest correlation among parameters between LTER and NPN datasets.
+Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and NPN datasets.
 In combination with this previous research, our results warrant caution in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.
 
-While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the Fixed GDD model, based on their ability to make out-of-sample predictions.
-Based on the RMSE, the GDD and Uniforc models produce the best out-of-sample predictions for the majority of species and phenophases at both NPN and LTER datasets (Fig.
+While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the \textit{Fixed GDD model}, based on their ability to make out-of-sample predictions.
+Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both NPN and LTER datasets (Fig.
 S1 \& S2).
@@
 Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}.
-In combination with our results showing observed differences in parameter estimates between LTS sites (Supplementary Material), this suggests that variation in phenological requirements across the the range is likely important.
+In combination with our results showing observed differences in parameter estimates between LTER sites (Fig.
+S4-S5), this suggests that variation in phenological requirements across the the range is likely important.
 However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses.
@@
 3), and out-of-sample error rates (Fig.
-4) for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively.
+4) for the \texit{MSB} and \textit{M1} models were essentially the same as the \textit{Alternating} and \textit{GDD} models, respectively.
 This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors.
-\cite{basler2016} used the M1 model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day.
+\cite{basler2016} used the \textit{M1} model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day.
 Their result was strengthened by having a 40-year time-series across a large region.
@@
 Models built using both data sources yield effective predictions for phenological events, but parameter estimates from the two data sources differ and models from each source best predict that data source's phenology events.
-The primary difference in the datasets is spatial scale, but due to tradeoffs in data collection efforts, the larger scale NPN data has shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3).
+The primary difference in the datasets is spatial scale, but due to trade-offs in data collection efforts, the larger scale NPN data has shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3).
 These differences can be strengths or potential limitations.
@@
     \hline
-    Dataset Name & Habitat &  Phenological Event\newline (Num. Species) & Reference \\ \hline
-    Harvard Forest & N.E. Deciduous\newline Forest & Budburst (16)\newline Flowering (15) & \citep{okeefe2015} \\
+    Dataset Name & Habitat &  Phenological Event (Num. Species) & Reference \\ \hline
+    Harvard Forest & N.E. Deciduous Forest & Budburst (17) Flowering (7) & \citep{okeefe2015} \\
     Jornada Experimental Range & Chihuahuan Desert & Flowering (2) &  \\
-    H.J. Andrews Experimental Forest & N.W. Wet Coniferous \newline Forest & Budburst (6)\newline Flowering (6) & \citep{schulze2017} \\
-    Hubbard Brook & N.E. Deciduous \newline Forest & Budburst (3) & \citep{bailey2018} \\
+    H.J. Andrews Experimental Forest & N.W. Wet Coniferous Forest & Budburst (5) Flowering (4) & \citep{schulze2017} \\
+    Hubbard Brook & N.E. Deciduous Forest & Budburst (3) & \citep{bailey2018} \\
     \hline
@@
 
-\textbf{Figure 2}: Comparisons of parameter estimates between NPN and LTS derived models.
+\textbf{Figure 2}: Comparisons of parameter estimates between NPN and LTER derived models.
 Each point represents a parameter value for a specific species and phenophase, and is the mean value from 250 bootstrap iterations.
@@
 
-\textbf{Figure 3}: Comparison of predicted day of year (DOY) of all phenological events between NPN and LTS-derived models.
-Top panels show comparisons at LTS sites and bottom panels show comparisons at NPN sites.
+\textbf{Figure 3}: Comparison of predicted day of year (DOY) of all phenological events between NPN and LTER-derived models.
+Top panels show comparisons at LTER sites and bottom panels show comparisons at NPN sites.
 Each point is an estimate for a single held-out observation.
@@
 
-\textbf{Figure 4}: Differences in prediction error between NPN and LTS-derived models.
-Density plots for comparisons of predictions on LTS data (top row) and NPN data (bottom row).
-Each plot represents the difference between the RMSE for LTS-derived model and the NPN-derived model, meaning that values less than zero indicate more accurate prediction by LTS-derived models and values greater than zero indicate more accurate prediction by NPN-derived models.
+\textbf{Figure 4}: Differences in prediction error between NPN and LTER-derived models.
+Density plots for comparisons of predictions on LTER data (top row) and NPN data (bottom row).
+Each plot represents the difference between the RMSE for LTER-derived model and the NPN-derived model, meaning that values less than zero indicate more accurate prediction by LTER-derived models and values greater than zero indicate more accurate prediction by NPN-derived models.
 Differences are calculated pairwise for the 38 species/phenophase comparisons.

### 2018-05-31 data_preprocessing/compile_npn_data.R
double space manuscript

@@ -27,7 +27,7 @@ non_npn_species = read_csv('./cleaned_data/non_npn_species_list.csv') %>%
 
 
 #The raw npn data
-all_observations = read_csv(paste0(data_dir,'status_intensity_observation_data.csv')) %>%
+all_observations = read_csv(paste0(data_dir,'status_intensity_observation_data.csv.gz')) %>%
   select(Site_ID, individual_id = Individual_ID, Phenophase_ID, Observation_Date, status = Phenophase_Status,
          intensity_id = Intensity_Category_ID, intensity = Intensity_Value, Genus, Species) %>%
   mutate(species= tolower(paste(Genus,Species,sep=' ')), 

### 2018-08-14 manuscript/Reviewer_Reply.md
progress as of today on revision
includes notes from meeting with Ethan

#  Comments from the SME:

**General comment: I appreciate the authors' efforts to ensure transparency and repeatability of the analysis.**

Thank you.

**l. 20: not all the models are process-based; revise?**

We added changed this to "statistical and processed based models".

**l. 41: could you delete "uncertainty in", or replace with "estimates of"? (Isn't it the timing itself that's important, not the uncertainty in the timing?)**

We deleted "uncertainty in".

**l. 81: fix "quanity".**

Fixed.

**l. 115: why these dates?**

These dates were used in accordance with Crimmins (2017), who used them prior, as budburst is unexpected to be first seen past DOY 172 and DOY 213 for flowers.

**l. 127: phenophase (singular)**

Fixed.

**ll. 210-211. I don't understand this. Why do some seemingly strong relationships (e.g., "naive" panel of Fig. 2; also some panels of Fig. 3) yield negative R-squared values? This probably just reflects my ignorance... but it's likely that if I'm mystified, other readers will be, too! Please explain more clearly how the <0 R-squareds arise.**

**I found the discussion quite insightful, although there are some additional points that should be addressed and a few areas where clarity could be improved (see Reviewer 1's comments)**.

Thank you. Please see the replies to Reviewer 1.

**Also:**
**l. 289: what do you mean by "internal phenology"? internal physiological processes?**

Specifically we meant the date of endodormancy break, but had not used this term prior and did not want to confuse readers unfamiliar with it. We have changed this text to be more descriptive of the Chuine et al. 2016 study results.

**l. 294. "warrant caution" -> "indicate that caution is warranted"**

Fixed.

**l. 305: "the" is repeated.**

Fixed.

**Fig. 4: the labels on the right-hand side ("Metric 1/2") are uninformative. Please use more intuitive labels, e.g. "LTER data/NPN data".**

We replaced the labels to reflect the data evaluated. We also adjusted the methods section slightly to hopefully make this more clear.

# Reviewer(s)' Comments to Author:

## Reviewer: 1

### Comments to the Author
**Comparison of large-scale citizen science data and long-term study data for phenology modeling**

**This is a very practical study and the evaluation of the approach is clear and pretty straightforward to follow. The authors might improve readability a bit by either adding subheads into the Methods & Results or repeatedly using the same terms to refer to the various steps (eg, cross-validation).**

After some discussion, we decided not to add subheadings.

**The most critical thing that I see that’s lacking is that there is no consideration of how decisions made regarding which NPN data to include in the analysis might be impacting model performance. The authors allowed observations with up to 30d between the last reported “no” and the first reported “yes” and chose the midpoint. This could still lead to up to 15d error in the assumed date of “yes” and the true date of “yes”; this variance may in many cases be larger than the signal in the model. Though actually addressing this sensitivity is probably beyond the scope of the analysis, the potential and implications for this error should be discussed explicitly in the paper.**

Gerst et al. 2016 looked at this issue and found that differences in site level means among more restrictive datasets was normally distributed. Thus having a shorter cutoff period here would result equally in less overestimates *and* underestimates. Since the models here are optimzed for RMSE we feel the results would not be affected overall by having a cutoff shorter than 30 days.

Shawn's note:
After talking with Ethan I'll redo most of the analysis with a shorter cutoff date and see how things compare.
also report the distribution of "days between" to see if they're very far off between 30day cutoff and a shorter cutoff.

**The authors lay out two goals for this effort – to determine “inferences about biological processes driving phenology” and to develop predictive phenology models. It may be just me…using the term “inference” in this way confused me.**

**Regarding the predictions that can result from these different models…this is of strong interest to many stakeholder realms. It would be fantastic to see a bit more attention given in the discussion to recommendations for implementation. The authors talk about the tradeoffs between more complex formulations and simpler models and – generally – reasons these differences may come about. It would be great to also see a bit of more practical discussion along the lines of, “if you can tolerate error of 7-14d, then model X may meet your needs, and it’s fine to build the models with either NPN or LTER data. However, if your needs are Y, then you’d be better off doing blah, blah…”… would this be possible?**

note:
This can essentially be done by looking at some of the supplement figures. Will point this out in the discussion.
shoot for 2-3 sentences describing some rules of thumb. with specific of error rates someone can be ok with.

**Thank you for your serious efforts to understand the USA-NPN data and to contribute to the collective understanding of this dataset’s potential as well as shortcomings. I’m so excited to see this sort of work happen! I’m happy to talk with you more regarding this manuscript as well as your other efforts, especially if any of these comments require clarification.**
Sincerely,
Theresa Crimmins, theresa@usanpn.org

### Smaller things

**Please refer to the Network as “USA National Phenology Network” and “USA-NPN” because there are lots of other phenology networks in other countries**

We have added this clarification in several places within the text to avoid any confusion.

**In several places “data” is treated as singular… data are plural**

Fixed.

**L67: the citizen science program run by the USA-NPN that yields the data housed in the National Phenology Database is Nature’s Notebook**
**Thank you for acknowledging the Nature’s Notebook contributors for the data!**

You're welcome. We're happy to credit all data providers, as this study would not have been possible without them.

**Please see the USA-NPN data attribution policy for recommendations for how to properly cite the dataset: https://www.usanpn.org/terms#DataAttribution**

### Abstract
**be sure to indicate that the study is focused on the U.S.**

This is now made clear by specifying that it's the USA NPN.

**L26-29: models performed best when applied to the same data with which they were built – you attribute this to scale…though other things could be at play here, such as local adaptation and species’ sensitivity to different forcing variables varying across the range.**

**L33: you mention that the NPN dataset offers many species… though this is a good point, it is actually irrelevant to the arguments that you are making in this paragraph, that is, if your comparison is focused on only a handful of species.  Oh, now I see you make this point several times in the paper…but I’m still struggling to see how it’s relevant. Is it that you could have built models with many more species in this study based on what NPN offers, but you were limited by what the LTER datasets had to offer?**

Reviewer 2 is correct in that our analysis was limited by species common to the NPN and four LTER datasets. Large scale phenology studies typically employ as many species as are available in a given dataset. Thus we point out the large species pool of the NPN dataset throughout as it is something other researchers should consider when designing their own analysis.

**L34-36: this concluding sentence could be made stronger if you kept the focus on the findings of the present paper – the strengths and limitations of the two datasets and their suitable applications.**

We felt that the strengths and limits of these datasets required more attention in the abstract than could be managed in the final sentence, so we discussed them in the prior sentences primarily and instead looked forward in the final sentence.

### Introduction

**L77: Nature’s Notebook has 1,000s of volunteer participants (www.usanpn.org/data/dashboard)**

Fixed.

**L80: may also want to check out Feldman et al. (2018) – https://doi.org/10.1007/s00484-018-1540-4 for further support on cit sci observer skill**

### Methods

**I think it’s pretty critical to provide your sample sizes for each of the species/phenophases under evaluation for each of the datasets… when I read that you held out 20% of observations for evaluation, I was left wondering, was this 5 points or 50 or 500? You could include this info in Table 1**

We have added details of the sample sizes to the supplement table S1.

**What is the sampling frequency for the LTER sites? This is quite relevant because you chose to use 30d as your cut-off for NPN data - you’ll allow for up to 30d between the last reported “no” and the first reported “yes”. How does this compare to what’s possible in the LTER datasets? It could definitely have an impact on the sensitivity of your results.**

Shawns notes:
put the numbers below in the table or ref. and re-reference them when i talk about reviewer 1's other issue with NPN sampling frequency

sampling freq. from metadata / from data
harvard  3-7 days / 5.6 days
hubbard  7 days / 7.1 days
hjandrews  7 days / 7.6 days
jornada  monthly / 30.5 days

**After reading about the phenology datasets, I expected to read about the temperature data used. I see now that it’s nested within the “Modeling” section… the organization just surprised/confused me.**

The phenological data are the crux of the study so we felt that they deserved special consideration, while the components used to generate the models could be kept within the modeling section. We will gladly change this if the editor feels it is necessary.

**L138-140: Provide some references for your GDD calculation – there are multiple ways to go about calculating GDDs**

The exact calculations are unique to each model and are provided along with references in Table 2. We have clarified this in the text. We also added references for the GDD and Fixed GDD model.

**L143: what is NCD?**

This is the number of chill days (mean daily temperature < 0 degrees C) since Jan. 1 in a given year. We have added the description to the text.

**L145: please provide some references to back up your statement that these formulations are the most common**

**L156: I’m sure you extracted the daily temp at the lat/long of each observation location – maybe just make that a bit clearer**

We have clarified this.

**I don’t think I saw “RMSEA – RMSEB” and “RMSEC-RMSED” referenced anywhere else in the paper**

We removed these specific equations and replaced them with a better description of this analysis.

**L196: what does “scenario A” refer to here?**

This referred to the RMSE values derived from LTER Model predictions for held out LTER data. We rewrote this paragraph slightly and changed the labels on Figure 4 to make this analysis more clear.

### Results

**L231-237: is it possible to quantify these comparisons in any way?**

Suggestion from Ethan:
do a paired t-test here to show distribution is different from 0. Make sure text still matches if a bunch are or are not significantly different than 0.

### Discussion

**L244: are you referring to the temporal depth and/or frequency of sampling when you say “intense”?**

We are referring to both of these as LTER sites typically have a long history of frequent and consistent sampling compared to USA-NPN sites.

**263: “limitations in the sampling of the NPN dataset” – what sort of limitations? Small sample size, frequency of sampling, or something else?**

The frequency of sampling, in a given season and also between seasons, is the primary limit we consider here and we've made this clearer by stating as much.

**L266-268: it would be helpful to add some references for local adaptation and differing sensitivity to phenological drivers across gradients**

We added two references here to highlight the potential variation in phenological requirements across different gradients.

**L270-272: another option could be to develop local or regional models, especially if a national-scale prediction is not needed**

We have briefly included this possibility, as there may indeed be circumstances (population conservation) where the national prediction is unneeded.

**L273: define “model identifiability” and/or provide a reference? I think you are explaining it in the subsequent sentence – if so, just make this more explicit.**

We have made this clearer by switching to a colon instead of a period between the two sentences.

**L276: what do you mean by “these analyses”?**

Maybe change to "our analyses" ?

**L293: not sure what you’re referring to when you say “this previous research”**

We were referencing the two studies cited in this paragraph (Chuine et al. 2016 and Basler 2016). We have changed this to "In combination with the aforementioned studies,..."

**L337: there are many recent studies that have brought together these various datsets; would be worth adding a couple of references**

Shawn's notes on some refs to add:
comparing phenocam + modis: https://doi.org/10.1038/s41598-018-23804-6
comparing modis phenology w/ NPN & Flux towers: 10.1016/j.ecolind.2017.02.024

## Reviewer: 2

### Comments to the Author
Summary:

**While I do think that this manuscript should be ultimately acceptable for publication at Ecology, there are a few major/minor comments and suggestions for which I would like the authors to consider.**

Major Comments and Suggestions:

**First, for NPN species with larger pools/spatial extents of observations, I wonder if would be useful to calibrate the models using subsets somehow stratified by climate and/or latitude. For example, train the models using only warm region (or southern) red maple and test on cold region maples (or vice versa). Perhaps this could further address the issue of model identifiability (as discussed in Lines  259-277). While the oft-mentioned Basler (2016) study certainly had nice spatiotemporal coverage, it did not have nearly the latitudinal/climate variability that NPN affords.**

We discussed this and ultimately decided that it is beyond the scope of the current paper. Although we agree that this approach has the potential to improve the paper, we are unsure of the extent to which this would address the issue of identifiability, which is itself noted as a potential issue rather than a definite one. We also feel that it would require too much space to adequately investigate and discuss, and that the current set of analyses sufficiently accomplish the goals we set out in the introduction. However, we do hope to pursue these confounding factors further in future efforts.

**Second, given the length of time series of the LTER datasets, I think it would be useful to further examine the performance of models based on other criteria such as the amount of interannual variability explained (correlation or R-square) or bias during anomalously warm or cold years. Perhaps the latter might be outside of the overall realm of the paper, but I think it would be informative to know which models capture this type of information which matters in a global change context.**

Shawn's note:
make a R^2 figure for all species/phenophase for the supplement
on the stratified sampling. We thought of this, there's not enough data even for Acer rubrum. Write a sentence or 2 about this in the discussion and how it could be done someday when enought data is collected.

### Minor Comments and Suggestions:

**Line 281: Please verify what is T* (I assume base temperature)**

### 2018-08-14 manuscript/main.tex
double space manuscript
progress as of today on revision
includes notes from meeting with Ethan

@@
 % This is the double space command
-\renewcommand{\baselinestretch}{1.6}
+\renewcommand{\baselinestretch}{2.0}
 
@@
 This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability.
-Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites.
-We built process based phenology models for species common to each dataset.
+Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science USA National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites.
+We built statistical and process based phenology models for species common to each dataset.
 From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both NPN and LTER data.
@@
 Plant phenology, the timing of recurring biological events such as flowering, plays an important role in ecological research extending from local to global scales \citep{cleland2007, richardson2013, tang2016}.
-At large scales, uncertainty in the timing of spring leaf out and fall senescence influence the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long-term climate forecasts \citep{richardson2012}.
+At large scales the timing of spring leaf out and fall senescence influence the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long-term climate forecasts \citep{richardson2012}.
 At smaller scales, species-specific responses to temperature and precipitation can alter flower communities \citep{diez2012, caradonna2014, theobald2017} and affect the abundance and richness of both pollinators \citep{ogilvie2017a, ogilvie2017b} and organisms at higher trophic levels \citep{tylianakis2008}.
@@
 This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, or differences in plant age or population density \citep{kramer1995, diez2012}.
-For these reasons, data from a single location is not adequate for larger scale phenology modeling.
+For these reasons, data from a single location are not adequate for larger scale phenology modeling.
 Accurately forecasting phenology at larger scales will require models that account for the full range of variation across a species' range \citep{richardson2013, tang2016, chuine2017}, which will necessitate the use of data sources beyond traditional single-site studies.
 
-Data from citizen science projects is becoming increasingly important for ecological research \citep{kelling2009, dickinson2010, tulloch2013}.
-Because this data is often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams.
-A relatively new citizen science project started in 2009, The National Phenology Network (NPN), collects phenology observations from volunteers throughout North America and makes the data openly available \citep{npncitation}.
-Data from this project has already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}.
+Data from citizen science projects are becoming increasingly important for ecological research \citep{kelling2009, dickinson2010, tulloch2013}.
+Because these data are often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams.
+A relatively new citizen science project started in 2009, The USA National Phenology Network (NPN), collects phenology observations from volunteers throughout North America and makes the data openly available \citep{npncitation}.
+Data from this project have already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}.
 Large-scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the NPN dataset has the potential to meet these needs for North American plant species and communities.
 However, the features that allow citizen science projects to collect data at large scales can also introduce spatial biases toward cities and easily-accessible areas, and variation in sampling effort and observer skill \citep{dickinson2010}.
-With hundreds of participants across North America, the potential for variation among observers in their determination of species identification and dating of phenological events is high.
+With thousands of participants across North America, the potential for variation among observers in their determination of species identification and dating of phenological events is high.
 While volunteers have been shown to be accurate at distinguishing different leaf and flower stages for plants, \citep{fuccillo2015}, observations are sometimes made sporadically across seasons, years, and locations.
-This means that the quanity and quality of data at a specific site will typically be more variable for citizen science efforts than for intensive, long-term studies.
+This means that the quantity and quality of data at a specific site will typically be more variable for citizen science efforts than for intensive, long-term studies.
 
@@
 
-The National Phenology Network protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}.
+The USA National Phenology Network protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}.
 Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves.
@@
 To match the NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods).
-As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophases combinations which had at least 30 total observations.
+As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations.
 After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table 1 \& S1).
@@
 For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$).
-Forcing units are a transformation of the daily mean temperature.
+Forcing units are a transformation of the daily mean temperature and are calculated differently for each model (Table 2).
 The start day can either be estimated or fixed.
-For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$.
+For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$ \citep{reaumur1735, wang1960, hunter1992}.
 The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0$^{\circ}$C).
 The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan.
-1 ($NCD$).
+1 (number of chill days - $NCD$).
 The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}.
@@
 
-We extracted corresponding daily mean temperature for all NPN and LTER observations from the gridded PRISM dataset \citep{prismdata}.
+We extracted corresponding daily mean temperature for all NPN and LTER observations from the gridded PRISM dataset using the latitude and longitude of the site associated with each observation \citep{prismdata}.
 We parameterized all models using differential evolution to minimize the root mean square error (RMSE) of the estimated DOY of the phenological event.
@@
 
-We also directly evaluated model performance using four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) NPN-derived models predicting LTER observations, C) LTER-derived models predicting NPN observations, and D) NPN-derived models predicting NPN observations.
-Within each of these scenarios, we calculated the RMSE of the held-out observations for each species, phenophase, and model type.
-Using the RMSE values, we calculated two different metrics to compare the performance of LTER and NPN-derived models on different data types.
-The first metric focuses on local-scale prediction by comparing the fits of LTER and NPN-derived models on LTER observations: $RMSE_{A} - RMSE_{B}$.
-The second metric focuses on large-scale prediction by comparing the fits of LTER and NPN-derived models on the NPN data: $RMSE_{C} - RMSE_{D}$.
-These metrics were calculated for each of the model types and 38 species-phenophase combinations.
+We also directly evaluated model performance using the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) NPN-derived models predicting LTER observations, C) LTER-derived models predicting NPN observations, and D) NPN-derived models predicting NPN observations.
+Using the RMSE values from held out observations, we  compared the  performance of LTER and NPN-derived models on different data types in two different ways.
+First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and NPN-derived models solely with LTER observations.
+Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely NPN data.
+These differences were calculated for each of the model types and 38 species-phenophase combinations.
 Negative values indicate that LTER-derived models perform better, while positive values indicate that the NPN-derived model performed better.
-In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison (scenario A) within each site, not across sites, to focus on local scale prediction when LTER data is available.
+In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available.
 Absolute RMSE values are provided in the supplement (Fig.
@@
 These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters.
-First, the differences could result from limitations in the sampling of the NPN dataset, leading to less accurate parameter estimates.
-If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the Fixed GDD model is best for making inferences when NPN data is all that is available.
-Second, spatial variation in phenology requirements could drive these differences, because NPN data integrates over that spatial variation, while LTER data only estimates the phenological requirements for a specific site.
+First, the differences could result from limitations in the sampling of the NPN dataset, such as irregular sampling of the same location within or between seasons, leading to less accurate parameter estimates.
+If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the Fixed GDD model is best for making inferences when NPN data are all that is available.
+Second, spatial variation in phenology requirements could drive these differences \citep{diez2012, zhang2017}, because NPN data integrate over that spatial variation, while LTER data only estimate the phenological requirements for a specific site.
 In this case, NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site.
-The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}.
-Third, these differences could result from issues with model identifiability.
-Since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same.
+The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}, although localized models could also be generated when large-scale predictions are unnecessary.
+Third, these differences could result from issues with model identifiability: since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same.
 Information about which of these issues may be causing the differences between datasets can be explored using these analyses, as will be explained below.
@@
 S4-S5).
-These results are consistent with research showing that models estimating the starting day of warming accumulation from budbreak time-series failed to accurately infer the internal phenology described in the models \citep{chuine2016}.
+These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estamates which were not biologically realistic \citep{chuine2016}.
 \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting.
 Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and NPN datasets.
-In combination with this previous research, our results warrant caution in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.
+In combination with the aforementioned studies, our results indicate that caution is warranted in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.
 
@@
 In combination with our results showing observed differences in parameter estimates between LTER sites (Fig.
-S4-S5), this suggests that variation in phenological requirements across the the range is likely important.
+S4-S5), this suggests that variation in phenological requirements across the range is likely important.
 However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses.
@@
 Models built using both data sources yield effective predictions for phenological events, but parameter estimates from the two data sources differ and models from each source best predict that data source's phenology events.
-The primary difference in the datasets is spatial scale, but due to trade-offs in data collection efforts, the larger scale NPN data has shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3).
+The primary difference in the datasets is spatial scale, but due to trade-offs in data collection efforts, the larger scale NPN data have shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3).
 These differences can be strengths or potential limitations.
 Observers sampling opportunistically allows the NPN dataset to have a large spatial scale, but also leads to low site fidelity which limits the ability to measure long-term trends at local scales \citep{gerst2016}.
-Tracking long-term trends is the major strength of LTER data, but having a relatively small species pool limits its use in species-level predictive modeling.
+Tracking long-term trends is the major strength of LTER data, but having a relatively small species pool limits their use in species-level predictive modeling.
 Due to these differences, the best data source for making predictions depends on the scale at which the predictions are being made.
 Identifying the most effective data sources for different types and scales of analysis is a useful first step, but the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018, melaas2016}.
-Our results suggest that methods that can learn from the intensive information available in LTER data in regions where it is available, and simultaneously use large-scale data to capture spatial variation in phenological requirements will help improve our ability to understand and predict phenology.
+Our results suggest that methods that can learn from the intensive information available in LTER data in regions where they are available, and simultaneously use large-scale data to capture spatial variation in phenological requirements will help improve our ability to understand and predict phenology.
 Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have both a large spatial extent and high temporal resolution \citep{richardson2018}.
@@
 \begin{adjustbox}{angle=90}
+\small
 %%{\def\arraystretch{2}\tabcolsep=5pt
@@
     Naive & \( \overline{DOY} \) & - & 1 & - \\
-    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = min(T_{i}, 0)$ & 1 & - \\
+    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = max(T_{i}, 0)$ & 1 & \citep{reaumur1735, wang1960, hunter1992} \\
     Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
-    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T^{*}, 0) $  & 3 & - \\
+    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T^{*}, 0) $  & 3 & \citep{reaumur1735, wang1960, hunter1992} \\
     M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T^{*}, 5) $  & 4 & \citep{blumel2012} \\

### 2018-08-14 manuscript/supplement_images.tex
progress as of today on revision
includes notes from meeting with Ethan

--- 2018-05-23 manuscript/supplement_images.tex
@@
 
-\textbf{\centering{\large Supplementary materials \par}} \newline
-Supplementary images S1 - S4
+\textbf{\large Supplementary materials}
+
+Supplementary images S1 - S7 and Table S1
 
@@
 		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
+	Figure S1
 \end{center}
@@
 		\includegraphics[width=1\textwidth]{supplement_best_lter_models.png}
-	\caption{Figure S2}
+	Figure S2
 \end{center}
@@
 		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
-	\caption{Figure S3}
+	Figure S3
 \end{center}
@@
 	\centering
-		\includegraphics[width=1\textwidth]{supplement_hubbard_harvard_comparison.png}
-	\caption{Figure S4}
+		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison1.png}
+	Figure S4
 \end{center}
 
+%%%%%%%%%%%%%%%%%
+%% Figure S5
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S5}: As in Figure S4, but for the Alternating and Uniforc models.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison2.png}
+	Figure S5
+\end{center}
+
+%%%%%%%%%%%%%%%%%
+%% Figure S6
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S6}: As in Figure S4, but for 4 selected species to show the difference in parameter distributions between LTER and NPN derived models.
+The phenophase for the four species is budburst.
+These 4 species are representative of the analysis, and for the remaining comparisons the reader is pointed to the script 'analysis/plot\_select\_species\_parameters.R' in the code repository to generate additional figures.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_select_species_param_comparison1.png}
+	Figure S6
+\end{center}
+
+%%%%%%%%%%%%%%%%%
+%% Figure S7
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S7}: As in Figure S6, but for the Alternating and Uniforc models.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_select_species_param_comparison2.png}
+	Figure S7
+\end{center}
+%%%%%%%%%%%%%%%%%
+%% Table S1
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Table S1}: Species used in the analysis along with the sample size from each dataset. The numbers indicate the sample size of the training data and the size of the testing data in parenthesis.
+
+\newpage
+
+\begin{table}
+\tiny
+\begin{tabular}{l|r|l|l|l|l|l|l}
+\hline
+species & phenophase & phenophase\_type & harvard & hjandrews & hubbard & jornada & npn\\
+\hline
+acer circinatum & 371 & Budburst & - & 266 (66) & - & - & 39 (10)\\
+\hline
+acer circinatum & 501 & Flowers & - & 116 (29) & - & - & 33 (8)\\
+\hline
+acer pensylvanicum & 371 & Budburst & 80 (20) & - & - & - & 34 (9)\\
+\hline
+acer rubrum & 371 & Budburst & 100 (25) & - & - & - & 957 (239)\\
+\hline
+acer rubrum & 501 & Flowers & 96 (24) & - & - & - & 668 (167)\\
+\hline
+acer saccharum & 371 & Budburst & 60 (15) & - & 164 (41) & - & 365 (91)\\
+\hline
+betula alleghaniensis & 371 & Budburst & 60 (15) & - & 178 (44) & - & 133 (33)\\
+\hline
+betula alleghaniensis & 501 & Flowers & 26 (7) & - & - & - & 64 (16)\\
+\hline
+betula lenta & 371 & Budburst & 58 (14) & - & - & - & 96 (24)\\
+\hline
+betula papyrifera & 371 & Budburst & 76 (19) & - & - & - & 96 (26)\\
+\hline
+betula papyrifera & 501 & Flowers & 18 (5) & - & - & - & 36 (9)\\
+\hline
+fagus grandifolia & 371 & Budburst & 76 (19) & - & 177 (44) & - & 259 (65)\\
+\hline
+fraxinus americana & 371 & Budburst & 84 (21) & - & - & - & 90 (23)\\
+\hline
+fraxinus americana & 501 & Flowers & 22 (6) & - & - & - & 52 (13)\\
+\hline
+ilex verticillata & 371 & Budburst & 35 (9) & - & - & - & 26 (6)\\
+\hline
+larrea tridentata & 501 & Flowers & - & - & - & 27 (7) & 118 (30)\\
+\hline
+nyssa sylvatica & 371 & Budburst & 27 (7) & - & - & - & 63 (16)\\
+\hline
+pinus strobus & 496 & Budburst & 38 (10) & - & - & - & 77 (19)\\
+\hline
+populus tremuloides & 371 & Budburst & 38 (10) & - & - & - & 208 (51)\\
+\hline
+populus tremuloides & 501 & Flowers & 17 (4) & - & - & - & 79 (22)\\
+\hline
+prosopis glandulosa & 501 & Flowers & - & - & - & 49 (12) & 78 (20)\\
+\hline
+prunus serotina & 371 & Budburst & 58 (14) & - & - & - & 228 (57)\\
+\hline
+pseudotsuga menziesii & 480 & Budburst & - & 182 (46) & - & - & 38 (10)\\
+\hline
+quercus alba & 371 & Budburst & 62 (15) & - & - & - & 174 (43)\\
+\hline
+quercus rubra & 371 & Budburst & 80 (20) & - & - & - & 242 (60)\\
+\hline
+quercus rubra & 501 & Flowers & 56 (14) & - & - & - & 127 (32)\\
+\hline
+quercus velutina & 371 & Budburst & 77 (19) & - & - & - & 72 (18)\\
+\hline
+rhododendron macrophyllum & 371 & Budburst & - & 84 (21) & - & - & 48 (12)\\
+\hline
+rhododendron macrophyllum & 501 & Flowers & - & 27 (7) & - & - & 50 (12)\\
+\hline
+trillium ovatum & 488 & Budburst & - & 222 (55) & - & - & 68 (17)\\
+\hline
+trillium ovatum & 501 & Flowers & - & 169 (42) & - & - & 60 (15)\\
+\hline
+vaccinium corymbosum & 371 & Budburst & 38 (10) & - & - & - & 60 (15)\\
+\hline
+vaccinium corymbosum & 501 & Flowers & 38 (10) & - & - & - & 65 (16)\\
+\hline
+vaccinium parvifolium & 371 & Budburst & - & 149 (37) & - & - & 25 (6)\\
+\hline
+vaccinium parvifolium & 501 & Flowers & - & 162 (41) & - & - & 25 (6)\\
+\hline
+\end{tabular}
+\newline
+Table S1
+\end{table}
+
 \end{document}}

### 2018-08-28 manuscript/Reviewer_Reply.md
progress on revision

@@
 
-These dates were used in accordance with Crimmins (2017), who used them prior, as budburst is unexpected to be first seen past DOY 172 and DOY 213 for flowers.
+These dates were used in accordance with Crimmins (2017), who used them prior, as budburst is unexpected to be first seen past DOY 172 (June 21) and DOY 213 (Aug. 1) for flowers.
 
@@
 
-We have added this clarification in several places within the text to avoid any confusion.
+We have made this change throughout the text to avoid any confusion.
 
@@
 
+We have added the correct citation for the dataset (USA National Phenology Network, 2017)
+
 ### Abstract
@@
 
+Spatial scale is the primary difference between the two datasets used here, but Reviewer 1 is correct in that local adaptation or species sensitivity could be the mechanisms behind the differences seen at different scales.
+We have clarified this in the abstract as well as the 2nd paragraph of the discussion.
+
 **L33: you mention that the NPN dataset offers many species… though this is a good point, it is actually irrelevant to the arguments that you are making in this paragraph, that is, if your comparison is focused on only a handful of species.
@@
 
-Reviewer 2 is correct in that our analysis was limited by species common to the NPN and four LTER datasets.
+Reviewer 1 is correct in that our analysis was limited by species common to the NPN and four LTER datasets.
 Large scale phenology studies typically employ as many species as are available in a given dataset.
-Thus we point out the large species pool of the NPN dataset throughout as it is something other researchers should consider when designing their own analysis.
+Thus we point out the large species pool of the USA-NPN dataset throughout as it is something other researchers should consider when designing their own analysis.
 
@@
 **L80: may also want to check out Feldman et al. (2018) – https://doi.org/10.1007/s00484-018-1540-4 for further support on cit sci observer skill**
+
+Thank you. We have added this reference as there are very few studies looking at observer skill at this level of detail.
 
@@
 
+While we feel anecdotally that this statement is true, to our knowledge no formal analysis on model usage has been done. Thus we have removed this line.
+
 **L156: I’m sure you extracted the daily temp at the lat/long of each observation location – maybe just make that a bit clearer**
@@
 
+We have adde
 Suggestion from Ethan:
@@
 
-We added two references here to highlight the potential variation in phenological requirements across different gradients.
+We added two references as well as some potential drivers here to highlight the potential variation in phenological requirements across different gradients.
 
@@
 
-Maybe change to "our analyses" ?
+We have clarified this by changing it to "..using the analyses in the current study".
 
@@
 
-We were referencing the two studies cited in this paragraph (Chuine et al. 2016 and Basler 2016). We have changed this to "In combination with the aforementioned studies,..."
+We were referencing the two studies cited in the same paragraph (Chuine et al. 2016 and Basler 2016). We have changed this to "In combination with the aforementioned studies,..."
 
@@
 
-Shawn's notes on some refs to add:
-comparing phenocam + modis: https://doi.org/10.1038/s41598-018-23804-6
-comparing modis phenology w/ NPN & Flux towers: 10.1016/j.ecolind.2017.02.024
+We have added two additional citations which perform analysis using different phenological datasets.
 

### 2018-08-28 manuscript/main.tex
progress on revision

@@
 This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability.
-Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science USA National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites.
+Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science USA National Phenology Network (USA-NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites.
 We built statistical and process based phenology models for species common to each dataset.
-From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both NPN and LTER data.
+From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both USA-NPN and LTER data.
 We found that model parameter estimates for the same species were most similar between the two datasets when using simple models, but parameter estimates varied widely as model complexity increased.
 Despite this, estimates for the date of phenological events and out-of-sample errors were similar, regardless of the model chosen.
-Predictions for NPN data had the lowest error when using models built from the NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built.
+Predictions for USA-NPN data had the lowest error when using models built from the USA-NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built.
+This difference is likely due to variation in phenological requirements within species.
+Models using the USA-NPN dataset can integrate parameters over a large spatial scale while those using an LTER dataset can only estimate parameters for a single location.
 Accordingly, the choice of dataset depends on the research question.
-Inferences about species-specific phenological requirements are best made with LTER data, and if NPN or similar data are all that is available, then analyses should be limited to simple models.
-Large-scale predictive modeling is best done with the larger-scale NPN data, which has high spatial representation and a large regional species pool.
+Inferences about species-specific phenological requirements are best made with LTER data, and if USA-NPN or similar data are all that is available, then analyses should be limited to simple models.
+Large-scale predictive modeling is best done with the larger-scale USA-NPN data, which has high spatial representation and a large regional species pool.
 LTER datasets, on the other hand, have high site fidelity and thus characterize inter-annual variability extremely well.
@@
 
-\keywords{LTER, NPN, forecasting, budburst, flowering, data integration, scale}
+\keywords{LTER, USA-NPN, forecasting, budburst, flowering, data integration, scale}
 
@@
 Because these data are often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams.
-A relatively new citizen science project started in 2009, The USA National Phenology Network (NPN), collects phenology observations from volunteers throughout North America and makes the data openly available \citep{npncitation}.
+A relatively new citizen science project started in 2009, The USA National Phenology Network (USA-NPN), collects phenology observations from volunteers throughout the United States and makes the data openly available \citep{npncitation}.
 Data from this project have already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}.
-Large-scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the NPN dataset has the potential to meet these needs for North American plant species and communities.
+Large-scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the USA-NPN dataset has the potential to meet these needs for North American plant species and communities.
 However, the features that allow citizen science projects to collect data at large scales can also introduce spatial biases toward cities and easily-accessible areas, and variation in sampling effort and observer skill \citep{dickinson2010}.
-With thousands of participants across North America, the potential for variation among observers in their determination of species identification and dating of phenological events is high.
-While volunteers have been shown to be accurate at distinguishing different leaf and flower stages for plants, \citep{fuccillo2015}, observations are sometimes made sporadically across seasons, years, and locations.
+With thousands of participants, the potential for variation among observers in their determination of species identification and dating of phenological events is high.
+While volunteers have been shown to be accurate at distinguishing different leaf and flower stages for plants \citep{fuccillo2015} and can have high agreement on abundance estimates \citep{feldman2018}, contributions to USA-NPN are sometimes made sporadically across seasons, years, and locations.
 This means that the quantity and quality of data at a specific site will typically be more variable for citizen science efforts than for intensive, long-term studies.
@@
 In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology and our ability to predict future phenology events.
-Here, we fit a suite of plant phenology models for the budburst and first-flowering phenophases of 24 plant species to data from both the NPN and a set of intensive long-term studies from the Long Term Ecological Research (LTER) network.
+Here, we fit a suite of plant phenology models for the budburst and first-flowering phenophases of 24 plant species to data from both the USA-NPN and a set of intensive long-term studies from the Long Term Ecological Research (LTER) network.
 We compare the resulting models based on both inference about models and parameters and predictions for unobserved events.
@@
 Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves.
-Sites in the NPN datasets are located across the U.S.
+Sites in the USA-NPN datasets are located across the U.S.
 and generally clustered around populated areas (Fig.
@@
 
-We downloaded all NPN observations from 2009, when collections began, to 2016 for the following phenophases: Breaking Leaf Buds, Breaking Needle Buds, Emerging Needles, and Open Flowers.
+We downloaded all USA-NPN observations from 2009, when collections began, to 2016 for the following phenophases: Breaking Leaf Buds, Breaking Needle Buds, Emerging Needles, and Open Flowers \citep{npndata}.
 The first three phenophases apply to the 'leaf out' phase for deciduous broadleafs, evergreen conifers, and pines, respectively.
@@
 Hereafter, we will refer to these as either 'Flowers' for the Open Flower phenophase, or 'Budburst' for all other phenophases.
-We subset the NPN observations similar to methods outlined in \cite{crimmins2017}.
+We subset the USA-NPN observations similar to methods outlined in \cite{crimmins2017}.
 First, 'yes' observations for individual plants were kept only if they were preceded by a 'no' observation within 30 days.
@@
 In the LTER datasets observation metrics varied widely due to different protocols.
-To match the NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods).
-As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations.
-After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table 1 \& S1).
+To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods).
+As with the USA-NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations.
+After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& S1).
 
@@
 It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}.
-For each of the 38 species and phenophase combinations in the five datasets (NPN and four LTER datasets), we fit eight phenology models (Table 2).
+For each of the 38 species and phenophase combinations in the five datasets (USA-NPN and four LTER datasets), we fit eight phenology models (Table 2).
 The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY.
@@
 The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}.
-These models are some of the most commonly used in phenology research and serve as a suitable baseline for comparing the NPN and LTER datasets.
 
@@
 The second, the \textit{Macroscale Species-specific Budburst model (MSB)}, uses the mean spring temperature as a linear correction on the total forcing required in the \textit{Alternating model} \citep{jeong2013}.
-Since there is little to no spatial variation in the LTER datasets, we fit the two spatial models to data from the NPN only.
-We compared the resulting parameters, estimates, and errors for the NPN-derived \textit{M1} and \textit{MSB} models to their non-spatial analogs (the \textit{GDD} and \textit{Alternating models}, respectively) for each species and phenophase in the LTER data.
+Since there is little to no spatial variation in the LTER datasets, we fit the two spatial models to data from the USA-NPN only.
+We compared the resulting parameters, estimates, and errors for the USA-NPN derived \textit{M1} and \textit{MSB} models to their non-spatial analogs (the \textit{GDD} and \textit{Alternating models}, respectively) for each species and phenophase in the LTER data.
 
-We extracted corresponding daily mean temperature for all NPN and LTER observations from the gridded PRISM dataset using the latitude and longitude of the site associated with each observation \citep{prismdata}.
+We extracted corresponding daily mean temperature for all USA-NPN and LTER observations from the gridded PRISM dataset using the latitude and longitude of the site associated with each observation \citep{prismdata}.
 We parameterized all models using differential evolution to minimize the root mean square error (RMSE) of the estimated DOY of the phenological event.
@@
 
-As described above, we fit two sets of models for each species and phenophase: one set of models parameterized using only NPN data, and one set parameterized using only LTER data (with the exception of the \textit{M1} and \textit{MSB} models, see above).
-To compare the inferences about process made by the two datasets, we compared the distribution of each parameter between LTER and NPN-derived models for each species and phenophase combination.
-Using the mean value of each bootstrapped parameter, we also calculated the coefficient of determination ($R^2$) between LTER and NPN-derived models among the 38 species-phenophases.
-In three cases where a species phenophase combination occurred in two LTER sites (Budburst for \textit{Acer saccharum}, \textit{Betula alleghaniensis}, and \textit{Fagus grandifolia} in the Harvard and Hubbard Brook datasets) they were compared separately to the NPN data.
+As described above, we fit two sets of models for each species and phenophase: one set of models parameterized using only USA-NPN data, and one set parameterized using only LTER data (with the exception of the \textit{M1} and \textit{MSB} models, see above).
+To compare the inferences about process made by the two datasets, we compared the distribution of each parameter between LTER and USA-NPN derived models for each species and phenophase combination.
+Using the mean value of each bootstrapped parameter, we also calculated the coefficient of determination ($R^2$) between LTER and USA-NPN derived models among the 38 species-phenophases.
+In three cases where a species phenophase combination occurred in two LTER sites (Budburst for \textit{Acer saccharum}, \textit{Betula alleghaniensis}, and \textit{Fagus grandifolia} in the Harvard and Hubbard Brook datasets) they were compared separately to the USA-NPN data.
 
 Models with different parameter values, and even entirely different structures, can produce similar estimates for the date of phenological events \citep{basler2016}.
-Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and NPN derived models for all held out observations.
-For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and NPN-derived estimates for estimates made at the four LTER sites and across all NPN sites.
+Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and USA-NPN derived models for all held out observations.
+For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and USA-NPN derived estimates for estimates made at the four LTER sites and across all USA-NPN sites.
 
-We also directly evaluated model performance using the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) NPN-derived models predicting LTER observations, C) LTER-derived models predicting NPN observations, and D) NPN-derived models predicting NPN observations.
-Using the RMSE values from held out observations, we  compared the  performance of LTER and NPN-derived models on different data types in two different ways.
-First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and NPN-derived models solely with LTER observations.
-Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely NPN data.
+We also directly evaluated model performance using the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) USA-NPN derived models predicting LTER observations, C) LTER-derived models predicting USA-NPN observations, and D) USA-NPN derived models predicting USA-NPN observations.
+Using the RMSE values from held out observations, we  compared the  performance of LTER and USA-NPN derived models on different data types in two different ways.
+First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and USA-NPN derived models solely with LTER observations.
+Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely USA-NPN data.
 These differences were calculated for each of the model types and 38 species-phenophase combinations.
-Negative values indicate that LTER-derived models perform better, while positive values indicate that the NPN-derived model performed better.
+Negative values indicate that LTER-derived models perform better, while positive values indicate that the USA-NPN derived model performed better.
+We used a t-test to test the difference from zero in these values.
 In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available.
@@
 
-The best matches between parameter estimates based on NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
+The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
 The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Fig.
 2).
-The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the NPN dataset (Fig.
+The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the USA-NPN dataset (Fig.
 2).
 The large outlier for the \textit{Fixed GDD model} is \textit{Larrea tridentata}; this species' flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
-While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between NPN and LTER-derived models (Fig.
+While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Fig.
 S6-S7).
 
-When comparing estimates of phenological events between the two sets of models, many NPN and LTER models produced similar estimates (Fig.
+When comparing estimates of phenological events between the two sets of models, many USA-NPN and LTER models produced similar estimates (Fig.
 3).
-The \textit{Fixed GDD model} had the highest correlation between the two models sets at NPN sites ($R^2 = 0.82$), while the \textit{GDD}, \textit{M1}, and \textit{Uniforc} models had the highest correlation at LTER sites ($R^2 =$ 0.51, 0.52, and 0.51, respectively).
+The \textit{Fixed GDD model} had the highest correlation between the two models sets at USA-NPN sites ($R^2 = 0.82$), while the \textit{GDD}, \textit{M1}, and \textit{Uniforc} models had the highest correlation at LTER sites ($R^2 =$ 0.51, 0.52, and 0.51, respectively).
 Comparing models with spatial corrections to the non-spatial alternatives, the \textit{MSB} (an extension of the \textit{Alternating model} with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the \textit{Alternating model}.
-The \textit{MSB model} improved the $R^2$ from 0.36 to 0.45 at LTER sites, and from -0.23 to -0.15 at NPN sites.
-The \textit{M1 model} (an extension of the \textit{GDD model} with a spatial correction based on day length) improved the correlation over the \textit{GDD model} only slightly at LTER sites (from 0.51 to 0.52) and did not improve the correlation at NPN sites.
+The \textit{MSB model} improved the $R^2$ from 0.36 to 0.45 at LTER sites, and from -0.23 to -0.15 at USA-NPN sites.
+The \textit{M1 model} (an extension of the \textit{GDD model} with a spatial correction based on day length) improved the correlation over the \textit{GDD model} only slightly at LTER sites (from 0.51 to 0.52) and did not improve the correlation at USA-NPN sites.
 
-When comparing the prediction accuracy on held-out data, NPN-derived models made more accurate predictions for held-out NPN observations, and LTER-derived models performed better on held-out LTER observations (Fig.
+When comparing the prediction accuracy on held-out data, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all p-values < 0.001, Fig.
 4).
-The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both NPN and LTER held-out observations.
-Although the \textit{Fixed GDD model} had the highest agreement in accuracy between NPN and LTER-derived models, it was not the best performing model overall.
-The \textit{GDD} and \textit{Uniforc} models commonly made the best predictions, having the lowest RMSE in 23\% and 40\% of cases among NPN-derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig.
+The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both USA-NPN and LTER held-out observations.
+Although the \textit{Fixed GDD model} had the highest agreement in accuracy between USA-NPN and LTER-derived models, it was not the best performing model overall.
+The \textit{GDD} and \textit{Uniforc} models commonly made the best predictions, having the lowest RMSE in 23\% and 40\% of cases among USA-NPN derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig.
 S1 \& S2).
@@
 
-Data used to build phenology models typically falls into two categories: intensive long-term data with long time-series at a small number of locations (e.g., LTER data in this study), and large-scale data with less intensive sampling at hundreds of locations (e.g., NPN data) (Table 3).
+Data used to build phenology models typically falls into two categories: intensive long-term data with long time-series at a small number of locations (e.g., LTER data in this study), and large-scale data with less intensive sampling at hundreds of locations (e.g., USA-NPN data) (Table 3).
 This data scenario--a small amount of intensive data and a large amount of less intensive data--is common in many areas of science and makes it necessary to understand how to choose between, or combine, data sources \citep{hanks2018}.
@@
 These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters.
-First, the differences could result from limitations in the sampling of the NPN dataset, such as irregular sampling of the same location within or between seasons, leading to less accurate parameter estimates.
-If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the Fixed GDD model is best for making inferences when NPN data are all that is available.
-Second, spatial variation in phenology requirements could drive these differences \citep{diez2012, zhang2017}, because NPN data integrate over that spatial variation, while LTER data only estimate the phenological requirements for a specific site.
-In this case, NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site.
+First, the differences could result from limitations in the sampling of the USA-NPN dataset, such as irregular sampling of the same location within or between seasons, leading to less accurate parameter estimates.
+If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the Fixed GDD model is best for making inferences when USA-NPN data are all that is available.
+Second, spatial variation (eg.
+from local adaptation, acclimation, microclimates, or plant age)  in phenology requirements and drivers could contribute to these differences \citep{diez2012, zhang2017}.
+Model built using USA-NPN data integrate over that spatial variation, while models built using LTER data only estimate the phenological requirements for a specific site.
+In this case, USA-NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site.
 The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}, although localized models could also be generated when large-scale predictions are unnecessary.
 Third, these differences could result from issues with model identifiability: since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same.
-Information about which of these issues may be causing the differences between datasets can be explored using these analyses, as will be explained below.
+Information about which of these issues may be causing the differences between datasets can be explored using the analyses in the current study, as will be explained below.
 
-Despite substantial differences in parameter estimates, LTER and NPN-derived models produced similar estimates for phenological events in most cases (Fig.
+Despite substantial differences in parameter estimates, LTER and USA-NPN derived models produced similar estimates for phenological events in most cases (Fig.
 3).
@@
 This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters.
-In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTER data (Fig.
+In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Fig.
 S6-S7).
@@
 S4-S5).
-These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estamates which were not biologically realistic \citep{chuine2016}.
+These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates which were not biologically realistic \citep{chuine2016}.
 \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting.
-Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and NPN datasets.
+Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and USA-NPN datasets.
 In combination with the aforementioned studies, our results indicate that caution is warranted in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.
@@
 While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the \textit{Fixed GDD model}, based on their ability to make out-of-sample predictions.
-Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both NPN and LTER datasets (Fig.
+Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both USA-NPN and LTER datasets (Fig.
 S1 \& S2).
@@
 4) for the \texit{MSB} and \textit{M1} models were essentially the same as the \textit{Alternating} and \textit{GDD} models, respectively.
-This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors.
+This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the USA-NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors.
 \cite{basler2016} used the \textit{M1} model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day.
@@
 \cite{chuine2017} listed the incorporation of spatial variation in warming requirements in models as a primary issue in future phenology research.
-Large-scale phenology datasets, like NPN, will be key in addressing this and other phenological research needs.
+Large-scale phenology datasets, like USA-NPN, will be key in addressing this and other phenological research needs.
 
-In conclusion, our results suggest that both LTER and NPN data provide valuable information on plant phenology.
+In conclusion, our results suggest that both LTER and USA-NPN data provide valuable information on plant phenology.
 Models built using both data sources yield effective predictions for phenological events, but parameter estimates from the two data sources differ and models from each source best predict that data source's phenology events.
-The primary difference in the datasets is spatial scale, but due to trade-offs in data collection efforts, the larger scale NPN data have shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3).
+The primary difference in the datasets is spatial scale, but due to trade-offs in data collection efforts, the larger scale USA-NPN data have shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3).
 These differences can be strengths or potential limitations.
-Observers sampling opportunistically allows the NPN dataset to have a large spatial scale, but also leads to low site fidelity which limits the ability to measure long-term trends at local scales \citep{gerst2016}.
+Observers sampling opportunistically allows the USA-NPN dataset to have a large spatial scale, but also leads to low site fidelity which limits the ability to measure long-term trends at local scales \citep{gerst2016}.
 Tracking long-term trends is the major strength of LTER data, but having a relatively small species pool limits their use in species-level predictive modeling.
@@
 Our results suggest that methods that can learn from the intensive information available in LTER data in regions where they are available, and simultaneously use large-scale data to capture spatial variation in phenological requirements will help improve our ability to understand and predict phenology.
-Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have both a large spatial extent and high temporal resolution \citep{richardson2018}.
+Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have both a large spatial extent and high temporal resolution \citep{peng2017,richardson2018a,richardson2018b}.
 Data integration provides the potential to use data from many sources to produce the best opportunity for accurate inference about, and forecasting of, the timing of biological events.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-%% Table 3, differences between LTER and NPN
+%% Table 3, differences between LTER and USA-NPN
 %%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{tabular}{lll}
-                                & \underline{LTER}  & \underline{NPN}           \\
+                                & \underline{LTER}  & \underline{USA-NPN}           \\
 Time-series length                          & \textbf{High} & \textbf{Low}  \\
@@
 
-\textbf{Figure 2}: Comparisons of parameter estimates between NPN and LTER derived models.
+\textbf{Figure 2}: Comparisons of parameter estimates between USA-NPN and LTER derived models.
 Each point represents a parameter value for a specific species and phenophase, and is the mean value from 250 bootstrap iterations.
@@
 
-\textbf{Figure 3}: Comparison of predicted day of year (DOY) of all phenological events between NPN and LTER-derived models.
-Top panels show comparisons at LTER sites and bottom panels show comparisons at NPN sites.
+\textbf{Figure 3}: Comparison of predicted day of year (DOY) of all phenological events between USA-NPN and LTER-derived models.
+Top panels show comparisons at LTER sites and bottom panels show comparisons at USA-NPN sites.
 Each point is an estimate for a single held-out observation.
@@
 
-\textbf{Figure 4}: Differences in prediction error between NPN and LTER-derived models.
-Density plots for comparisons of predictions on LTER data (top row) and NPN data (bottom row).
-Each plot represents the difference between the RMSE for LTER-derived model and the NPN-derived model, meaning that values less than zero indicate more accurate prediction by LTER-derived models and values greater than zero indicate more accurate prediction by NPN-derived models.
+\textbf{Figure 4}: Differences in prediction error between USA-NPN and LTER-derived models.
+Density plots for comparisons of predictions on LTER data (top row) and USA-NPN data (bottom row).
+Each plot represents the difference between the RMSE for LTER-derived model and the USA-NPN derived model, meaning that values less than zero indicate more accurate prediction by LTER-derived models and values greater than zero indicate more accurate prediction by NPN-derived models.
+\textit{p} <0.001 for all t-tests.
 Differences are calculated pairwise for the 38 species/phenophase comparisons.

### 2018-08-28 manuscript/supplement_images.tex
progress on revision

--- 2018-05-23 manuscript/supplement_images.tex
@@
 \usepackage{lineno}
-\usepackage{setspace}
-  \doublespacing
+    \linenumbers
+%\usepackage{setspace}
+%  \doublespacing
 \usepackage[document]{ragged2e}
@@
 \begin{document}
-\centering{\Large Paper title goes here \par}
 
-\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White}
+{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par} \newline
 
-\textbf{\centering{\large Supplementary materials \par}} \newline
-Supplementary images S1 - S4
+\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White} \newline
 
-%%%%%%%%%%%%%%%%%
-%% Figure S1
-%%%%%%%%%%%%%%%%%
-\textbf{Figure S1}: RMSE for specific species and phenophases of the NPN dataset. X marks the best performing models for the respective data type.
+{\Large \textbf{Supplementary materials}} \newline
+Methods describing processing of the 4 LTER datasets
 
-\newpage
+The four LTER datasets each had different protocols for recording phenology observations.
+Below are details for converting the data from each to a status based yes/no as used in the National Phenology Network.
+As in the USA-NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation.
+The years used for each datasets were all years available at the time of analysis.
 
-\begin{center}
-	\centering
-		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
-\end{center}
+\textbf{Harvard Forest} \newline
+We used observations from 1990-2014.
+Observations here were recorded as relative percentage of flowering or budburst for individual plants.
+We set "Yes" observations budburst and flower to the DOY when the percentage of each tree had was greater than or equal to 10\%
 
-\newpage
-%%%%%%%%%%%%%%%%%
-%% Figure S2
-%%%%%%%%%%%%%%%%%
-\textbf{Figure S2}: RMSE for specific species and phenophases of the LTER datasets. X marks the best performing models for the respective data type.
+\textbf{H.J. Andrews Experimental Forest} \newline
+We used observations from 2009-2015.
+We set "yes" observations for budburst to the first DOY when an individual was marked as "bud break" and "yes" observations for flowering were when an indiviudal was first marked as "Flowers open".
+Note that each species has slightly different ordinal codes to mark each of these events.
 
-\newpage
+\textbf{Jornada Experimental Range} \newline
+We used observations from 1992-2009.
+Observations for this dataset represent, for each zone, the percent of plants for a particular species which is observed within each phenophase.
+We set "Yes" observations for flowers to the first DOY where the flower phenophase was 10\% or greater.
 
-\begin{center}
-	\centering
-		\includegraphics[width=1\textwidth]{supplement_best_lter_models.png}
-	\caption{Figure S2}
-\end{center}
+\textbf{Hubbard Brook} \newline
+We used observations from 1989-2015. Observations for this dataset represent for each species the average, among 3 individuals, of an ordinal description of phenophase. \newline
 
-%%%%%%%%%%%%%%%%%
-%% Figure S3
-%%%%%%%%%%%%%%%%%
+0: winter conditions \newline
+1: bud swelling \newline
+2: small leaves or flowers \newline
+3: leafs 1/2 of final length, leafs obscure half the sky as seen thru crown \newline
+3.5: leaves 3/4 expanded, sky mostly obscured, crown not yet in summer condition \newline
+4: fully expanded, canopy in summer conditions \newline
 
-\newpage
-
-\textbf{Figure S3}: RMSE of all species and phenophases of the four scenarios described in the text. These values are calculated using held out test data.
-
-\newpage
-
-\begin{center}
-	\centering
-		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
-	\caption{Figure S3}
-\end{center}
-
-%%%%%%%%%%%%%%%%%
-%% Figure S4
-%%%%%%%%%%%%%%%%%
-\newpage
-
-\textbf{Figure S4}: Distribution of model parameters for the three species common to the Hubbard Brook, Harvard, and NPN datasets.
-The phenophase is budburst for all three species.
-Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution.
-
-\newpage
-
-\begin{center}
-	\centering
-		\includegraphics[width=1\textwidth]{supplement_hubbard_harvard_comparison.png}
-	\caption{Figure S4}
-\end{center}
+We set "Yes" observations for budburst to the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].
 

### 2018-08-28 manuscript/supplement_methods.tex
progress on revision

--- 2018-05-23 manuscript/supplement_methods.tex
@@
 \usepackage{lineno}
-    \linenumbers
-%\usepackage{setspace}
-%  \doublespacing
+\usepackage{setspace}
+  \doublespacing
 \usepackage[document]{ragged2e}
@@
 \begin{document}
+{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par}
+
+\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White}
+
+\textbf{\large Supplementary materials}
+
+Supplementary images S1 - S7 and Table S1
+
+%%%%%%%%%%%%%%%%%
+%% Figure S1
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S1}: RMSE for specific species and phenophases of the USA-NPN dataset. X marks the best performing models for each respective data type.
+
+\newpage
+
 \begin{center}
-{\huge Paper title goes here \par} \newline
-
-\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White} \newline
-
-{\Large \textbf{Supplementary materials}} \newline
-Methods describing processing of the 4 LTER datasets
+	\centering
+		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
+	Figure S1
 \end{center}
 
-The four LTER datasets each had different protocols for recording phenology observations.
-Below are details for converting the data from each to a status based yes/no as used in the National Phenology Network.
-As in the NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation.
+\newpage
+%%%%%%%%%%%%%%%%%
+%% Figure S2
+%%%%%%%%%%%%%%%%%
+\textbf{Figure S2}: RMSE for specific species and phenophases of the LTER datasets. X marks the best performing models for each respective data type.
 
-\textbf{Harvard Forest} \newline
-Observations here were recorded as relative percentage of flowering or budburst for individual plants.
-"Yes" observations budburst and flower where set to the DOY when the percentage of each tree had was greater than or equal to 10\%
+\newpage
 
-\textbf{H.J. Andrews Experimental Forest} \newline
-"yes" observations for budburst was the first DOY when an individual was marked as "bud break".
-"yes" observations for flowering was was when an indiviudal was first marked as "Flowers open".
-Note that each species has slightly different ordinal codes to make each of these events.
+\begin{center}
+	\centering
+		\includegraphics[width=1\textwidth]{supplement_best_lter_models.png}
+	Figure S2
+\end{center}
 
-\textbf{Jornada Experimental Range} \newline
-Observations for this dataset represent, for each zone, the percent of plants for a particular species which is observed within each phenophase.
-"Yes" observations for flowers was the first DOY where the flower phenophase was 10\% or greater.
+%%%%%%%%%%%%%%%%%
+%% Figure S3
+%%%%%%%%%%%%%%%%%
 
-\textbf{Hubbard Brook} \newline
-Observations for this dataset represent for each species the average, among 3 individuals, of an ordinal description of phenophase. \newline
+\newpage
 
-0: winter conditions \newline
-1: bud swelling \newline
-2: small leaves or flowers \newline
-3: leafs 1/2 of final length, leafs obscure half the sky as seen thru crown \newline
-3.5: leaves 3/4 expanded, sky mostly obscured, crown not yet in summer condition \newline
-4: fully expanded, canopy in summer conditions \newline
+\textbf{Figure S3}: RMSE of all species and phenophases of the four scenarios described in the text. These values were calculated using held out test data.
 
-"Yes" observations for budburst were the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
+	Figure S3
+\end{center}
+
+%%%%%%%%%%%%%%%%%
+%% Figure S4
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S4}: Distribution of parameters of the Naive, GDD, Fixed GDD, and Linear models for the three species common to the Hubbard Brook, Harvard, and USA-NPN datasets.
+The phenophase is budburst for all three species.
+Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution.
+Note the heading for each sub figure.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison1.png}
+	Figure S4
+\end{center}
+
+%%%%%%%%%%%%%%%%%
+%% Figure S5
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S5}: As in Figure S4, but for the Alternating and Uniforc models.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison2.png}
+	Figure S5
+\end{center}
+
+%%%%%%%%%%%%%%%%%
+%% Figure S6
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S6}: As in Figure S4, but for 4 selected species to show the difference in parameter distributions between LTER and USA-NPN derived models.
+The phenophase for the four species is budburst.
+These 4 species are representative of the analysis, and for the remaining comparisons the reader is pointed to the script 'analysis/plot\_select\_species\_parameters.R' in the code repository to generate additional figures.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_select_species_param_comparison1.png}
+	Figure S6
+\end{center}
+
+%%%%%%%%%%%%%%%%%
+%% Figure S7
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S7}: As in Figure S6, but for the Alternating and Uniforc models.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[scale=0.5]{supplement_select_species_param_comparison2.png}
+	Figure S7
+\end{center}
+%%%%%%%%%%%%%%%%%
+%% Table S1
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Table S1}: Species used in the analysis along with the sample size from each dataset. The numbers indicate the sample size of the training data and the size of the testing data in parenthesis.
+
+\newpage
+
+\begin{table}
+\tiny
+\begin{tabular}{l|r|l|l|l|l|l|l}
+\hline
+species & phenophase & phenophase\_type & harvard & hjandrews & hubbard & jornada & npn\\
+\hline
+acer circinatum & 371 & Budburst & - & 266 (66) & - & - & 39 (10)\\
+\hline
+acer circinatum & 501 & Flowers & - & 116 (29) & - & - & 33 (8)\\
+\hline
+acer pensylvanicum & 371 & Budburst & 80 (20) & - & - & - & 34 (9)\\
+\hline
+acer rubrum & 371 & Budburst & 100 (25) & - & - & - & 957 (239)\\
+\hline
+acer rubrum & 501 & Flowers & 96 (24) & - & - & - & 668 (167)\\
+\hline
+acer saccharum & 371 & Budburst & 60 (15) & - & 164 (41) & - & 365 (91)\\
+\hline
+betula alleghaniensis & 371 & Budburst & 60 (15) & - & 178 (44) & - & 133 (33)\\
+\hline
+betula alleghaniensis & 501 & Flowers & 26 (7) & - & - & - & 64 (16)\\
+\hline
+betula lenta & 371 & Budburst & 58 (14) & - & - & - & 96 (24)\\
+\hline
+betula papyrifera & 371 & Budburst & 76 (19) & - & - & - & 96 (26)\\
+\hline
+betula papyrifera & 501 & Flowers & 18 (5) & - & - & - & 36 (9)\\
+\hline
+fagus grandifolia & 371 & Budburst & 76 (19) & - & 177 (44) & - & 259 (65)\\
+\hline
+fraxinus americana & 371 & Budburst & 84 (21) & - & - & - & 90 (23)\\
+\hline
+fraxinus americana & 501 & Flowers & 22 (6) & - & - & - & 52 (13)\\
+\hline
+ilex verticillata & 371 & Budburst & 35 (9) & - & - & - & 26 (6)\\
+\hline
+larrea tridentata & 501 & Flowers & - & - & - & 27 (7) & 118 (30)\\
+\hline
+nyssa sylvatica & 371 & Budburst & 27 (7) & - & - & - & 63 (16)\\
+\hline
+pinus strobus & 496 & Budburst & 38 (10) & - & - & - & 77 (19)\\
+\hline
+populus tremuloides & 371 & Budburst & 38 (10) & - & - & - & 208 (51)\\
+\hline
+populus tremuloides & 501 & Flowers & 17 (4) & - & - & - & 79 (22)\\
+\hline
+prosopis glandulosa & 501 & Flowers & - & - & - & 49 (12) & 78 (20)\\
+\hline
+prunus serotina & 371 & Budburst & 58 (14) & - & - & - & 228 (57)\\
+\hline
+pseudotsuga menziesii & 480 & Budburst & - & 182 (46) & - & - & 38 (10)\\
+\hline
+quercus alba & 371 & Budburst & 62 (15) & - & - & - & 174 (43)\\
+\hline
+quercus rubra & 371 & Budburst & 80 (20) & - & - & - & 242 (60)\\
+\hline
+quercus rubra & 501 & Flowers & 56 (14) & - & - & - & 127 (32)\\
+\hline
+quercus velutina & 371 & Budburst & 77 (19) & - & - & - & 72 (18)\\
+\hline
+rhododendron macrophyllum & 371 & Budburst & - & 84 (21) & - & - & 48 (12)\\
+\hline
+rhododendron macrophyllum & 501 & Flowers & - & 27 (7) & - & - & 50 (12)\\
+\hline
+trillium ovatum & 488 & Budburst & - & 222 (55) & - & - & 68 (17)\\
+\hline
+trillium ovatum & 501 & Flowers & - & 169 (42) & - & - & 60 (15)\\
+\hline
+vaccinium corymbosum & 371 & Budburst & 38 (10) & - & - & - & 60 (15)\\
+\hline
+vaccinium corymbosum & 501 & Flowers & 38 (10) & - & - & - & 65 (16)\\
+\hline
+vaccinium parvifolium & 371 & Budburst & - & 149 (37) & - & - & 25 (6)\\
+\hline
+vaccinium parvifolium & 501 & Flowers & - & 162 (41) & - & - & 25 (6)\\
+\hline
+\end{tabular}
+\newline
+Table S1
+\end{table}
 

### 2018-08-29 analysis/observation_error_analysis.R
add t-test to figure 4

@@ -128,6 +128,15 @@ rmse_metrics = scenarios_error_data %>%
   mutate("With LTER Data" = A-B, "With USA-NPN Data" = C-D) %>%
   gather(metric, metric_value, "With LTER Data", "With USA-NPN Data")
 
+# All p-values for this figure are < 0.001, so these are not actually in the figure
+# but only mentioned in the text and caption
+test_statistics = rmse_metrics %>%
+  group_by(metric, model_name) %>%
+  summarise(t_stat = t.test(metric_value)$statistic,
+            p_value = t.test(metric_value)$p.value) %>%
+  ungroup() %>%
+  mutate(t_stat_text = paste0('italic(t) == ',round(t_stat,2)))
+
 # The annotated lines
 y_pos_line=0.15
 indicator_lines=data.frame(x=c(-12, 12), xend=c(-27, 27),
@@ -147,6 +156,7 @@ rmse_metrics_figure=ggplot(rmse_metrics, aes(metric_value)) +
   geom_segment(data=indicator_lines, aes(x=x, xend=xend, y=y, yend=yend), size=0.8, arrow = arrow(length=unit(0.25,'cm')),
                inherit.aes = FALSE) +
   geom_text(data=indicator_text, aes(x=x,y=y, label=t),size=4.5, inherit.aes = F) +
+  geom_text(data=test_statistics, aes(x=-24, y=0.07, label=t_stat_text), size=5, parse = TRUE) + 
   labs(y='',x='Difference between scenarios') +
   theme_bw() +
   theme(strip.text.x = element_text(size=20),
@@ -283,6 +293,7 @@ supplement_fig2 = ggplot(supplement_fig2_data, aes(x=model_name, y=error_value, 
 
 ggsave(supplement_fig2, filename = 'manuscript/supplement_best_lter_models.png',
        width = 40, height = 50, units = 'cm')
+
 ############################################
 ############################################
 library(kableExtra)

### 2018-09-12 manuscript/Reviewer_Reply.md
final revisions edits from shawn before sending to co-auths

@@
 Please explain more clearly how the <0 R-squareds arise.**
+
+Maybe change the Fig 2. caption to "..can be negative if the relationship between the two parameter sets is substantialy worse than a 1:1 relationship"
 
@@
 
+We tested the sensitivity of this cutoff time by repeating the analysis with a 15 day cutoff.
+We found no significant differences throughout the analysis.
+Three species (flowers and budburst for Vaccinium parvifolium and  budburst for Acer circinatum) were dropped from the 15 day analysis due to inadequate sample size.
+The overall sample size for the 15 day analysis was decreased by 10% (619 observations) and by 1-35% for individual species.
+We have included notes on this sensitivity test in the methods and results as well as additional supplementary material.
+
+Potential discussion on the technical points:
+
 Gerst et al.
-2016 looked at this issue and found that differences in site level means among more restrictive datasets was normally distributed.
-Thus having a shorter cutoff period here would result equally in less overestimates *and* underestimates.
-Since the models here are optimzed for RMSE we feel the results would not be affected overall by having a cutoff shorter than 30 days.
+2016 looked at the sensitivity of different thresholds in the USA-NPN dataset and found that differences in site level means among more restrictive datasets was normally distributed.
+We feel having a shorter cutoff period here would result equally in less overestimates as well underestimates, keeping the optimization surface of the model essentially the same.
+This is supported by the parameter estimates being nearly identical for the majority of cases (see new images S1-S3).
 
@@
 
-note:
-This can essentially be done by looking at some of the supplement figures. Will point this out in the discussion.
-shoot for 2-3 sentences describing some rules of thumb. with specific of error rates someone can be ok with.
+We have added an additional note describing the best models overall from our analysis and pointed readers to the supplemental images for specific details of model performance.
+We have also updated supplemental images S1 and S2 to better show model performance in the cross-dataset comparison, and also added an additional table (Table S2) which gives the overall model performance across all species and phenophases.
 
@@
 
+Three of the four LTER sites (Harard, Hubbard Brook, & H.J.
+Andrews) have very short sampling intervals (3-7 days).
+The Jornada LTER has a sample interval of 30 days.
+We have added these details to the supplementary methods.
+With the exception of the Jornada we feel the sampling interval of the LTER sites is very precise, and is one of the strengths of long term data that we point out throughout the manuscript.
+Given that, the issue of potentially 30 (or 15) days between sampling versus 7 is one of several points relating to volunteer based sampling we discuss in the manuscript.
+
+We note that only 2 of the 38 comparisons are from the Jornada datasets, so it's larger sampling frequency does not affect the overall analysis done in this paper.
+
 Shawns notes:
@@
 
-We have adde
-Suggestion from Ethan:
-do a paired t-test here to show distribution is different from 0. Make sure text still matches if a bunch are or are not significantly different than 0.
+We have added a t-test to show the difference from 0 in these distributions.
 
@@
 
-The frequency of sampling, in a given season and also between seasons, is the primary limit we consider here and we've made this clearer by stating as much.
+The frequency of sampling, in a given season and also between seasons, is the primary limit we consider here and we've added this clarification in the text.
 
@@
 We discussed this and ultimately decided that it is beyond the scope of the current paper.
-Although we agree that this approach has the potential to improve the paper, we are unsure of the extent to which this would address the issue of identifiability, which is itself noted as a potential issue rather than a definite one.
+Although we agree that this approach has the potential to improve the paper, we are doubtful if it would address the issue of identifiability.
+The Harvard and Hubbard Brook LTER share 3 species in common and are relatively close (~150km), yet their parameter estimates varied widely for most models (see images S4-S5 in the original and revised text).
 We also feel that it would require too much space to adequately investigate and discuss, and that the current set of analyses sufficiently accomplish the goals we set out in the introduction.
-However, we do hope to pursue these confounding factors further in future efforts.
 
@@
 
-Shawn's note:
-make a R^2 figure for all species/phenophase for the supplement
-on the stratified sampling.
-We thought of this, there's not enough data even for Acer rubrum.
-Write a sentence or 2 about this in the discussion and how it could be done someday when enought data is collected.
+See the note to Reviewer 1 about general model recommendations.
+We have updated the Supplemental images S1 and S2 to included Pearson's correlation in addition to the RMSE for all dataset inter comparisons.
 
@@
 **Line 281: Please verify what is T* (I assume base temperature)**
+
+This is correct. We have changed T* to T_base throughout as it is more descriptive.

### 2018-09-12 manuscript/main.tex
final revisions edits from shawn before sending to co-auths

@@
 
-In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology and our ability to predict future phenology events.
+In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology (eg.
+warming requirements for a specific plant) and our ability to predict future phenology events (eg.
+forecasting when flowering or leaf out occurs).
 Here, we fit a suite of plant phenology models for the budburst and first-flowering phenophases of 24 plant species to data from both the USA-NPN and a set of intensive long-term studies from the Long Term Ecological Research (LTER) network.
@@
 
-The USA National Phenology Network protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}.
+The USA National Phenology Network (USA-NPN) protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}.
 Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves.
@@
 We used 30 days to allow for a greater number of species to be compared.
+We tested the sensitivity of this choice by also performing the analysis using a 15 day cutoff.
 We chose not to group multiple individuals at a single site to better incorporate intra-site variability.
@@
 After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& S1).
+Using a 15 day cutoff in the USA-NPN dataset resulted in 35 unique combinations with 23 species .
 
@@
 The start day can either be estimated or fixed.
-For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$ \citep{reaumur1735, wang1960, hunter1992}.
-The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0$^{\circ}$C).
+For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T_{base}$ \citep{reaumur1735, wang1960, hunter1992}.
+The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T_{base}$ = 0$^{\circ}$C).
 The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan.
@@
 Absolute RMSE values are provided in the supplement (Fig.
-S1-S3).
+S5-S7).
 
@@
 
+Throughout the analysis no significant difference was found between a 30 day or 15 day threshold between the first 'yes' and most recent 'no' observation in the USA-NPN dataset.
+Results presented here reflect the 30 day cutoff, see the supplemtary figures S2-S4 for matching figures using a 15 day cutoff.
 The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
@@
 While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Fig.
-S6-S7).
+S10-S11).
 
@@
 
-When comparing the prediction accuracy on held-out data, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all p-values < 0.001, Fig.
+When comparing the prediction accuracy on held-out data, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Fig.
 4).
@@
 The \textit{GDD} and \textit{Uniforc} models commonly made the best predictions, having the lowest RMSE in 23\% and 40\% of cases among USA-NPN derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig.
-S1 \& S2).
+S5 \& S6).
 
@@
 This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues.
-For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T^*$=0 and $t_{1}$=5, $F$=5, $T^*$=0 produce nearly identical estimates in many scenarios.
+For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T_{base}$=0 and $t_{1}$=5, $F$=5, $T_{base}$=0 produce nearly identical estimates in many scenarios.
 This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters.
 In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Fig.
-S6-S7).
+S10-S11).
 Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig.
-S4-S5).
+S8-S9).
 These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates which were not biologically realistic \citep{chuine2016}.
@@
 Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both USA-NPN and LTER datasets (Fig.
-S1 \& S2).
+S5 \& S6).
 This demonstrates that the more complex models are capturing additional information about phenology, and that some of the differences between datasets result from differences in either the scales or the sampling of the data.
@@
 In combination with our results showing observed differences in parameter estimates between LTER sites (Fig.
-S4-S5), this suggests that variation in phenological requirements across the range is likely important.
+S8-S9), this suggests that variation in phenological requirements across the range is likely important.
 However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses.
@@
 Large-scale phenology datasets, like USA-NPN, will be key in addressing this and other phenological research needs.
+
+In addition to exploring differences between phenology datasets our analysis provides guidance on which models to use when making predictions at a local scale using models built from large scale data, or vice versa.
+Among the eight models tested the Uniforc model performed the best overall in the cross dataset comparison in terms of pearson correlation and RMSE (Table S2).
+One exception was the Fixed GDD model had a slightly higher pearson value when using LTER derived models to make predictions for USA-NPN data.
+When the scale is the same between predictions and training data (ie.
+using LTER derived models to make predictions for LTER data) the Uniforc, GDD, and M1 models performed equally well, though the Uniforc had slightly better RMSE (Table S2).
+The best model for specific species and phenophases varied, but was commonly the Uniforc and GDD models (Figs.
+S5-S6).
 
@@
     Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
-    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T^{*}, 0) $  & 3 & \citep{reaumur1735, wang1960, hunter1992} \\
-    M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T^{*}, 5) $  & 4 & \citep{blumel2012} \\
+    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T_{base}, 0) $  & 3 & \citep{reaumur1735, wang1960, hunter1992} \\
+    M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T_{base}, 5) $  & 4 & \citep{blumel2012} \\
     Alternating & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD(t)} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 3 & \citep{cannell1983} \\

### 2018-09-12 manuscript/supplement_images.tex
final revisions edits from shawn before sending to co-auths

--- 2018-08-14 manuscript/supplement_images.tex
@@
 \usepackage{a4wide}
+\usepackage{adjustbox}
+\usepackage{float}
 \usepackage{lineno}
@@
 \begin{document}
-{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par}
+{\Comparison of large-scale citizen science data and long-term study data for phenology modeling \par}
 
@@
 
-Supplementary images S1 - S7 and Table S1
+Supplementary images S1 - S11 and Table S1
 
@@
 
-\textbf{Figure S1}: RMSE for specific species and phenophases of the NPN dataset. X marks the best performing models for each respective data type.
+\textbf{Figure S1}: Sensitivity test results from using a 15 versus 30 day cutoff between the 'yes' and most recent 'no' in the USA-NPN dataset.
+Each point represents the value of a single parameter for one of eight models for 33 species and phenophases in the USA-NPN dataset.
+Note only 33 comparisons were possible here as the stricter 15 day cutoff resulted in three of the species not having sufficient observations.
 
@@
 	\centering
-		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
+		\includegraphics[width=1\textwidth]{parameter_estimates_from_cutoff_sensitivity.png}
 	Figure S1
@@
 \newpage
+
 %%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%
-\textbf{Figure S2}: RMSE for specific species and phenophases of the LTER datasets. X marks the best performing models for each respective data type.
+\newpage
+
+\textbf{Figure S2}: Comparisons of parameter estimates between NPN and LTER derived models.
+As in Figure 2 in the main text, but using a threshold of 15 instead of 30 days between the first 'yes' and most recent 'no' in the USA-NPN dataset.
+See methods for details.
 
@@
 	\centering
-		\includegraphics[width=1\textwidth]{supplement_best_lter_models.png}
+		\includegraphics[width=1\textwidth]{figure_param_comparison_15day_cutoff.png}
 	Figure S2
 \end{center}
+
+\newpage
 
@@
 %%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S3}: Comparison of predicted day of year (DOY) of all phenological events between NPN
+and LTER-derived models. As in Figure 3 in the main text, but using a threshold of 15 instead of 30 days between the first 'yes' and most recent 'no' in the USA-NPN dataset. See methods for details.
 
@@
 
-\textbf{Figure S3}: RMSE of all species and phenophases of the four scenarios described in the text. These values were calculated using held out test data.
+\begin{center}
+	\centering
+		\includegraphics[width=1\textwidth]{figure_estimate_compare_15day_cutoff.png}
+	Figure S3
+\end{center}
+
+\newpage
+
+%%%%%%%%%%%%%%%%%
+%% Figure S4
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S4}: Differences in prediction error between NPN and LTER-derived models.
+As in Figure 4 in the main text, but using a threshold of 15 instead of 30 days between the first 'yes' and most recent 'no' in the USA-NPN dataset.
+See methods for details.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[width=1\textwidth]{figure_rmse_metrics_density_plot_15day_cutoff.png}
+	Figure S4
+\end{center}
+
+\newpage
+
+%%%%%%%%%%%%%%%%%
+%% Figure S5
+%%%%%%%%%%%%%%%%%
+\newpage
+
+\textbf{Figure S5}: RMSE for specific species and phenophases using all combinations of models and data sources. Red X's mark the best performing models for each respective dataset.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[width=1\textwidth]{supplement_all_model_rmse.png}
+	Figure S5
+\end{center}
+
+\newpage
+%%%%%%%%%%%%%%%%%
+%% Figure S6
+%%%%%%%%%%%%%%%%%
+\textbf{Figure S6}: Pearson correlation coefficients for specific species and phenophases using all combinations of models and data sources.
+Note that since all predictions from each Naive model are the same pearson's correlation cannot be calculated here.
+Red X's mark the best performing models for each respective dataset.
+
+\newpage
+
+\begin{center}
+	\centering
+		\includegraphics[width=1\textwidth]{supplement_all_model_pearson.png}
+	Figure S6
+\end{center}
+
+%%%%%%%%%%%%%%%%%
+%% Figure S7
+%%%%%%%%%%%%%%%%%
+
+\newpage
+
+\textbf{Figure S7}: RMSE of all species and phenophases of the four scenarios described in the text. These values were calculated using held out test data.
 
@@
 		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
-	Figure S3
+	Figure S7
 \end{center}
@@
 %%%%%%%%%%%%%%%%%
-%% Figure S4
+%% Figure S8
 %%%%%%%%%%%%%%%%%
@@
 
-\textbf{Figure S4}: Distribution of parameters of the Naive, GDD, Fixed GDD, and Linear models for the three species common to the Hubbard Brook, Harvard, and NPN datasets.
+\textbf{Figure S8}: Distribution of parameters of the Naive, GDD, Fixed GDD, and Linear models for the three species common to the Hubbard Brook, Harvard, and USA-NPN datasets.
 The phenophase is budburst for all three species.
@@
 		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison1.png}
-	Figure S4
+	Figure S8
 \end{center}
@@
 %%%%%%%%%%%%%%%%%
-%% Figure S5
+%% Figure S9
 %%%%%%%%%%%%%%%%%
@@
 
-\textbf{Figure S5}: As in Figure S4, but for the Alternating and Uniforc models.
+\textbf{Figure S9}: As in Figure S4, but for the Alternating and Uniforc models.
 
@@
 		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison2.png}
-	Figure S5
+	Figure S9
 \end{center}
@@
 %%%%%%%%%%%%%%%%%
-%% Figure S6
+%% Figure S10
 %%%%%%%%%%%%%%%%%
@@
 
-\textbf{Figure S6}: As in Figure S4, but for 4 selected species to show the difference in parameter distributions between LTER and NPN derived models.
+\textbf{Figure S10}: As in Figure S4, but for 4 selected species to show the difference in parameter distributions between LTER and USA-NPN derived models.
 The phenophase for the four species is budburst.
@@
 		\includegraphics[scale=0.5]{supplement_select_species_param_comparison1.png}
-	Figure S6
+	Figure S10
 \end{center}
@@
 %%%%%%%%%%%%%%%%%
-%% Figure S7
+%% Figure S11
 %%%%%%%%%%%%%%%%%
@@
 
-\textbf{Figure S7}: As in Figure S6, but for the Alternating and Uniforc models.
+\textbf{Figure S11}: As in Figure S6, but for the Alternating and Uniforc models.
 
@@
 		\includegraphics[scale=0.5]{supplement_select_species_param_comparison2.png}
-	Figure S7
+	Figure S11
 \end{center}
@@
 
-\begin{table}
 \tiny
@@
 \newline
-Table S1
-\end{table}
+Table S1 \newline
+
+%%%%%%%%%%%%%%%%%
+%% Table S2
+%%%%%%%%%%%%%%%%%
+\newpage
+\normalsize
+\textbf{Table S2}: Overall best models when doing cross dataset comparisons.
+Observations for all species and phenophases were aggregated together to calculate RMSE and Pearsons coefficient for each combination of Parameter source (either USA-NPN or LTER), observation source (either USA-NPN or LTER), and model (6 possible phenology models).
+Bold indicates the best performing model for a specific parameter and observation combination, with some combinations having ties among multiple models.
+
+\newpage
+
+\tiny
+\begin{adjustbox}{angle=90}
+\begin{tabular}{|c|c|cc|cc|cc|cc|cc|cc|cc|cc|}
+\hline
+\begin{tabular}[c]{@{}c@{}}Parameter\\ Source\end{tabular} & \begin{tabular}[c]{@{}c@{}}Held Out\\ Observation Source\end{tabular} & \multicolumn{2}{c|}{Alternating} & \multicolumn{2}{c|}{Fixed GDD} & \multicolumn{2}{c|}{GDD} & \multicolumn{2}{c|}{Linear} & \multicolumn{2}{c|}{M1} & \multicolumn{2}{c|}{MSB} & \multicolumn{2}{c|}{Naive} & \multicolumn{2}{c|}{Uniforc} \\
+ &  & p & RMSE & p & RMSE & p & RMSE & p & RMSE & p & RMSE & p & RMSE & p & RMSE & p & RMSE \\ \hline
+LTER & LTER & 0.87 & 8.73 & 0.84 & 10.14 & \textbf{0.90} & 7.89 & 0.81 & 10.27 & \textbf{0.90} & 7.89 & 0.87 & 8.73 & 0.72 & 12.35 & \textbf{0.90} & \textbf{7.86} \\ \hline
+LTER & USA-NPN & 0.44 & 26.07 & \textbf{0.72} & 22.70 & 0.68 & 20.52 & 0.65 & 22.94 & 0.68 & 20.52 & 0.44 & 26.07 & 0.34 & 31.25 & 0.70 & \textbf{19.69} \\ \hline
+USA-NPN & LTER & 0.63 & 16.17 & 0.69 & 16.15 & 0.70 & 13.73 & 0.68 & 16.59 & 0.71 & 13.79 & 0.66 & 15.79 & 0.61 & 27.08 & \textbf{0.72} & \textbf{13.48} \\ \hline
+USA-NPN & USA-NPN & 0.80 & 15.27 & 0.76 & 20.18 & \textbf{0.82} & 14.67 & 0.77 & 16.19 & \textbf{0.82} & 14.71 & 0.80 & 15.20 & 0.53 & 21.57 & \textbf{0.82} & \textbf{14.37} \\ \hline
+\end{tabular}
+\end{adjustbox}
+\newline
+Table S2
 

### 2018-09-12 manuscript/supplement_methods.tex
final revisions edits from shawn before sending to co-auths

--- 2018-05-30 manuscript/supplement_methods.tex
@@
 
-{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par} \newline
+{\Comparison of large-scale citizen science data and long-term study data for phenology modeling \par} \newline
 
@@
 Below are details for converting the data from each to a status based yes/no as used in the National Phenology Network.
-As in the NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation.
+As in the USA-NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation.
 The years used for each datasets were all years available at the time of analysis.
@@
 Observations here were recorded as relative percentage of flowering or budburst for individual plants.
-We set "Yes" observations budburst and flower to the DOY when the percentage of each tree had was greater than or equal to 10\%
+We set "Yes" observations budburst and flower to the DOY when the percentage of each tree had was greater than or equal to 10\%.
+Harvard Forest has a sampling interval of 3-7 days.
 
@@
 We used observations from 2009-2015.
-We set "yes" observations for budburst to the first DOY when an individual was marked as "bud break" and "yes" observations for flowering were when an indiviudal was first marked as "Flowers open".
+We set "yes" observations for budburst to the first DOY when an individual was marked as "bud break" and "yes" observations for flowering were when an individual was first marked as "Flowers open".
 Note that each species has slightly different ordinal codes to mark each of these events.
+H.J.
+Andrews has a sampling interval of 7 days.
 
@@
 We set "Yes" observations for flowers to the first DOY where the flower phenophase was 10\% or greater.
+Jornada has a sampling interval of 30 days.
 
@@
 
-We set "Yes" observations for budburst to the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].
+We set "Yes" observations for budburst to the DOY when the average value was greater than or equal to 1.6.
+This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].
+Hubbard Brook has a sampling interval of 7 days.
 

### 2018-09-19 analysis/observation_error_analysis.R
remove overal best model stats

@@ -300,29 +300,3 @@ supplement_fig2 = ggplot(supplement_fig2_data, aes(x=model_name, y=error_value, 
 ggsave(supplement_fig2, filename = paste0(config$image_save_directory,'supplement_all_model_pearson.png'),
        width = 40, height = 50, units = 'cm')
 
-############################################
-############################################
-###
-# Stats for manuscript, not actually used anywhere. 
-
-# Winning model percentages for NPN observations
-npn_overal_best_models = winning_models_npn %>%
-  group_by(model_name, data_type) %>%
-  summarize(n=n()) %>%
-  ungroup() %>%
-  mutate(data_source='NPN')
-
-# Winning model percentages for LTS observations
-lts_overal_best_models = winning_models_lts %>%
-  group_by(model_name, data_type) %>%
-  summarize(n=n()) %>%
-  ungroup() %>%
-  mutate(data_source='LTER')
-
-overall_best_models = npn_overal_best_models %>%
-  bind_rows(lts_overal_best_models) %>%
-  spread(model_name, n, fill=0) %>%
-  select(data_source, data_type, Naive, Linear, 'Fixed GDD', GDD, M1, Uniforc, Alternating, MSB) %>%
-  arrange(data_source, data_type)
-
-#kable(overall_best_models, 'latex')

### 2018-09-19 analysis/observation_error_analysis.R
use only rmse in scenario error, remove old rankings

@@ -94,7 +94,7 @@ abbreviate_species_names = function(s){
 ###########################################################
 
 scenarios_error_data = model_errors %>%
-  filter(data_type=='test') %>%
+  filter(data_type=='test', error_type=='rmse') %>%
   mutate(scenario = case_when(
     is_lts_model & is_lts_obs ~ 'A',
     is_npn_model & is_lts_obs ~'B',
@@ -192,26 +192,6 @@ scenario_error = ggplot(scenarios_error_data, aes(x=species_phenophase, y=error_
   labs(y='RMSE',x='Species & Phenophase', color='Scenario') 
 
 ggsave(scenario_error, filename = paste0(config$image_save_directory,'supplement_scenario_absolute_rmse.png'), width = 30, height = 20, units = 'cm')
-
-
-#########################################################
-# relative rankings of the different scenarios
-# scenario_rankings = scenarios_error_data %>%
-#   select(model_name, species_phenophase, scenario, error_value) %>%
-#   group_by(model_name, species_phenophase) %>%
-#   arrange(error_value) %>%
-#   mutate(ranking = 1:4) %>%
-#   ungroup() %>%
-#   group_by(model_name, scenario) %>%
-#   summarise(ranking_1 = sum(ranking==1)/n(),
-#             ranking_2 = sum(ranking==2)/n(),
-#             ranking_3 = sum(ranking==3)/n(),
-#             ranking_4 = sum(ranking==4)/n()) %>%
-#   ungroup() %>%
-#   gather(ranking, rank_percent, ranking_1, ranking_2, ranking_3, ranking_4) %>%
-#   group_by(scenario, ranking) %>%
-#   summarize(rank_percent = round(mean(rank_percent),2)) %>%
-#   spread(scenario, rank_percent)
 
 
 ###########################################################

### 2018-09-19 analysis/lter_sampling_frequency.R
calculate sampling frequency from the data

@@ -1,0 +1,66 @@
+library(tidyverse)
+
+#############################
+# This calculates the sampling frequency from the actual data for all LTER sites
+# It matches the documented sampling frequency of 30 days for the Jornada and
+# 7 days for Harvard, HJ Andrews, and Hubbard Brook
+
+harvard = read_csv('raw_data/harvard_forest/hf003-03-spring.csv') %>%
+  mutate(year=lubridate::year(date)) %>%
+  select(doy=julian, year) %>%
+  distinct() %>%
+  mutate(dataset='harvard',site_id='1')
+
+jornada = read_csv('raw_data/jornada/JornadaStudy_287_npp_perennial_plant_phenology_0.csv', skip=45) %>%
+  mutate(date = as.Date(DATE, '%m/%d/%Y'), year=lubridate::year(date), doy = lubridate::yday(date)) %>%
+  select(year, doy, site_id = SITE) %>%
+  distinct() %>%
+  mutate(dataset='jornada')
+
+hjandrews = read_csv('raw_data/hjandrews/TV07501_v1_without_encoding.csv') %>%
+  mutate(year = lubridate::year(SAMPLEDATE), doy=lubridate::yday(SAMPLEDATE)) %>%
+  select(year, doy, site_id = PLOT) %>%
+  distinct() %>%
+  mutate(dataset='hjandrews')
+
+hubbard = read_csv('raw_data/hubbard_brook/phn.txt') %>%
+  gather(site, status, -Date, -DAY, -SEASON, -SPECIES) %>%
+  filter(SEASON=='SPRING') %>%
+  mutate(year=lubridate::year(Date)) %>%
+  select(year, doy=DAY, site_id = site) %>%
+  distinct() %>%
+  mutate(dataset='hubbard')
+
+sampling_dates = harvard %>%
+  bind_rows(jornada) %>%
+  bind_rows(hjandrews) %>%
+  bind_rows(hubbard)
+
+################################################
+sampling_dates = sampling_dates %>%
+  group_by(dataset, site_id, year) %>%
+  arrange(doy) %>%
+  mutate(sample_id = 1:n())
+
+sampling_dates_offset = sampling_dates %>%
+  mutate(sample_id = sample_id - 1) %>%
+  rename(next_doy=doy)
+
+sampling_dates = sampling_dates %>%
+  left_join(sampling_dates_offset, by=c('year','dataset','site_id','sample_id')) %>%
+  mutate(sampling_interval = next_doy - doy)
+
[…]

### 2018-09-19 manuscript/Reviewer_Reply.md
final version for 1st resubmission

@@
+We would like to thank the editor and two reviewers for the  productive feedback on our manuscript.
+We have addressed comments related to the sensitivity of the 30-day cutoff in using the USA-NPN data, and included general recommendations for which models performed best in different contexts.
+More detailed responses to individual comments are included below.
+
 #  Comments from the SME:
@@
 
-We added changed this to "statistical and processed based models".
+We changed this to "statistical and processed based models".
 
@@
 
-Maybe change the Fig 2. caption to "..can be negative if the relationship between the two parameter sets is substantialy worse than a 1:1 relationship"
+These values are coefficients of determination which compare the predictions from the model to the observed values.
+If there is a strong bias in the predictions this can lead to negative values even if the observations and predictions are correlated.
+Specifically, in cases where choosing the mean value of the observations provides a better prediction than the model these numbers will be negative.
+While this is not possible in simple regression without held out data (because the model by definition does at least as well as the mean) it is possible for other kinds of models and in cases where the model is not fit to the data it is evaluated on.
+We have attempted to clarify this in the text.
 
@@
 
-# Reviewer(s)' Comments to Author:
-
-## Reviewer: 1
+# Reviewer 1 Comments to Author:
 
@@
 
-After some discussion, we decided not to add subheadings.
+After some discussion, we decided not to add subheadings as both sections are relatively short, and each have a single paragraph dedicated to each analysis step.
+Instead we have made the Analysis introduction paragraph lay out the three primary steps, and also adjusted the wording to be more consistent.
 
@@
 We tested the sensitivity of this cutoff time by repeating the analysis with a 15 day cutoff.
-We found no significant differences throughout the analysis.
+This change did not influence our results.
 Three species (flowers and budburst for Vaccinium parvifolium and  budburst for Acer circinatum) were dropped from the 15 day analysis due to inadequate sample size.
 The overall sample size for the 15 day analysis was decreased by 10% (619 observations) and by 1-35% for individual species.
-We have included notes on this sensitivity test in the methods and results as well as additional supplementary material.
-
-Potential discussion on the technical points:
-
-Gerst et al.
-2016 looked at the sensitivity of different thresholds in the USA-NPN dataset and found that differences in site level means among more restrictive datasets was normally distributed.
-We feel having a shorter cutoff period here would result equally in less overestimates as well underestimates, keeping the optimization surface of the model essentially the same.
-This is supported by the parameter estimates being nearly identical for the majority of cases (see new images S1-S3).
-
-Shawn's note:
-After talking with Ethan I'll redo most of the analysis with a shorter cutoff date and see how things compare.
-also report the distribution of "days between" to see if they're very far off between 30day cutoff and a shorter cutoff.
+We have added notes on this sensitivity analysis to the methods and results as well as reporting the results in new supplementary material.
 
@@
 It may be just me…using the term “inference” in this way confused me.**
+
+This is a common usuage in the data science community and we are unclear how to better communicate this distinction between making predictions and understanding processes.
+To avoid confusion we have added a specific example to describe our usage of inference in the last introduction paragraph.
 
@@
 
-We have added an additional note describing the best models overall from our analysis and pointed readers to the supplemental images for specific details of model performance.
-We have also updated supplemental images S1 and S2 to better show model performance in the cross-dataset comparison, and also added an additional table (Table S2) which gives the overall model performance across all species and phenophases.
+We have added additional text to the Discussion describing the best models from our analysis and pointed readers to the supplemental images for specific details of model performance.
+We have also updated supplemental images S1 and S2 (now S5 & S6) to better show model performance in the cross-dataset comparison, and also added an additional table (Table S2) which gives the overall model performance across all species and phenophases.
 
@@
 
-We have made this change throughout the text to avoid any confusion.
+We have made this change throughout the text.
 
@@
 
-You're welcome. We're happy to credit all data providers, as this study would not have been possible without them.
+You're welcome.
+We're happy to credit all data providers, as this study would not have been possible without them.
+We have adjusted this line of text to reflect that Nature's Notebook is the citizen science program run by USA-NPN.
 
@@
 
-This is now made clear by specifying that it's the USA NPN.
+This is now made clear by specifying that it's the USA National Phenology Network.
 
@@
 The Jornada LTER has a sample interval of 30 days.
-We have added these details to the supplementary methods.
+We have added these details to the manuscript.
 With the exception of the Jornada we feel the sampling interval of the LTER sites is very precise, and is one of the strengths of long term data that we point out throughout the manuscript.
@@
 We note that only 2 of the 38 comparisons are from the Jornada datasets, so it's larger sampling frequency does not affect the overall analysis done in this paper.
-
-Shawns notes:
-put the numbers below in the table or ref. and re-reference them when i talk about reviewer 1's other issue with NPN sampling frequency
-
-sampling freq. from metadata / from data
-harvard  3-7 days / 5.6 days
-hubbard  7 days / 7.1 days
-hjandrews  7 days / 7.6 days
-jornada  monthly / 30.5 days
 
@@
 
-This referred to the RMSE values derived from LTER Model predictions for held out LTER data. We rewrote this paragraph slightly and changed the labels on Figure 4 to make this analysis more clear.
+This referred to the RMSE values derived from LTER Model predictions for held out LTER data. We rewrote this paragraph and changed the labels on Figure 4 to make this analysis more clear.
 
@@
 
-## Reviewer: 2
+# Reviewer 2 Comments to Author:
 
@@
 
-We discussed this and ultimately decided that it is beyond the scope of the current paper.
-Although we agree that this approach has the potential to improve the paper, we are doubtful if it would address the issue of identifiability.
-The Harvard and Hubbard Brook LTER share 3 species in common and are relatively close (~150km), yet their parameter estimates varied widely for most models (see images S4-S5 in the original and revised text).
-We also feel that it would require too much space to adequately investigate and discuss, and that the current set of analyses sufficiently accomplish the goals we set out in the introduction.
+We agree that this approach is a valuable next step, but ultimately decided that doing this properly would require a follow up paper rather than an additional analysis in this one.
+That said, we are doubtful that it would address the issue of identifiability.
+The Harvard and Hubbard Brook LTER share 3 species in common and are relatively close (~150km), yet their parameter estimates varied widely for most models (see images S4-S5 in the original text, now S8-S9 in the revised text).
 
@@
 We have updated the Supplemental images S1 and S2 to included Pearson's correlation in addition to the RMSE for all dataset inter comparisons.
+We decided that more expansive analyses were beyond the scope of this paper, but are hopeful that the open and reproducible nature of the code and data for this project will allow others (or our future selves) to conduct the important follow up analyses suggested by the reviewer.
 

### 2018-09-19 manuscript/main.tex
final version for 1st resubmission

@@
 Because these data are often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams.
-A relatively new citizen science project started in 2009, The USA National Phenology Network (USA-NPN), collects phenology observations from volunteers throughout the United States and makes the data openly available \citep{npncitation}.
+A relatively new citizen science project started in 2009, Nature's Notebook run by The USA National Phenology Network (USA-NPN), collects phenology observations from volunteers throughout the United States and makes the data openly available \citep{npncitation}.
 Data from this project have already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}.
@@
 
-In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology (eg.
-warming requirements for a specific plant) and our ability to predict future phenology events (eg.
+In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology (e.g.
+warming requirements for a specific plant) and our ability to predict future phenology events (e.g.
 forecasting when flowering or leaf out occurs).
@@
 To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods).
+Three of the LTER datasets (Hubbard Brook, Harvard Forest, and H.J.
+Andrews) had a sampling frequency of 3-7 days during the growing season.
+The Jornada dataset had a sampling frequency of 30 days.
 As with the USA-NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations.
 After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& S1).
-Using a 15 day cutoff in the USA-NPN dataset resulted in 35 unique combinations with 23 species .
+Using a 15 day cutoff in the USA-NPN dataset resulted in 35 unique combinations with 23 species.
 
@@
 1 (number of chill days - $NCD$).
-The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}.
+The \textit{Uniforc model} is like the \textit{GDD model} but with the forcing units transformed via a sigmoid function \citep{chuine2000}.
 
@@
 As described above, we fit two sets of models for each species and phenophase: one set of models parameterized using only USA-NPN data, and one set parameterized using only LTER data (with the exception of the \textit{M1} and \textit{MSB} models, see above).
+We performed three primary analyses from these model outputs by comparing: 1) the model parameters, 2) estimates from the models, and 3) out-of-sample errors from each model.
+
 To compare the inferences about process made by the two datasets, we compared the distribution of each parameter between LTER and USA-NPN derived models for each species and phenophase combination.
@@
 
+Next we compared the estimates of phenological events between models.
 Models with different parameter values, and even entirely different structures, can produce similar estimates for the date of phenological events \citep{basler2016}.
@@
 
-We also directly evaluated model performance using the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) USA-NPN derived models predicting LTER observations, C) LTER-derived models predicting USA-NPN observations, and D) USA-NPN derived models predicting USA-NPN observations.
+Finally, we directly evaluated performance using out-of-sample errors from the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) USA-NPN derived models predicting LTER observations, C) LTER-derived models predicting USA-NPN observations, and D) USA-NPN derived models predicting USA-NPN observations.
 Using the RMSE values from held out observations, we compared the performance of LTER and USA-NPN derived models on different data types in two different ways.
@@
 In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available.
-Absolute RMSE values are provided in the supplement (Fig.
-S5-S7).
+Absolute RMSE values as well as Pearson correlation coefficients are provided in the supplement for specific species (Fig.
+S5-S7) and with all observations aggregated together (Table S2).
 
@@
 
-Throughout the analysis no significant difference was found between a 30 day or 15 day threshold between the first 'yes' and most recent 'no' observation in the USA-NPN dataset.
-Results presented here reflect the 30 day cutoff, see the supplemtary figures S2-S4 for matching figures using a 15 day cutoff.
+Throughout the analysis there were no qualitative differences between a 30-day or 15-day threshold between the first 'yes' and most recent 'no' observation in the USA-NPN dataset.
+Results presented here reflect the 30 day cutoff, see the supplementary figures S2-S4 for matching figures using a 15 day cutoff.
+
 The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
@@
 
-When comparing the prediction accuracy on held-out data, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Fig.
+When comparing the predictive performance using out-of-sample errors, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Fig.
 4).
@@
 Although the \textit{Fixed GDD model} had the highest agreement in accuracy between USA-NPN and LTER-derived models, it was not the best performing model overall.
-The \textit{GDD} and \textit{Uniforc} models commonly made the best predictions, having the lowest RMSE in 23\% and 40\% of cases among USA-NPN derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig.
-S5 \& S6).
+The \textit{GDD} and \textit{Uniforc} models made the best out of sample predictions, having the lowest RMSE and Pearson coefficient when aggregating all observations together (Table S2).
+One exception was that the Fixed GDD model had a slightly higher Pearson value when using LTER-derived models to make predictions for USA-NPN data.
+The best model for specific species and phenophases varied, but was commonly the \textit{Uniforc} and \textit{GDD} models (Figs.
+S5-S6).
 
@@
 
-In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the Fixed GDD (Fig.
+In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the \textit{Fixed GDD} (Fig.
 2).
@@
 First, the differences could result from limitations in the sampling of the USA-NPN dataset, such as irregular sampling of the same location within or between seasons, leading to less accurate parameter estimates.
-If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the Fixed GDD model is best for making inferences when USA-NPN data are all that is available.
-Second, spatial variation (eg.
+If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the \textit{Fixed GDD model} is best for making inferences when USA-NPN data are all that is available.
+Second, spatial variation (e.g.
 from local adaptation, acclimation, microclimates, or plant age)  in phenology requirements and drivers could contribute to these differences \citep{diez2012, zhang2017}.
-Model built using USA-NPN data integrate over that spatial variation, while models built using LTER data only estimate the phenological requirements for a specific site.
+Models built using USA-NPN data integrate over that spatial variation, while models built using LTER data only estimate the phenological requirements for a specific site.
 In this case, USA-NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site.
@@
 This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues.
-For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T_{base}$=0 and $t_{1}$=5, $F$=5, $T_{base}$=0 produce nearly identical estimates in many scenarios.
+For example, two \textit{GDD} models with parameters of $t_{1}$=1, $F$=10, $T_{base}$=0 and $t_{1}$=5, $F$=5, $T_{base}$=0 produce nearly identical estimates in many scenarios.
 This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters.
@@
 S8-S9).
-These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates which were not biologically realistic \citep{chuine2016}.
+These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates that were not biologically realistic \citep{chuine2016}.
 \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting.
@@
 
-In addition to exploring differences between phenology datasets our analysis provides guidance on which models to use when making predictions at a local scale using models built from large scale data, or vice versa.
-Among the eight models tested the Uniforc model performed the best overall in the cross dataset comparison in terms of pearson correlation and RMSE (Table S2).
-One exception was the Fixed GDD model had a slightly higher pearson value when using LTER derived models to make predictions for USA-NPN data.
-When the scale is the same between predictions and training data (ie.
-using LTER derived models to make predictions for LTER data) the Uniforc, GDD, and M1 models performed equally well, though the Uniforc had slightly better RMSE (Table S2).
-The best model for specific species and phenophases varied, but was commonly the Uniforc and GDD models (Figs.
-S5-S6).
+In addition to exploring differences between phenology datasets, our analyses provide guidance on which models to use when making predictions at a local scale using models built from large-scale data, or vice versa.
+Among the eight models tested, the \textit{Uniforc} and \textit{GDD} models performed the best overall in the cross dataset comparison in terms of Pearson correlation and RMSE (Fig.
+S5-S6, Table S2).
+The \textit{GDD} model has one less parameter than the \textit{Uniforc} model, thus the \textit{GDD} model is a suitable choice for making predictions when there is little to no information at the location of interest (e.g.
+making phenology forecasts at a new location distant from any observed data).
+This guidance can vary between species though and model testing should still be performed when suitable data is available.
 
@@
 These differences can be strengths or potential limitations.
-Observers sampling opportunistically allows the USA-NPN dataset to have a large spatial scale, but also leads to low site fidelity which limits the ability to measure long-term trends at local scales \citep{gerst2016}.
+Observers sampling opportunistically allows the USA-NPN dataset to have a large spatial scale, but also leads to low site fidelity, which limits the ability to measure long-term trends at local scales \citep{gerst2016}.
 Tracking long-term trends is the major strength of LTER data, but having a relatively small species pool limits their use in species-level predictive modeling.
@@
 White.
+J.M.
+Meiners was supported by the University of Florida Biodiversity Institute Graduate Research Fellowship.
 We thank the developers and providers of the data that made this research possible including: the USA National Phenology Network and the many participants who contribute to its Nature’s Notebook program; the H.J.

### 2018-09-19 analysis/compare_hubbard_harvard.R
first revision (#42)
* progress as of today on revision

includes notes from meeting with Ethan

* progress on revision

* add t-test to figure 4

* add a supp. table with overall aggregate errors for best models

* change the layout of supp images 1 and 2

Before they were all RMSE of train and test data for all
NPN models (S1) and LTER models (S2). Now they are RMSE
for all models of only held out data (S1) and pearson
correlation for all models of only held out data (S2)

* change T* to T_base in all images

* differentat the new cutoff time differences in config

* new sensitivity analysis

@@ -28,6 +28,9 @@ all_parameters = all_parameters %>%
 
 all_parameters$phenophase = as.numeric(all_parameters$phenophase)
 
+#Make the threshold temperature name a bit more descriptive
+all_parameters$parameter_name[all_parameters$parameter_name=='T'] = 'T_base'
+
 ########################################################################
 datasets = c('harvard','hubbard','npn')
 pretty_dataset_names = c('Harvard Forest','Hubbard Brook','NPN')
@@ -93,5 +96,5 @@ hubbard_hubbard_2 =
   (uniforc + no_legend) +
   plot_layout(ncol=1, heights=c(1,1))
 
-ggsave('manuscript/supplement_hubbard_harvard_comparison1.png', plot=hubbard_hubbard_1, height=40, width=30, units = 'cm')
-ggsave('manuscript/supplement_hubbard_harvard_comparison2.png', plot=hubbard_hubbard_2, height=40, width=30, units = 'cm')
+ggsave(paste0(config$image_save_directory,'supplement_hubbard_harvard_comparison1.png'), plot=hubbard_hubbard_1, height=40, width=30, units = 'cm')
+ggsave(paste0(config$image_save_directory,'supplement_hubbard_harvard_comparison2.png'), plot=hubbard_hubbard_2, height=40, width=30, units = 'cm')

### 2018-09-19 analysis/compare_parameters.R

@@ -141,10 +141,10 @@ parameter_name_plotmath = tribble(
   'linear_temp','slope','beta[2]',
   'gdd','F','F',
   'gdd','t1','t[1]',
-  'gdd','T','T',
+  'gdd','T','T[base]',
   'm1','F','F',
   'm1','t1','t[1]',
-  'm1','T','T',
+  'm1','T','T[base]',
   'm1','k','k',
   'alternating','a','a',
   'alternating','b','b',
@@ -203,8 +203,8 @@ gdd=get_subplot(model_name = 'gdd', y_label = "GDD\n")
 gdd_fixed=get_subplot(model_name = 'gdd_fixed', y_label = "Fixed GDD\n")
 linear_temp=get_subplot('linear_temp', y_label = "Linear\n")
 naive=get_subplot('naive', y_label = "Naive\n") 
-m1=get_subplot('m1', y_label = "GDD\nCorrected")
-msb=get_subplot('msb', y_label = "Alternating\nCorrected")
+m1=get_subplot('m1', y_label = "M1")
+msb=get_subplot('msb', y_label = "MSB")
 
 legend = cowplot::get_legend(ggplot(filter(parameter_means, model=='uniforc'), aes(x=npn_derived_parameter, y=lts_derived_parameter, color=dataset, group=dataset))+
                                geom_point(size=4, aes(shape = phenophase)) +
@@ -234,6 +234,6 @@ whole_plot=gridExtra::grid.arrange(empty_space,naive, linear_temp, gdd, uniforc,
 
 
 
-ggsave('manuscript/figure_param_comparison.png', plot=whole_plot, height=33, width=20, units = 'cm')
+ggsave(paste0(config$image_save_directory,'figure_param_comparison.png'), plot=whole_plot, height=33, width=20, units = 'cm')
 
 

### 2018-09-19 analysis/cutoff_sensitivity_analysis.R

@@ -1,0 +1,78 @@
+library(tidyverse)
+
+# During review there was a question of the sensitivity of the 30 day threshold to NPN observations (see methods)
+# I tested that sensitivity by running all analysis again with a 15 day cutoff. This compares some of the stats on 
+# that (ie. the decrease in sample size from a more conservative cutoff) and also plots the 1:1 comparison of 
+# each estimated parameter between the two cutoffs
+
+
+##########################
+# The 1:1 charts of different parameters
+
+cutoff_15_day = read_csv('results/cutoff_15days/model_parameters.csv.gz')
+cutoff_15_day$cutoff = 'cutoff_15'
+
+cutoff_30_day = read_csv('results/cutoff_30days/model_parameters.csv.gz') 
+cutoff_30_day$cutoff = 'cutoff_30'
+
+
+both = cutoff_15_day %>%
+  bind_rows(cutoff_30_day) %>%
+  filter(dataset=='npn', !parameter_name %in% c('run_time','num_iterations') )  %>%
+  group_by(model, parameter_name, species, cutoff) %>%
+  summarise(parameter_value = mean(value)) %>%
+  ungroup() %>%
+  spread(cutoff, parameter_value) %>%
+  fortify() %>%
+  filter(complete.cases(.))
+
+sensitivity_figure = ggplot(both, aes(x=cutoff_15, y=cutoff_30)) + 
+  geom_point() + 
+  geom_abline(intercept = 0, slope=1) + 
+  facet_wrap(model~parameter_name, scales='free') + 
+  labs(x='Parameter estimates using a 15 day cutoff',
+       y='Parameter estimates using a 30 day cutoff')
+
+ggsave('manuscript/parameter_estimates_from_cutoff_sensitivity.png', plot=sensitivity_figure, height=20, width=30, units = 'cm')
+
+#############################################
+##########################
+# The decrease in sample size moving from a 30 day to a 15 day cutoff
+npn_species = cutoff_30_day %>%
+  filter(!parameter_name %in% c('run_time','num_iterations')) %>%
+  select(dataset, species) %>%
+  filter(dataset=='npn') %>%
+  distinct()
+
+###################
+
+sample_sizes_15d = read_csv('results/cutoff_15days/predictions.csv.gz') %>%
+  select(observation_source, species, observation_id) %>%
+  distinct() %>%
+  filter(species %in% npn_species$species) %>%
[…]

### 2018-09-19 analysis/observation_error_analysis.R

@@ -1,4 +1,6 @@
 library(tidyverse)
+library(kableExtra)
+library(knitr)
 
 config = yaml::yaml.load_file('config.yaml')
 
@@ -58,9 +60,11 @@ rm(lts_models)
 
 model_errors = predictions %>%
   group_by(data_type, model_name, observation_source, parameter_source, species, phenophase) %>%
-  summarise(rmse = sqrt(mean((doy_estimated - doy_observed)^2)), sample_size=n()) %>%
+  summarise(rmse = sqrt(mean((doy_estimated - doy_observed)^2)), 
+            pearson = cor(doy_estimated, doy_observed, method='pearson'),
+            sample_size=n()) %>%
   ungroup() %>%
-  gather(error_type, error_value, rmse) %>%
+  gather(error_type, error_value, rmse, pearson) %>%
   mutate(error_value = round(error_value,4)) %>%
   mutate(is_lts_model = parameter_source!='npn', is_lts_obs = observation_source != 'npn',
          is_npn_model = !is_lts_model, is_npn_obs = !is_lts_obs)
@@ -90,7 +94,7 @@ abbreviate_species_names = function(s){
 ###########################################################
 
 scenarios_error_data = model_errors %>%
-  filter(data_type=='test') %>%
+  filter(data_type=='test', error_type=='rmse') %>%
   mutate(scenario = case_when(
     is_lts_model & is_lts_obs ~ 'A',
     is_npn_model & is_lts_obs ~'B',
@@ -123,10 +127,20 @@ scenarios_error_data$species_phenophase = with(scenarios_error_data, paste(abbre
 # Main figure of the two metrics (scenario A - B and C -D)
 
 rmse_metrics = scenarios_error_data %>%
+  filter(error_type=='rmse') %>%
   select(model_name, species, phenophase, error_value, scenario) %>%
   spread(scenario, error_value) %>%
   mutate("With LTER Data" = A-B, "With USA-NPN Data" = C-D) %>%
   gather(metric, metric_value, "With LTER Data", "With USA-NPN Data")
+
+# All p-values for this figure are < 0.001, so these are not actually in the figure
+# but only mentioned in the text and caption
+test_statistics = rmse_metrics %>%
+  group_by(metric, model_name) %>%
+  summarise(t_stat = t.test(metric_value)$statistic,
+            p_value = t.test(metric_value)$p.value) %>%
+  ungroup() %>%
[…]

### 2018-09-19 analysis/observation_estimate_comparison.R

@@ -102,4 +102,4 @@ fig2=ggplot(observation_estimate_comparison, aes(x=doy_estimated_npn_model, y=do
   labs(x='Estimates from NPN derived models',
        y='Estimates from LTER derived models')
 
-ggsave(fig2, filename = 'manuscript/figure_estimate_compare.png', height = 14, width = 40, units = 'cm')
+ggsave(fig2, filename = paste0(config$image_save_directory,'figure_estimate_compare.png'), height = 14, width = 40, units = 'cm')

### 2018-09-19 analysis/plot_select_species_parameters.R

@@ -21,6 +21,9 @@ all_parameters = all_parameters %>%
 
 all_parameters$phenophase = as.numeric(all_parameters$phenophase)
 
+#Make the threshold temperature name a bit more descriptive
+all_parameters$parameter_name[all_parameters$parameter_name=='T'] = 'T_base'
+
 select_species_params = all_parameters %>%
   filter(species %in% select_species, 
          phenophase %in% select_phenophase)
@@ -91,5 +94,5 @@ select_species_2 =
   (uniforc + no_legend) +
   plot_layout(ncol=1, heights=c(1,1))
 
-ggsave('manuscript/supplement_select_species_param_comparison1.png', plot=select_species_1, height=40, width=30, units = 'cm')
-ggsave('manuscript/supplement_select_species_param_comparison2.png', plot=select_species_2, height=40, width=30, units = 'cm')
+ggsave(paste0(config$image_save_directory,'supplement_select_species_param_comparison1.png'), plot=select_species_1, height=40, width=30, units = 'cm')
+ggsave(paste0(config$image_save_directory,'supplement_select_species_param_comparison2.png'), plot=select_species_2, height=40, width=30, units = 'cm')

### 2018-10-22 analysis/compare_parameters.R
major improvements for figure 2
all subplots are rendered at once in a single plot now, making
everything nice and pretty. 5 minutes of editing in GIMP
is needed now to remove some spacer plots.

@@ -104,6 +104,13 @@ ggplot(x, aes(x=value, group=dataset, fill=dataset)) +
 
 
 ###############################################################################
+###############################################################################
+# Everything below if for making the fairly complicated Figure 2. 
+
+# Once this image is rendered I do some slight editing in a photo editor to mask out all the 
+# dummy variables and adjust the text and legend positioning. 
+###############################################################################
+###############################################################################
 #scatter plots of npn vs long term datasets
 
 budburst_phenophases = c(371, 496, 488, 480)
@@ -169,71 +176,117 @@ r2_values = parameter_means %>%
 # labels of the plot
 parameter_means = parameter_means %>%
   left_join(r2_values, by=c('model','parameter_name')) %>%
-  left_join(parameter_name_plotmath, by = c('model','parameter_name')) %>%
-  mutate(parameter_name2 = paste0('list(',parameter_symbol,',', r2_text,')'))
+  left_join(parameter_name_plotmath, by = c('model','parameter_name')) 
 
 #################################################################################
+# These dummy lines help create padding on the right side of Figure 2.
+dummy_parameters = tribble(
+~parameter_name,~facet_strip_text, ~model, ~lts_derived_parameter, ~npn_derived_parameter, ~r2_text,
+'naive_dummy1','naive_dummy1', 'naive', 1,1,'naive_dummy1',
+'naive_dummy2','naive_dummy2', 'naive', 1,1,'naive_dummy2',
+'naive_dummy3','naive_dummy3', 'naive', 1,1,'naive_dummy3',
 
-common_plot_theme = theme(strip.text = element_text(size=9),
-                          strip.background = element_rect(fill='grey95'),
-                          axis.text = element_text(size=9),
-                          axis.title = element_text(size=12))
+'fixed_gdd_dummy1','fixed_gdd_dummy1', 'gdd_fixed', 1,1,'fixed_gdd_dummy1',
+'fixed_gdd_dummy2','fixed_gdd_dummy2', 'gdd_fixed', 1,1,'fixed_gdd_dummy2',
+'fixed_gdd_dummy3','fixed_gdd_dummy3', 'gdd_fixed', 1,1,'fixed_gdd_dummy3',
 
-point_size=3
-point_shapes = c(17,13)
[…]

### 2018-10-22 manuscript_map.R
make the map work with ggsave, increase resolution

@@ -36,15 +36,16 @@ site_map = ggplot() +
   geom_raster(data = elevation, aes(x=lat, y=lon, fill=elevation)) +
   scale_fill_gradient(low='white', high='grey50') + 
   geom_polygon(data = basemap, aes(x=long, y = lat, group = group), fill=NA, color='grey20', size=0.5) + 
-  geom_point(data=npn_sites, aes(x=Longitude, y=Latitude), color='black', size=2) + 
+  geom_point(data=npn_sites, aes(x=Longitude, y=Latitude), color='black', size=1.5) + 
   geom_point(data=lts_sites, aes(x=Longitude, y=Latitude), color='black', shape=16, size=5) +
   geom_point(data=lts_sites, aes(x=Longitude, y=Latitude), color='white', shape=16, size=3) +
   geom_label_repel(data=filter(lts_sites, dataset %in% c('Hubbard Brook','Harvard Forest')), aes(x=Longitude, y=Latitude, label=dataset), force=10, 
-                   label.size = 0.8, segment.size = 1, nudge_y = 2.8, nudge_x = -5) + 
+                   label.size = 0.6, segment.size = 1, nudge_y = 2.8, nudge_x = -7, size=4) + 
   geom_label_repel(data=filter(lts_sites, !dataset %in% c('Hubbard Brook','Harvard Forest')), aes(x=Longitude, y=Latitude, label=dataset), force=10, 
-                   label.size = 0.8, segment.size = 1, nudge_y = -1, nudge_x = 3) + 
+                   label.size = 0.6, segment.size = 1, nudge_y = -2, nudge_x = 4, size=4) + 
   coord_fixed(1.3) +
   theme_bw() + 
+  labs(x=NULL, y=NULL) +
   theme(axis.title = element_blank(),
         axis.text = element_blank(),
         panel.grid = element_blank(),
@@ -52,8 +53,8 @@ site_map = ggplot() +
         panel.border = element_blank(),
         legend.position = 'none',
         plot.margin = unit(c(0,0,0,0),'mm'))
-print(site_map)
+#print(site_map)
 # Note, I saved this manually using the export tool to get ride of uneeded whitespace in the figure.
-ggsave('manuscript/figure_site_map.png', plot=site_map, height=20, width=50, units = 'cm')
+ggsave('manuscript/figure_site_map.png', width=20, height=16, units = 'cm',dpi=1000, plot=site_map)
 
 

### 2018-10-22 analysis/compare_hubbard_harvard.R
image name updates mostly

@@ -3,7 +3,7 @@ library(cowplot)
 config = yaml::yaml.load_file('config.yaml')
 
 ###################################################
-# This produced the supplementary figure comparing parameters
+# This produced the supplementary figures S8 and S9, comparing parameters
 # of hubbard, harvard, and NPN
 #################################################
 
@@ -96,5 +96,5 @@ hubbard_hubbard_2 =
   (uniforc + no_legend) +
   plot_layout(ncol=1, heights=c(1,1))
 
-ggsave(paste0(config$image_save_directory,'supplement_hubbard_harvard_comparison1.png'), plot=hubbard_hubbard_1, height=40, width=30, units = 'cm')
-ggsave(paste0(config$image_save_directory,'supplement_hubbard_harvard_comparison2.png'), plot=hubbard_hubbard_2, height=40, width=30, units = 'cm')
+ggsave(paste0(config$image_save_directory,'figure_s8_hubbard_harvard_comparison1.png'), plot=hubbard_hubbard_1, height=40, width=30,dpi=1000, units = 'cm')
+ggsave(paste0(config$image_save_directory,'figure_s9_hubbard_harvard_comparison2.png'), plot=hubbard_hubbard_2, height=40, width=30,dpi=1000, units = 'cm')

### 2018-10-22 analysis/compare_parameters.R

@@ -286,7 +286,7 @@ whole_plot = ggplot(parameters_with_dummy_vars, aes(x=npn_derived_parameter, y=l
         legend.text = element_text(size=14),
         legend.title = element_text(size=16)) 
 
-ggsave(paste0(config$image_save_directory,'figure_param_comparison.png'), plot=whole_plot, height=28, width=25, units = 'cm', dpi=1000)
+ggsave(paste0(config$image_save_directory,'figure_2_param_comparison.png'), plot=whole_plot, height=28, width=25, units = 'cm', dpi=1000)
 
 # Note, after rendering I deleted the dummy placeholder subplots inside a photo editing program and save as figure_param_comparison_final.png
 

### 2018-10-22 analysis/cutoff_sensitivity_analysis.R

@@ -33,7 +33,7 @@ sensitivity_figure = ggplot(both, aes(x=cutoff_15, y=cutoff_30)) +
   labs(x='Parameter estimates using a 15 day cutoff',
        y='Parameter estimates using a 30 day cutoff')
 
-ggsave('manuscript/parameter_estimates_from_cutoff_sensitivity.png', plot=sensitivity_figure, height=20, width=30, units = 'cm')
+ggsave('manuscript/figure_s1_parameter_estimates_from_cutoff_sensitivity.png', plot=sensitivity_figure, height=20, width=30,dpi=1000, units = 'cm')
 
 #############################################
 ##########################

### 2018-10-22 analysis/observation_error_analysis.R

@@ -170,11 +170,10 @@ rmse_metrics_figure=ggplot(rmse_metrics, aes(metric_value)) +
         axis.title.x = element_text(size=20),
         strip.background = element_rect(fill='grey95'))
 
-ggsave(rmse_metrics_figure, filename = paste0(config$image_save_directory,'figure_rmse_metrics_density_plot.png'), width = 60, height = 15, units = 'cm')
-
+ggsave(rmse_metrics_figure, filename = paste0(config$image_save_directory,'figure_4_rmse_metrics_density_plot.png'), width = 60, height = 15,dpi=1000, units = 'cm')
 ######################################################
 ######################################################
-# supplement figure, absolute RMSE values of all scenarios
+# supplement figure 7, absolute RMSE values of all scenarios
 
 scenario_error = ggplot(scenarios_error_data, aes(x=species_phenophase, y=error_value, group=scenario, color=scenario)) +
   geom_point() +
@@ -191,7 +190,7 @@ scenario_error = ggplot(scenarios_error_data, aes(x=species_phenophase, y=error_
         legend.direction = "horizontal") +
   labs(y='RMSE',x='Species & Phenophase', color='Scenario') 
 
-ggsave(scenario_error, filename = paste0(config$image_save_directory,'supplement_scenario_absolute_rmse.png'), width = 30, height = 20, units = 'cm')
+ggsave(scenario_error, filename = paste0(config$image_save_directory,'figure_s7_scenario_absolute_rmse.png'), width = 30, height = 20, dpi=1000, units = 'cm')
 
 
 ###########################################################
@@ -226,57 +225,58 @@ kable(best_scenario_model, 'latex')
 
 ############################################
 ############################################
-# Supplement figure 1. Species level out of sample RMSE
+# Supplement figure 5. Species level out of sample RMSE
 # or, if using  NPN data which model is best for each species/phenophase?
 
-supplement_fig_12_theme =  theme(axis.text.x = element_text(size=10,angle=90,hjust = 0.2,vjust = 0.8, debug = FALSE),
+supplement_fig_5_6_theme =  theme(axis.text.x = element_text(size=10,angle=90,hjust = 0.2,vjust = 0.8, debug = FALSE),
                                  axis.title = element_text(size=15),
-                                 strip.text = element_text(size=6.5)) 
+                                 strip.text = element_text(size=6.5)) +
[…]

### 2018-10-22 analysis/observation_estimate_comparison.R

@@ -1,5 +1,6 @@
 library(tidyverse)
 #Compare out of sample verification between datasets
+# produces Figure 3, a comparison of model estimates. 
 config = yaml::yaml.load_file('config.yaml')
 
 predictions = read_csv(config$predictions_file)
@@ -102,4 +103,4 @@ fig2=ggplot(observation_estimate_comparison, aes(x=doy_estimated_npn_model, y=do
   labs(x='Estimates from NPN derived models',
        y='Estimates from LTER derived models')
 
-ggsave(fig2, filename = paste0(config$image_save_directory,'figure_estimate_compare.png'), height = 14, width = 40, units = 'cm')
+ggsave(fig2, filename = paste0(config$image_save_directory,'figure_3_estimate_compare.png'), height = 14, width = 40, dpi=1000, units = 'cm')

### 2018-10-22 analysis/plot_select_species_parameters.R

@@ -3,7 +3,7 @@ library(cowplot)
 config = yaml::yaml.load_file('config.yaml')
 
 ###################################################
-# This produced the supplementary figure comparing parameters
+# This produced the supplementary figures S10 and S11, comparing parameters
 # of 4 species. Adjust the ones below to create different plots. 
 #################################################
 
@@ -94,5 +94,5 @@ select_species_2 =
   (uniforc + no_legend) +
   plot_layout(ncol=1, heights=c(1,1))
 
-ggsave(paste0(config$image_save_directory,'supplement_select_species_param_comparison1.png'), plot=select_species_1, height=40, width=30, units = 'cm')
-ggsave(paste0(config$image_save_directory,'supplement_select_species_param_comparison2.png'), plot=select_species_2, height=40, width=30, units = 'cm')
+ggsave(paste0(config$image_save_directory,'figure_s10_select_species_param_comparison1.png'), plot=select_species_1, height=40,dpi=1000, width=30, units = 'cm')
+ggsave(paste0(config$image_save_directory,'figure_s11_select_species_param_comparison2.png'), plot=select_species_2, height=40,dpi=1000, width=30, units = 'cm')

### 2018-10-22 manuscript_map.R

@@ -55,6 +55,6 @@ site_map = ggplot() +
         plot.margin = unit(c(0,0,0,0),'mm'))
 #print(site_map)
 # Note, I saved this manually using the export tool to get ride of uneeded whitespace in the figure.
-ggsave('manuscript/figure_site_map.png', width=20, height=16, units = 'cm',dpi=1000, plot=site_map)
+ggsave('manuscript/figure_1_site_map.png', width=20, height=16, units = 'cm',dpi=1000, plot=site_map)
 
 

### 2018-10-22 manuscript/supplement_images.tex
first revision (#42)
* progress as of today on revision

includes notes from meeting with Ethan

* progress on revision

* add t-test to figure 4

* add a supp. table with overall aggregate errors for best models

* change the layout of supp images 1 and 2

Before they were all RMSE of train and test data for all
NPN models (S1) and LTER models (S2). Now they are RMSE
for all models of only held out data (S1) and pearson
correlation for all models of only held out data (S2)

* change T* to T_base in all images

* differentat the new cutoff time differences in config

* new sensitivity analysis

@@
 \begin{document}
-{\Comparison of large-scale citizen science data and long-term study data for phenology modeling \par}
 
+\Large
+Appendix S2
+
+AppendixS2.pdf
+
+Comparison of large-scale citizen science data and long-term study data for phenology modeling
+
+\normalsize
 \textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White}
@@
 
-Supplementary images S1 - S11 and Table S1
+Supplementary images S1 - S11 and Tables S1-S2
 
@@
 	\centering
-		\includegraphics[width=1\textwidth]{parameter_estimates_from_cutoff_sensitivity.png}
+		\includegraphics[width=1\textwidth]{figure_s1_parameter_estimates_from_cutoff_sensitivity.png}
 	Figure S1
@@
 	\centering
-		\includegraphics[width=1\textwidth]{figure_param_comparison_15day_cutoff.png}
+		\includegraphics[scale=0.7]{figure_s2_param_comparison.png}
 	Figure S2
@@
 	\centering
-		\includegraphics[width=1\textwidth]{figure_estimate_compare_15day_cutoff.png}
+		\includegraphics[width=1\textwidth]{figure_s3_estimate_compare.png}
 	Figure S3
@@
 	\centering
-		\includegraphics[width=1\textwidth]{figure_rmse_metrics_density_plot_15day_cutoff.png}
+		\includegraphics[width=1\textwidth]{figure_s4_rmse_metrics_density_plot.png}
 	Figure S4
@@
 	\centering
-		\includegraphics[width=1\textwidth]{supplement_all_model_rmse.png}
+		\includegraphics[width=1\textwidth]{figure_s5_all_model_rmse.png}
 	Figure S5
@@
 \textbf{Figure S6}: Pearson correlation coefficients for specific species and phenophases using all combinations of models and data sources.
-Note that since all predictions from each Naive model are the same pearson's correlation cannot be calculated here.
+Note that since all predictions from each Naive model are the same the Pearsons correlation cannot be calculated here.
 Red X's mark the best performing models for each respective dataset.
@@
 	\centering
-		\includegraphics[width=1\textwidth]{supplement_all_model_pearson.png}
+		\includegraphics[width=1\textwidth]{figure_s6_all_model_pearson.png}
 	Figure S6
@@
 	\centering
-		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
+		\includegraphics[width=1\textwidth]{figure_s7_scenario_absolute_rmse.png}
 	Figure S7
@@
 	\centering
-		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison1.png}
+		\includegraphics[scale=0.5]{figure_s8_hubbard_harvard_comparison1.png}
 	Figure S8
@@
 	\centering
-		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison2.png}
+		\includegraphics[scale=0.5]{figure_s9_hubbard_harvard_comparison2.png}
 	Figure S9
@@
 	\centering
-		\includegraphics[scale=0.5]{supplement_select_species_param_comparison1.png}
+		\includegraphics[scale=0.5]{figure_s10_select_species_param_comparison1.png}
 	Figure S10
@@
 	\centering
-		\includegraphics[scale=0.5]{supplement_select_species_param_comparison2.png}
+		\includegraphics[scale=0.5]{figure_s11_select_species_param_comparison2.png}
 	Figure S11

### 2018-10-22 manuscript/supplement_methods.tex
finalize supplement/appendix

@@
 \usepackage{lineno}
-    \linenumbers
+%    \linenumbers
 %\usepackage{setspace}
@@
 
-{\Comparison of large-scale citizen science data and long-term study data for phenology modeling \par} \newline
+Appendix S1
 
-\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White} \newline
+AppendixS1.pdf
 
-{\Large \textbf{Supplementary materials}} \newline
+\Large
+Comparison of large-scale citizen science data and long-term study data for phenology modeling
+
+\normalsize
+\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White}
+
+\textbf{\large Supplementary materials}
+
 Methods describing processing of the 4 LTER datasets

### 2018-10-23 analysis/cutoff_sensitivity_analysis.R
corrections for the sensitivity sample sizes

@@ -7,14 +7,13 @@ library(tidyverse)
 
 
 ##########################
-# The 1:1 charts of different parameters
+# Figure S1, the 1:1 charts of different parameters
 
 cutoff_15_day = read_csv('results/cutoff_15days/model_parameters.csv.gz')
 cutoff_15_day$cutoff = 'cutoff_15'
 
 cutoff_30_day = read_csv('results/cutoff_30days/model_parameters.csv.gz') 
 cutoff_30_day$cutoff = 'cutoff_30'
-
 
 both = cutoff_15_day %>%
   bind_rows(cutoff_30_day) %>%
@@ -25,6 +24,8 @@ both = cutoff_15_day %>%
   spread(cutoff, parameter_value) %>%
   fortify() %>%
   filter(complete.cases(.))
+
+total_sensitivity_comparisons = length(unique(both$species))
 
 sensitivity_figure = ggplot(both, aes(x=cutoff_15, y=cutoff_30)) + 
   geom_point() + 
@@ -65,13 +66,31 @@ sample_sizes_30d = read_csv('results/cutoff_30days/predictions.csv.gz') %>%
   ungroup()
 
 all_sample_sizes = sample_sizes_15d %>%
-  full_join(sample_sizes_30d, by='species')
+  full_join(sample_sizes_30d, by='species') %>%
+  mutate(phenophase = stringr::word(species,2,2, ' - '),
+         species = stringr::word(species,1,1,' - '))
 
 all_sample_sizes$n_decrease = all_sample_sizes$sample_size_30d - all_sample_sizes$sample_size_15d
 all_sample_sizes$n_percent_decrease = round(all_sample_sizes$n_decrease / all_sample_sizes$sample_size_15d, 2)
 
-total_30_day_spp = sum(!is.na(all_sample_sizes$sample_size_30d))
-total_15_day_spp = sum(!is.na(all_sample_sizes$sample_size_15d))
+# Total unique species in the analaysis
+total_30_day_spp = all_sample_sizes %>%
+  filter(!is.na(sample_size_30d)) %>%
+  select(species) %>%
+  distinct() %>%
+  nrow()
+
+total_15_day_spp = all_sample_sizes %>%
+  filter(!is.na(sample_size_15d)) %>%
+  select(species) %>%
+  distinct() %>%
+  nrow()
+
+# Total species/phenophase combinations for comparison. Since 1 species can have either budburst, flowering, or both.
[…]

### 2018-10-23 manuscript/supplement_images.tex
updated appendix2 latex file

@@
   \doublespacing
-\usepackage[document]{ragged2e}
 
@@
 \textbf{Figure S1}: Sensitivity test results from using a 15 versus 30 day cutoff between the 'yes' and most recent 'no' in the USA-NPN dataset.
-Each point represents the value of a single parameter for one of eight models for 33 species and phenophases in the USA-NPN dataset.
-Note only 33 comparisons were possible here as the stricter 15 day cutoff resulted in three of the species not having sufficient observations.
+Each point represents the value of a single parameter for one of eight models for 32 unique species/phenophase combinations.
+This is from 23 species with varying combinations of the budburst and flowering phenophases (see Table S1).
+Only 32 comparisons were possible here as the stricter 15 day cutoff resulted in three species/phenophase combinations not having sufficient observations.
+This figure does not include any data from the LTER datasets.
+Note that the primary analysis states 35 combinations using the 15 day cutoff.
+This is because 3 species are duplicated in the Harvard Forest and Hubbard Brook datasets, thus 3 extra comparisons are available for the primary analysis.
 

### 2018-10-25 manuscript/main.tex
final main text edits

@@
 Predictions for USA-NPN data had the lowest error when using models built from the USA-NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built.
-This difference is likely due to variation in phenological requirements within species.
+This difference in the cross-scale model comparison is likely due to variation in phenological requirements within species.
 Models using the USA-NPN dataset can integrate parameters over a large spatial scale while those using an LTER dataset can only estimate parameters for a single location.
@@
 Sites in the USA-NPN datasets are located across the U.S.
-and generally clustered around populated areas (Fig.
-1).
-To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig.
-1).
+and generally clustered around populated areas (Figure 1).
+To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Figure 1).
 All four long-term studies are located in the U.S.
@@
 In the LTER datasets observation metrics varied widely due to different protocols.
-To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods).
+To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see Appendix S1 for details).
 Three of the LTER datasets (Hubbard Brook, Harvard Forest, and H.J.
@@
 As with the USA-NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations.
-After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& S1).
+After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& Appendix S2: Table S1).
 Using a 15 day cutoff in the USA-NPN dataset resulted in 35 unique combinations with 23 species.
@@
 In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available.
-Absolute RMSE values as well as Pearson correlation coefficients are provided in the supplement for specific species (Fig.
-S5-S7) and with all observations aggregated together (Table S2).
+Absolute RMSE values as well as Pearson correlation coefficients are provided in the supplement for specific species (Appendix S2: Figure S5-S7) and with all observations aggregated together (Appendix S2: Table S2).
 
@@
 Throughout the analysis there were no qualitative differences between a 30-day or 15-day threshold between the first 'yes' and most recent 'no' observation in the USA-NPN dataset.
-Results presented here reflect the 30 day cutoff, see the supplementary figures S2-S4 for matching figures using a 15 day cutoff.
+Results presented here reflect the 30 day cutoff; see the supplementary figures S2-S4 in Appendix S2 for matching figures using a 15 day cutoff.
 
 The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$).
-The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Fig.
-2).
-The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the USA-NPN dataset (Fig.
-2).
+The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Figure 2).
+The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the USA-NPN dataset (Figure 2).
 The large outlier for the \textit{Fixed GDD model} is \textit{Larrea tridentata}; this species' flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}.
-While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Fig.
-S10-S11).
+While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Appendix S2: Figure S10-S11).
 
-When comparing estimates of phenological events between the two sets of models, many USA-NPN and LTER models produced similar estimates (Fig.
-3).
-The \textit{Fixed GDD model} had the highest correlation between the two models sets at USA-NPN sites ($R^2 = 0.82$), while the \textit{GDD}, \textit{M1}, and \textit{Uniforc} models had the highest correlation at LTER sites ($R^2 =$ 0.51, 0.52, and 0.51, respectively).
+When comparing estimates of phenological events between the two sets of models, many USA-NPN and LTER models produced similar estimates (Figure 3).
+The \textit{Fixed GDD model} had the highest correlation between the two model sets at USA-NPN sites ($R^2 = 0.82$), while the \textit{GDD}, \textit{M1}, and \textit{Uniforc} models had the highest correlation at LTER sites ($R^2 =$ 0.51, 0.52, and 0.51, respectively).
 Comparing models with spatial corrections to the non-spatial alternatives, the \textit{MSB} (an extension of the \textit{Alternating model} with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the \textit{Alternating model}.
@@
 
-When comparing the predictive performance using out-of-sample errors, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Fig.
-4).
+When comparing the predictive performance using out-of-sample errors, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Figure 4).
 The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both USA-NPN and LTER held-out observations.
 Although the \textit{Fixed GDD model} had the highest agreement in accuracy between USA-NPN and LTER-derived models, it was not the best performing model overall.
-The \textit{GDD} and \textit{Uniforc} models made the best out of sample predictions, having the lowest RMSE and Pearson coefficient when aggregating all observations together (Table S2).
+The \textit{GDD} and \textit{Uniforc} models made the best out of sample predictions, having the lowest RMSE and Pearson coefficient when aggregating all observations together (Appendix S2: Table S2).
 One exception was that the Fixed GDD model had a slightly higher Pearson value when using LTER-derived models to make predictions for USA-NPN data.
-The best model for specific species and phenophases varied, but was commonly the \textit{Uniforc} and \textit{GDD} models (Figs.
+The best model for specific species and phenophases varied, but was commonly the \textit{Uniforc} or \textit{GDD} models (Appendix S2: Figs.
 S5-S6).
@@
 
-Data used to build phenology models typically falls into two categories: intensive long-term data with long time-series at a small number of locations (e.g., LTER data in this study), and large-scale data with less intensive sampling at hundreds of locations (e.g., USA-NPN data) (Table 3).
+Data used to build phenology models typically fall into two categories: intensive long-term data with long time-series at a small number of locations (e.g., LTER data in this study), and large-scale data with less intensive sampling at hundreds of locations (e.g., USA-NPN data) (Table 3).
 This data scenario--a small amount of intensive data and a large amount of less intensive data--is common in many areas of science and makes it necessary to understand how to choose between, or combine, data sources \citep{hanks2018}.
@@
 
-In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the \textit{Fixed GDD} (Fig.
-2).
+In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the \textit{Fixed GDD} (Figure 2).
 These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters.
@@
 
-Despite substantial differences in parameter estimates, LTER and USA-NPN derived models produced similar estimates for phenological events in most cases (Fig.
-3).
+Despite substantial differences in parameter estimates, LTER and USA-NPN derived models produced similar estimates for phenological events in most cases (Figure 3).
 This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues.
@@
 This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters.
-In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Fig.
-S10-S11).
-Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig.
-S8-S9).
+In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Appendix S2: Figure S10-S11).
+Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Appendix S2: Figure S8-S9).
 These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates that were not biologically realistic \citep{chuine2016}.
@@
 While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the \textit{Fixed GDD model}, based on their ability to make out-of-sample predictions.
-Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both USA-NPN and LTER datasets (Fig.
-S5 \& S6).
+Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both USA-NPN and LTER datasets (Appendix S2: Figure S5 \& S6).
 This demonstrates that the more complex models are capturing additional information about phenology, and that some of the differences between datasets result from differences in either the scales or the sampling of the data.
 Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}.
-In combination with our results showing observed differences in parameter estimates between LTER sites (Fig.
-S8-S9), this suggests that variation in phenological requirements across the range is likely important.
+In combination with our results showing observed differences in parameter estimates between LTER sites (Appendix S2: Figure S8-S9), this suggests that variation in phenological requirements across the range is likely important.
 However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses.
-Specifically, correspondence between parameter estimates (Fig.
-2), estimates of phenological events (Fig.
-3), and out-of-sample error rates (Fig.
-4) for the \texit{MSB} and \textit{M1} models were essentially the same as the \textit{Alternating} and \textit{GDD} models, respectively.
+Specifically, correspondence between parameter estimates (Figure 2), estimates of phenological events (Figure 3), and out-of-sample error rates (Figure 4) for the \textit{MSB} and \textit{M1} models were essentially the same as the \textit{Alternating} and \textit{GDD} models, respectively.
 This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the USA-NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors.
@@
 In addition to exploring differences between phenology datasets, our analyses provide guidance on which models to use when making predictions at a local scale using models built from large-scale data, or vice versa.
-Among the eight models tested, the \textit{Uniforc} and \textit{GDD} models performed the best overall in the cross dataset comparison in terms of Pearson correlation and RMSE (Fig.
-S5-S6, Table S2).
+Among the eight models tested, the \textit{Uniforc} and \textit{GDD} models performed the best overall in the cross dataset comparison in terms of Pearson correlation and RMSE (Appendix S2: Figure S5-S6, Table S2).
 The \textit{GDD} model has one less parameter than the \textit{Uniforc} model, thus the \textit{GDD} model is a suitable choice for making predictions when there is little to no information at the location of interest (e.g.
 making phenology forecasts at a new location distant from any observed data).
-This guidance can vary between species though and model testing should still be performed when suitable data is available.
+This guidance can vary between species, though, and model testing should still be performed when suitable data are available.
 
@@
 \newpage
-
-\begin{adjustbox}{angle=90}
+\scriptsize
 %%renewcommand{\arraystretch}{2}\tabcolsep=5pt
@@
     \end{tabular}
-\end{adjustbox} \newline
+\newline
 \textbf{Table 1}: LTER datasets used in the analysis
@@
 \newpage
-\begin{adjustbox}{angle=90}
-\small
+
+\scriptsize
 %%{\def\arraystretch{2}\tabcolsep=5pt
@@
     Naive & \( \overline{DOY} \) & - & 1 & - \\
-    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = max(T_{i}, 0)$ & 1 & \citep{reaumur1735, wang1960, hunter1992} \\
     Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
     GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T_{base}, 0) $  & 3 & \citep{reaumur1735, wang1960, hunter1992} \\
+    Fixed GDD &$\sum_{t=1}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = max(T_{i}, 0)$ & 1 & \citep{reaumur1735, wang1960, hunter1992} \\
+    Alternating & $\sum_{t=1}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD(t)} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 3 & \citep{cannell1983} \\
+    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = \frac{1}{1 + e^{b(T_{i}-c)}} $ & 4 & \citep{chuine2000} \\
     M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T_{base}, 5) $  & 4 & \citep{blumel2012} \\
-    Alternating & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD(t)} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 3 & \citep{cannell1983} \\
-    MSB & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD_{i}} +dT_{mean} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 4 & \citep{jeong2013} \\
-    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = \frac{1}{1 + e^{b(T_{i}-c)}} $ & 4 & \citep{chuine2000} \\
+    MSB & $\sum_{t=1}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD_{i}} +dT_{mean} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 4 & \citep{jeong2013} \\
     \hline
     \end{tabular}
-\end{adjustbox} \newline 
-\textbf{Table 2}: Phenology models used in the analysis
+\newline
+
+\textbf{Table 2}: Phenology models used in the analysis.
+For all models, except the Naive and Linear models, the daily mean temperature $T_{i}$ is first transformed via the specified forcing equation.
+The cumulative sum of forcing is then calculated from a specific start date (either $DOY=1$ or using the fitted parameter $t_{1}$).
+The phenological event is estimated as the $DOY$ in which cumulative forcing is greater than or equal to the specified total required forcing (either $F^{*}$ or the specified equation).
+Parameters for each model are as follows: For the Naive model $\overline{DOY}$ is the mean day of year (ie.
+the Julian date) of a phenological event; for the Linear model $\beta_{1}$ and $\beta_{2}$ are the intercept and slope, respectively and $T_{mean}$ is the average daily temperature between January 1 and March 31; for the GDD model $F^{*}$ is the total accumulated forcing required, $t_{1}$ is the start date of forcing accumulation, and $T_{base}$ is the threshold daily mean temperature above which forcing accumulates; for the Fixed GDD model $F^{*}$ is the total accumulated forcing required; for the Alternating model $NCD$ is the number of chill days (daily mean temperature below 0$^{\circ}$C) from $DOY=0$ to the $DOY$ of the phenological event, $a$, $b$, and $c$ are the three fitted model coefficients; for the Uniforc model, is $F^{*}$ is the total accumulated forcing required, $t_{1}$ is the start date of forcing accumulation, and $b$ and $c$ are two additional fitted parameters which define the sigmoid function; the M1 model is the same as the GDD model, but with the additional fitted parameter $k$ that adjusts the total forcing accumulation according to day length; the MSB model is the same as the Alternating model, but with the additional fitted parameter $d$ to correct the model according to mean spring temperature.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-%% Figure 1, the map
+%% Figure 1 Legend, the map
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-\textbf{Figure 1}: Locations of National Phenology Network sites used (black points) and Long Term Ecological Research sites (labeled circles), with greyscale showing elevation.
-
-\newpage
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.6, angle=90]{figure_site_map.png}
-	\caption{.}
-\end{figure}
+\textbf{Figure 1}: Locations of U.S.A. National Phenology Network sites used (black points) and Long Term Ecological Research sites (labeled circles), with greyscale showing elevation.
 
 %%%%%%%%%%%%%%%%%%%%%%
-%% Figure 2, parameter comparison
+%% Figure 2 Legend , parameter comparison
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-\begin{figure}[H]
-	\centering
-		\includegraphics[scale=0.65]{figure_param_comparison.png}
-    \caption{.}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%
-%% Figure 3, estimate comparison (1:1 graphs w/ yellow and
+%% Figure 3 Legend, estimate comparison (1:1 graphs w/ yellow and
 %% green points
@@
 
-\newpage
-
-\begin{figure}[H]
-	\centering
-		\includegraphics[scale=0.5, angle=90]{figure_estimate_compare.png}
-	\caption{.}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%
-%% Figure 4, error comparison (2 rows of density plots)
+%% Figure 4 Legend, error comparison (2 rows of density plots)
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+%% Figure 1 the map
+%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{figure}[H]
 	\centering
-		\includegraphics[scale=0.4, angle=90]{figure_rmse_metrics_density_plot.png}
+	\includegraphics[scale=0.6]{figure_1_site_map.png}
+	\caption{.}
+\end{figure}
+
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%
+%% Figure 2 parameter comparison
+%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[H]
+	\centering
+		\includegraphics[scale=0.65]{figure_2_param_comparison_final.png}
+    \caption{.}
+\end{figure}
+
+\newpage
+
+%%%%%%%%%%%%%%%%%%%%%%
+%% Figure 3, estimate comparison (1:1 graphs w/ yellow and
+%% green points
+%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[H]
+	\centering
+		\includegraphics[scale=0.4]{figure_3_estimate_compare.png}
+	\caption{.}
+\end{figure}
+
+\newpage
+
+%%%%%%%%%%%%%%%%%%%%%%
+%% Figure 4  error comparison (2 rows of density plots)
+%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[H]
+	\centering
+		\includegraphics[scale=0.3]{figure_4_rmse_metrics_density_plot.png}
 	\caption{.}

### 2018-10-25 manuscript/main.tex
getting the abstract from 353 to 350 words

@@
 Large-scale observational data from citizen science efforts are becoming increasingly common in ecology, and researchers often choose between these and data from intensive local-scale studies for their analyses.
-This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability.
+This choice has trade-offs related to spatial scale, observer variance, and inter-annual variability.
 Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science USA National Phenology Network (USA-NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites.
@@
 We found that model parameter estimates for the same species were most similar between the two datasets when using simple models, but parameter estimates varied widely as model complexity increased.
-Despite this, estimates for the date of phenological events and out-of-sample errors were similar, regardless of the model chosen.
-Predictions for USA-NPN data had the lowest error when using models built from the USA-NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built.
+Despite this, estimates for the date of phenological events and out-of-sample errors were similar, regardless of the model.
+Predictions for USA-NPN data had the lowest error using models built from the USA-NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built.
 This difference in the cross-scale model comparison is likely due to variation in phenological requirements within species.
